### 2024-05-03 report/src/draft.tex
Initial commit of files from CCKM

\documentclass[runningheads]{llncs}
%
\usepackage{amsmath, amssymb, lineno, lscape,hyperref,subcaption, verbatimbox, bm}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\usepackage{color}
\newcommand{\todo}[1]{{\color{blue}TODO : {#1}}}

% Set control to 1 to include all proofs.
% Set control to 0 to shrink the paper to 12 pages
\def\control{1}

% Control of line numbering
%\linenumbers
% \renewcommand\linenumberfont{\normalfont\scriptsize}

% Graphics path
\graphicspath{{../fig/}}

\begin{document}
%
\title{The need for accuracy and smoothness in numerical simulations}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Carl Christian Kjelgaard Mikkelsen \inst{1}\orcidID{0000-0002-9158-1941 } \and
  Lori{\'e}n L{\'o}pez-Villellas \inst{2}\orcidID{0000-0002-1891-4359}
  % \and Pablo Garc{\'i}a-Risue{\~no} \inst{3} \orcidID{0000-0002-8142-9196}
}
%
\authorrunning{C. C. Kjelgaard Mikkelsen et al. }
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
  Department of Computing Science, Ume{\aa} University, 90187 Ume{\aa}, Sweden \email{spock@cs.umu.se} \\
  \and Barcelona Supercomputing Center, Barcelona, Spain, \\
  \email{lorien.lopez@bsc.es} \\
 %  \and Independent scholar \\
 %  \email{risueno@unizar.es, garcia.risueno@gmail.com}
}

%
\maketitle % typeset the header of the contribution
%
\begin{abstract} We consider the problem of estimating the error when solving a system of differential algebraic equations. Richardson extrapolation is a standard technique in numerical analysis that can be used to judge when rounding errors are irrelevant and estimate the discretization error. We have simulated molecular dynamics with constraints using GROMACS-library and found that the output is not always ameable to Richardson's analysis. We derive and illustrate Richardson's technique using a variety of numerical experiments. We identify two necessary conditions that are not always satisfied by GROMACS. 
\keywords{}
\end{abstract}
%
%
%

\section{Motivation} Consider the problem of solving the following system of differential algebraic equations
\begin{align}
  \bm{q}'(t) &= \bm{v}(t) \\
  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t) \\
  \bm{g}(\bm{q}(t)) &= \bm{0}. 
\end{align}
This is Newton's law for a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints specified by the function $\bm{g}$.
The vector $\bm{q}$ represents the position of the atoms.
The vector $\bm{v}$ represents the velocities of the atoms.
The nonsingular diagonal matrix $\bm{M}$ lists the masses of the atoms.
The function $\bm{G}$ is the Jacobian of the function $\bm{g}$.
In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm.
It uses a pair of staggered grids with uniform step size $h$ and takes the form
\begin{align}
  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda}_n \right), \\
  \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
  \bm{g}(\bm{q}_{n+1}) &= \bm{0}.
\end{align}
In reality, the constraint equation is a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
Let $T$ denote any value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
Superficially, our task is to estimate $T - A_h$, but the presence of truncation and rounding errors ensure that we cannot hope to compute the exact value of $A_h$. 
Let $\hat{A}_h(\tau,u)$ denote the value obtained from running the SHAKE algorithm using floating point arithmetic with unit roundoff $u$ and solving the constraint equations with a relative error that is bounded by $\tau$.
This is the value returned by a computer. 
We have the following trivial identity
\begin{equation}
  T - \hat{A}_h(\tau,u) = T - A_h + A_h - \hat{A}_h(\tau,u)
\end{equation}
We see that the error $T-\hat{A}_h(\tau,u)$ can be written as sum of two terms.
The discretization error $E_h = T - A_h$ and the computational error $A_h - \hat{A}_h(\tau,u)$.

We can reduce the computational error by running the constraint solver until stagnation and by reducing the unit roundoff.
It is frequently possible to use Richardson extrapolation to determine when computational errors are insignificant and estimate the discretization error.
In the past, the authors have sought to accelerate the solution of nonlinear constraint equations in the context of molecular dynamics by integrating Newton's method into the GROMACS library for constrained molecular dynamics.
We have now discovered that the output of GROMACS is not always amenable to Richardson's analysis.
In this paper we derive and illustrate the use of Richardson's technique using a variety of numerical experiments.
We identify two conditions that are not always satified by GROMACS and we demonstrate that each condition is necessary for the succesful application of Richardson's techniques.
Our data and software is freely available from our GitHub along with every script needed to generate every number, table and figure from scratch.

\section{Theory}

Consider the problem of approximation a target value $T$ using a numerical method $A = A_h$ that depends on a single real parameter $h$. We shall assume there exists nonzero real constants $\alpha$ and $\beta$ and exponents
\begin{equation}
  0 < p < q < r
\end{equation}
such that the error $E_h = T - A_h$ satisfies
\begin{equation} \label{equ:aex}
  E_h  = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
\end{equation}
We say that the error $E_h$ satisfies an asymptotic error expansion. Frequently, the exponents $(p,q,r)$ are all integers, but since we shall allow exponents that are not integers, we must insist that $h$ is strictly positive.

Our first task to estimate the error $E_h$ for a specific value of $h$. Richardson's error estimate $R_h$ is defined by the equation
\begin{equation}
 R_h =  \frac{A_h - A_{2h}}{2^p - 1}
\end{equation}
The following theorem shows that Richardson's error estimate is a good approximation of the error when $h$ is sufficiently small.

\begin{theorem} If $E_h$ satisfies equation \eqref{equ:aex}, then
  \begin{equation}
    \frac{E_h - R_h}{h^q} \rightarrow  \left(1 - \frac{2^q-1}{2^p-1} \right) \beta, \quad h \rightarrow 0_+.
  \end{equation}
\end{theorem}
\begin{proof}
  By assumption, there is a function $g(h)$ 
  \begin{equation}
    T - A_h  = \alpha h^p + \beta h^q + g(h)
  \end{equation}
  and constants $C>0$ and $h_0 > 0$ such that
  \begin{equation}
    \forall h \leq h_0 \: : \: |g(h)| \leq Ch^r.
  \end{equation}
  It follows that
  \begin{equation}
    T- A_{2h} = 2^p \alpha h^p + 2^q \beta h^q + g(2h).
  \end{equation}
  We conclude that
  \begin{equation} \label{equ:Dh:1}
    A_h - A_{2h} = (2^p - 1) \alpha h^p + (2^q - 1) h^q + g(2h) - g(h).
  \end{equation}
  It follows that
  \begin{equation}
    R_h = \frac{A_h - A_{2h}}{2^p - 1} = \alpha h^p + \frac{2^q-1}{2^p-1} \beta h^q + \frac{g(2h)-g(h)}{2^p-1}.
  \end{equation}
 This implies that
  \begin{equation}
    \alpha h^p =  R_h - \frac{2^q-1}{2^p-1} \beta h^q + O(h^r).
  \end{equation}
  We conclude that
  \begin{equation}
    E_h = R_h + \left(1 - \frac{2^q-1}{2^p-1} \right) \beta h^q + O(h^r)
  \end{equation}
  The theorem follows immediately from this expression because $q < r$.
\end{proof}
We shall now demonstrate how to demonstrate the existence of an asymptotic error expansion experimentally.
We define Richardson's fraction $F_h$ using the expression
\begin{equation}
  F_h = \frac{A_{2h} - A_{4h}}{A_h - A_{2h}}.
\end{equation}
The behavior of the function $h \rightarrow F_h$ is described by the following theorem.
\begin{theorem} Assume that $T-A_h$ satisfies equation \eqref{equ:aex} and let $(m,n)$ be given by
  \begin{equation}
    m = q - p, \quad n = r - p.
  \end{equation}
  Then Richardson's fraction satisfies
  \begin{equation}
    F_h \rightarrow 2^p, \quad h \rightarrow 0_+
  \end{equation}
  and 
  \begin{equation}
    \frac{F_h - 2^p}{h^m} \rightarrow (2^m-1) \nu, \quad \nu = \frac{2^q-1}{2^p-1} \frac{\beta}{\alpha}.
    \end{equation}
\end{theorem}

\begin{proof} It is convenient to rewrite equation \eqref{equ:Dh:1} as
  \begin{equation}
    A_h - A_{2h} = (2^p-1) \alpha h^p \Big [ 1  + \nu h^m + \phi(h) \Big]
  \end{equation}
  where $\phi(h) \in O(h^n)$. It follows immediately that
   \begin{equation}
      A_{2h} - A_{4h} = 2^p (2^p-1) \alpha h^p \Big[ 1 + 2^m \nu h^m + \phi(2h) \Big]
  \end{equation}
  This allows us to write
  \begin{equation}
    F_h = \frac{ A_{2h} - A_{4h}}{ A_h - A_{2h}} = 2^p \left[ \frac{ 1 + 2^m \nu h^m + \phi(h)}{ 1  + \nu h^m + \phi(2h)} \right]
  \end{equation}
  The fraction on the right-hand side is of the form
  \begin{equation}
    \frac{1 + f(h)}{1 + g(h)} = 1 + \frac{f(h) - g(h)}{1 + g(h)}
  \end{equation}
  where
  \begin{equation}
    f(h) = 2^m \nu h^m + \phi(h), \quad g(h) = \nu h^m + \phi(h)
  \end{equation}
  It follows immediately that
  \begin{equation}
    F_h = 2^p \left( 1 + \frac{(2^m-1)\nu h^m}{1+g(h)} + \frac{\phi(h) - \phi(2h)}{1+g(h)}\right) \rightarrow 2^p, \quad h \rightarrow 0_+
  \end{equation}
  and
  \begin{equation}
    \frac{F_h - 2^p}{h^m}  = \frac{(2^m-1)\nu }{1+g(h)} + \frac{\phi(h) - \phi(2h)}{(1+g(h))h^m} \rightarrow (2^m - 1) \nu, \quad h \rightarrow 0_+,
  \end{equation}
  because $m < n$, so that
  \begin{equation}
     \frac{\phi(h) - \phi(2h)}{h^m} \rightarrow 0, \quad h \rightarrow 0_+.
  \end{equation}
  This completes the proof.
\end{proof}
We conclude that if the error $E_h$ satisfies equation \eqref{equ:aex}, then the order of the primary error term can be determined as the limit
\begin{equation}
p = \underset{h \rightarrow 0_+}{\lim} F_h
\end{equation}
and the difference $m = q-p$ can be determined from the fact
\begin{equation} \label{equ:asymtotic-behavior-of-Fh}
  \log |F_h - 2^p| \approx \log(2^m-1) + \log|\nu| + m \log(h)
\end{equation}
is a good approximation for $h$ sufficiently small. In particular, we note that the left-hand side of equation \ref{equ:asymtotic-behavior-of-Fh} is essentially a linear function of $\log(h)$ with slope $m$.
\section{Elementary examples}

The theory applies to the difference $E_h = T - A_h$between the target value $T$ and the exact value of the approximation $A_h$.
In practice, rounding errors ensure that the computed value $\hat{A}_h$ is different from the exact error $A_h$. However, it is frequently possible to assert that $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the computed error $\hat{E}_h = T - \hat{A}_h$. We shall now demonstrate the procedure. To this end, we consider the problem of computing definite integrals
\begin{equation} \label{equ:integral}
  T = \int_{a}^b f(x) dx
\end{equation}
using the composite trazoidal rule $A_h$ given by
\begin{equation}
  A_h = \frac{1}{2}h \sum_{j=0}^{n-1} \left[ f(x_j) + f(x_{j+1}) \right], \quad x_j = jh, \quad nh = b-a, \quad n \in \mathbb{N}.
\end{equation}
It is well-known that if the function $f : [a,b] \rightarrow \mathbb{R} $ is \emph{everywhere} smooth, then there exists a sequence $\{\alpha_j\}_{j=1}^\infty \subset \mathbb{R}$ such that 
\begin{equation}
  E_h = \sum_{j=1}^k \alpha _j h^{2j} + O(h^{2k+1}), \quad h \rightarrow 0_+
\end{equation}

\paragraph{Integration of a function that is everywhere smooth}

Let $[a,b]= [0,1]$, let $f : [a, b] \rightarrow \mathbb{R}$ be given by $f(x) = e^x$ and $T$ be given by equation \eqref{equ:integral}.
The script {\tt rint\_mwe1} computes the composite trapezoidal sum $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,20\}$ and generates \ref{fig:rint_mwe1a} and \ref{fig:rint_mwe1b}.
The raw data shows that $\hat{A}_{h_k}$ approach $4 = 2^2$ as $k$ increases and $k \in \{2,3,\dots,14\}$.
This suggests that $p=2$.
Figure \ref{fig:rint_mwe1a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
We observe that $k \rightarrow \log_2|\hat{F}_{h_k} - 4|$ is essentially a linear function of $k$ with slope $-2$ for $k \in \{2,3,\dots,10\}$.
This is the so-called asymptotic range, where the computed value $\hat{A}_h$  behaves in a manner that is indistinguishable from the exact value $A_h$.
We conclude that the experiments supports the existence of an asymptotic error expansion with $(p,q) = (2,4)$.
Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T - \hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe1b}.
We observe that computed value of Richardson's error estimate is a good approxmation of the error $T - \hat{A}_h$.

% tiny scriptsize footnotesize small normalsize large Large LARGE

% \begin{myverbbox}[\scriptsize]{\valpha}
%  k |                A_h |         F_h |         R_h |         E_h | (E_h-R_h)/R_h
%  0 | 1.859140914230e+00 |  0.00000000 |  0.0000e+00 | -1.4086e-01 |    1.0000e+00
%  1 | 1.753931092465e+00 |  0.00000000 | -3.5070e-02 | -3.5649e-02 |    1.6251e-02
%  2 | 1.727221904558e+00 |  3.93908726 | -8.9031e-03 | -8.9401e-03 |    4.1402e-03
%  3 | 1.720518592164e+00 |  3.98447608 | -2.2344e-03 | -2.2368e-03 |    1.0400e-03
%  4 | 1.718841128580e+00 |  3.99610010 | -5.5915e-04 | -5.5930e-04 |    2.6031e-04
%  5 | 1.718421660316e+00 |  3.99902383 | -1.3982e-04 | -1.3983e-04 |    6.5098e-05
%  6 | 1.718316786850e+00 |  3.99975588 | -3.4958e-05 | -3.4958e-05 |    1.6276e-05
%  7 | 1.718290568083e+00 |  3.99993897 | -8.7396e-06 | -8.7396e-06 |    4.0690e-06
%  8 | 1.718284013367e+00 |  3.99998474 | -2.1849e-06 | -2.1849e-06 |    1.0167e-06
%  9 | 1.718282374686e+00 |  3.99999619 | -5.4623e-07 | -5.4623e-07 |    2.5759e-07
% 10 | 1.718281965016e+00 |  3.99999904 | -1.3656e-07 | -1.3656e-07 |    6.5583e-08
% 11 | 1.718281862598e+00 |  3.99999982 | -3.4139e-08 | -3.4139e-08 |    8.2385e-08
% 12 | 1.718281836994e+00 |  3.99999982 | -8.5348e-09 | -8.5348e-09 |    1.4743e-07
% 13 | 1.718281830593e+00 |  3.99999566 | -2.1337e-09 | -2.1337e-09 |   -3.7464e-06
% 14 | 1.718281828992e+00 |  3.99999681 | -5.3343e-10 | -5.3342e-10 |   -1.8177e-05
% 15 | 1.718281828592e+00 |  4.00004551 | -1.3335e-10 | -1.3335e-10 |   -2.7197e-05
% 16 | 1.718281828492e+00 |  4.00022202 | -3.3337e-11 | -3.3341e-11 |    1.1322e-04
% 17 | 1.718281828467e+00 |  4.00046186 | -8.3333e-12 | -8.3409e-12 |    9.1400e-04
% 18 | 1.718281828461e+00 |  3.98094194 | -2.0933e-12 | -2.0610e-12 |   -1.5658e-02
% 19 | 1.718281828460e+00 |  4.13782004 | -5.0589e-13 | -5.4334e-13 |    6.8928e-02
% 20 | 1.718281828459e+00 |  3.95086705 | -1.2805e-13 | -1.5921e-13 |    1.9572e-01
% \end{myverbbox}

% \begin{figure}
%   \centering
%   \valpha
%   \caption{Numerical integration of a function that is everywhere smooth.} \label{fig:rint_mwe1}
% \end{figure}

\begin{figure}[h]
\begin{subfigure}[h]{0.49\linewidth}
\includegraphics[width=\linewidth]{rint_mwe1a.png}
\caption{The evolution of $F_h$} \label{fig:rint_mwe1a}
\end{subfigure}
\hfill
\begin{subfigure}[h]{0.49\linewidth}
\includegraphics[width=\linewidth]{rint_mwe1b.png}
\caption{The size of $E_h$ and the accuracy of $R_h$} \label{fig:rint_mwe1b}
\end{subfigure}%
\caption{The evolution of $F_h$ and the accuracy of $R_h$ for a method with $(p,q) = (2,4)$.}
\end{figure}

\paragraph{Integration of a function that is smooth in all but one point}

Let $f : [0,1] \rightarrow \mathbb{R}$ be given by $f(x) = \sqrt{x}$ and let $T$ be given by equation \eqref{equ:integral}.
Then $T = \frac{2}{3}$. Since $f$ is not differentible at $x=0$ we have no guarantee that there exists an asymptotic error expansion.
The script {\tt rint\_mwe2} computes $A_h$ using $h_k = 2^{-k}$ for $k \in {0,1,\dots,25}$ and generates Figures \ref{fig:rint_mwe2a} and \ref{fig:rint_mwe2b}.
From the raw data it is clear $p = 2$ cannot be true, but it is plausible that $p = \frac{3}{2}$.
Figure \ref{fig:rint_mwe1a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
We observe that $k \rightarrow \log_2|\hat{F}_{h_k} - 2^{3/2}|$ is essentially a linear function of $k$ with slope $-\frac{1}{2}$ for $k \in \{2,3,\dots,18\}$.
This is the so-called asymptotic range where the computed numbers $\hat{A}_h$ behave in a manner that is indistinguishable from the exact value $A_h$.
We conclude that the experiment is consistent with an asymptotic error expansion with $(p,q)=(\tfrac{3}{2},2)$.
Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T-\hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe2b}.
We observe that computed value of Richardson's error estimate is a good approxmation of the error $T - \hat{A}_h$.
In fact, the corresponding relative error decreases when $k$ increases and we remain inside the asymptotic region.

% \begin{myverbbox}[\scriptsize]{\vbeta}
%  k |                A_h |         F_h |         R_h |         E_h | (E_h-R_h)/R_h
%  0 | 5.000000000000e-01 |  0.00000000 |  0.0000e+00 |  1.6667e-01 |    1.0000e+00
%  1 | 6.035533905933e-01 |  0.00000000 |  5.6635e-02 |  6.3113e-02 |    1.0264e-01
%  2 | 6.432830462427e-01 |  2.60645075 |  2.1729e-02 |  2.3384e-02 |    7.0765e-02
%  3 | 6.581302216245e-01 |  2.67590667 |  8.1202e-03 |  8.5364e-03 |    4.8762e-02
%  4 | 6.635811968772e-01 |  2.72376496 |  2.9812e-03 |  3.0855e-03 |    3.3782e-02
%  5 | 6.655589362789e-01 |  2.75616456 |  1.0817e-03 |  1.1077e-03 |    2.3534e-02
%  6 | 6.662708113785e-01 |  2.77821124 |  3.8934e-04 |  3.9586e-04 |    1.6465e-02
%  7 | 6.665256572968e-01 |  2.79335492 |  1.3938e-04 |  1.4101e-04 |    1.1556e-02
%  8 | 6.666165489765e-01 |  2.80384210 |  4.9710e-05 |  5.0118e-05 |    8.1285e-03
%  9 | 6.666488815500e-01 |  2.81114895 |  1.7683e-05 |  1.7785e-05 |    5.7264e-03
% 10 | 6.666603622190e-01 |  2.81626213 |  6.2790e-06 |  6.3044e-06 |    4.0386e-03
% 11 | 6.666644335930e-01 |  2.81985125 |  2.2267e-06 |  2.2331e-06 |    2.8505e-03
% 12 | 6.666658761272e-01 |  2.82237604 |  7.8895e-07 |  7.9054e-07 |    2.0130e-03
% 13 | 6.666663869116e-01 |  2.82415482 |  2.7936e-07 |  2.7976e-07 |    1.4221e-03
% 14 | 6.666665676940e-01 |  2.82540939 |  9.8873e-08 |  9.8973e-08 |    1.0049e-03
% 15 | 6.666666316585e-01 |  2.82629477 |  3.4983e-08 |  3.5008e-08 |    7.1017e-04
% 16 | 6.666666542854e-01 |  2.82692054 |  1.2375e-08 |  1.2381e-08 |    5.0217e-04
% 17 | 6.666666622882e-01 |  2.82736352 |  4.3769e-09 |  4.3785e-09 |    3.5681e-04
% 18 | 6.666666651184e-01 |  2.82765569 |  1.5479e-09 |  1.5483e-09 |    2.3781e-04
% 19 | 6.666666661192e-01 |  2.82794845 |  5.4735e-10 |  5.4746e-10 |    1.9396e-04
% 20 | 6.666666664731e-01 |  2.82781700 |  1.9356e-10 |  1.9355e-10 |   -6.1533e-05
% 21 | 6.666666665982e-01 |  2.82856545 |  6.8431e-11 |  6.8428e-11 |   -3.5721e-05
% 22 | 6.666666666425e-01 |  2.82593073 |  2.4215e-11 |  2.4152e-11 |   -2.6041e-03
% 23 | 6.666666666581e-01 |  2.84485993 |  8.5119e-12 |  8.5889e-12 |    8.9627e-03
% 24 | 6.666666666636e-01 |  2.81554159 |  3.0232e-12 |  3.0612e-12 |    1.2421e-02
% 25 | 6.666666666657e-01 |  2.66194397 |  1.1357e-12 |  9.8466e-13 |   -1.5341e-01
% \end{myverbbox}

% \begin{figure}
%   \centering
%   \vbeta
%   \caption{Numerical integration of a function that is smooth in all but one point.}
%   \label{fig:rint_mwe2} 
% \end{figure}

\begin{figure}
\begin{subfigure}[h]{0.49\linewidth}
\includegraphics[width=\linewidth]{rint_mwe2a.png}
\caption{The evolution of $F_h$} \label{fig:rint_mwe2a}
\end{subfigure}
\hfill
\begin{subfigure}[h]{0.49\linewidth}
\includegraphics[width=\linewidth]{rint_mwe2b.png}
\caption{The size of $E_h$ and the accuracy of $R_h$.} \label{fig:rint_mwe2b}
\end{subfigure}%
\caption{The evolution $F_h$ and the accuracy of $R_h$ for a method with $(p,q) = (\frac{3}{2},2)$.}
\end{figure}

We mention in passing that low order methods are more pratical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small and this issue is more acute for high order methods than for low order methods.

\section{Practical examples}

In this section we present the results of more elaborate examples that highlight the practical limitations of Richardson's extrapolation as well as some limitations of contemporary software for doing molecular dynamics. Every function or script necessary to replicate the numbers needed to reproduce the figures is freely available from the authors GitHub.

\subsection{A successful application of the theory}

In this section we shall demonstrate how the theory can be applied to analyze a nontrivial problem.

\paragraph{Example: Identify the shells fired by a howitzer}

Consider the D-20 howitzer whose effective range is known to be about 17.4 km. We have access to tables that describe the drag coefficient of 6 different shell types. Given the mass, diameter and muzzle velocity of the shells, can we determine the best possible match with any degree of confidence?

The script {\tt maxrange\_rk1} models a shells point particle moving in a plane subject to Earth's standard gravity and the international standard atmosperic model. Each trajectory is integrated using Euler's explicit method ({\tt 'rk1'}) and all but the final step has the same size $h$. The final step is adjusted to place the shell directly on the ground. While the relevant equation is linear and could be solved to machine precision, the bisection method is deliberately used to compute the final step with an error that is bounded by ${\tt tol} \cdot h$, where ${\tt tol}$ is set by the script. The drag function are interpolated from tables using cubic spline interpolation.

For each drag coefficient our target value $T$ is the maximum range of the shell as the elevation varies continuously from $0$ to $\frac{\pi}{2}$. For each drag coefficient, we compute $12$ different approximations $A_{h_k}$ of $T$ using the step size $h_k = 2^{3-k}$ seconds, where $k \in \{1,2,\dots,12\}$.
For each drag coefficient and for each value of the time step $h$, a range function is defined which returns the range of the shell as a function of the howitzers elevation $\theta$.
The range functions are unimodal and the maximum range is found using the golden section search algorithm.
The initial search bracket is $[0,\pi/2$] and this bracket is systematically reduced in length until it is shorter than $\frac{pi}{2}u$.

The script will either read the raw data from a file or generate it from scratch. In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}. These figures represent calculations where the shell is placed on the ground with ``excessive'' accuracy in the sense that the final step is computed with an error that is bounded $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff. Figure \ref{fig:maxrange_rk1_table_tol53} list the maximum range and the corresponding error estimate for each of the shells know to us computed using a time step of $h = 2^{-9} s$. In each case the error estimate suggests that the computed range is exact to number of figures shown. In particular, we see while a G7 type shell achieves a maximum range of 17.4 km, all other shells have ranges that are less than 16.9 km. However, it is a fallacy to conclude anything on the basis of the computed ranges and the associated error estimates. We need to assert that we are inside the asymptotic range and that the error estimates are reliable. To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}. For each of the 6 different drag coefficients we find that evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$. This result is entirely consistent with the use of Euler's explicit method. We observe that for each drag coefficient, $k=12$ is still inside the asymtotic range and we have no reason to doubt the quality of the error estimate. We conclude that the best shell model for the D-20 howitzer is the G7 shell.

\begin{myverbbox}[\normalsize]{\vgamma}
Shell type  Maximum range (m) Error estimate (m)
    G1           12832                0.4 
    G2           16857                0.1 
    G5           15918                0.2 
    G6           15556                0.2 
    G7           17461                0.1 
    G8           15914                0.1 
\end{myverbbox}

\begin{figure}
  \centering
  \vgamma
  \caption{The computed maximum range for 6 different shell types fired from the D-20 howitzer.}
  \label{fig:maxrange_rk1_table_tol53} 
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=12cm]{maxrange_rk1_tol53.png}
  \caption{The evolution of Richardson's fraction corresponding to the maximum range of 6 different shells fired from the D-20 howitzer.} \label{fig:maxrange_rk1_fraction_tol53}
\end{figure}

\subsection{An unsuccesful application of the theory}

In this section we present the experiment that triggered the production of the present paper. We attempted to apply Richardson's techniques in the context of molecular dynamics with constraints. It was a spectacular failure with no obvious pattern controlling the computed numbers. While we cannot claim to fully understand our failure, we have identified two key differences between the our two test cases and we have through a sequence of experiments verified that each difference is enough to destroy any hope of having an asymptotic error expansion.

We utilized GROMACS v2021 to conduct experiments on the behavior of hen egg white lysozyme submerged in water within a cubic simulation box, following Justin Lemkul's Lysozyme in Water GROMACS Tutorial. Several steps were taken to prepare the system for production simulation: first, ions were introduced to achieve electrical neutrality. Subsequently, energy minimization was performed using the steepest descent algorithm until the maximum force reached below 1000.0 kJ/(mol·nm). Following this, the system underwent 100 ps of equilibration in an NVT ensemble to stabilize temperature, followed by another 100 ps of equilibration in an NPT ensemble to stabilize pressure. The described process was replicated using two different force fields, OPLS-AA/L and CHARMM36. We conducted production simulations of 1 ps for both force fields, varying the number of executed steps from 250 to 16000 and employing two different tolerances for the SHAKE algorithm (1E-4 and 1E-12). For each experiment we computed the total kinetic and potential energy of the system at the end of the simulation.

\begin{figure}
  \centering
  \includegraphics[width=12cm]{charmm36tol04.pdf}
\end{figure}

\section{The need for sufficient accuracy}

There are reasons beyond error estimation for solving the constraint equations accurately. The chemistry literature contains several papers that bemoan the fact that the energy drift is unacceptably large unless the constraint equations are solved with sufficient accuracy. However, what is perhaps less well known is \emph{direction} of the energy drift can depend on the choice of algorithm used to solve the constraint equations! We shall now present two examples that illustrate this point. We shall simulate the motion of a simple pendulum over a large number of periods. The only difference between the two simulation shall be the choice of algorithm used solve the scalar constraint equation.

\paragraph{Example: A pendulum that gains energy through numerical error}

\paragraph{Example: A pendulum that loses energy through numerical error}

\section{The need for sufficient smoothness}

\section{Conclusion}

\subsubsection{Acknowledgments}

The first author is supported by eSSENCE, a collaborative e-Science programme funded by the Swedish Research Council within the framework of the strategic research areas designated by the Swedish Government.

% @Lorien: update these lines with current information.
% This work has been partially supported by the Spanish Ministry of Science and Innovation (contract PID2019-107255GB-C21/AEI/10.13039/501100011033), by the Generalitat de Catalunya (contract 2017-SGR-1328), and by Lenovo-BSC Contract-Framework Contract (2020).

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{references}

\end{document}

### 2024-05-03 report/experiments/gromacs/src/gromacs_all.m

@@ -1,0 +1,419 @@
+% Defines results obtained from GROMACS simulations as MATLAB structures
+% Experiments performed by Lorien Villelas-Lopez
+% Communicated to Carl Christian K. Mikkelsen via email.
+% Time-stamp: 03-MAY-2024 12:56
+
+% IMPORTANT NOTE: 
+% Columns marked F_h and E_h contain useless data
+% They are calculated under the assumption that the step size
+% is cut in half from one row to the next. 
+% This is not the case here, because the admissible range of stepsize
+% is too small.
+
+%Force field: charmm36 | Tolerance: 1e-12 | Parameter: positions
+% Idx   time-step       A_h       F_h       E_h
+C1.field='charmm36'; C1.tol=1e-12; C1.param='positions';
+C1.data=[
+0    0.004000  4.198331  0.000000  0.000000
+1    0.002000  4.201656  0.000000  0.001108
+2    0.001000  4.165227 -0.091273 -0.012143
+3    0.000909  4.185225 -1.821665  0.006666
+4    0.000833  4.166800 -1.085352 -0.006142
+5    0.000769  4.144273  0.817934 -0.007509
+6    0.000714  4.178890 -0.650738  0.011539
+7    0.000667  4.203821  1.388467  0.008310
+8    0.000625  4.248702  0.555508  0.014960
+9    0.000588  4.239574 -4.917246 -0.003042
+10   0.000556  4.197813  0.218554 -0.013920
+11   0.000526  4.202374 -9.157256  0.001520
+12   0.000500  4.246244  0.103952  0.014624
+13   0.000333  4.284727  1.140032  0.012827
+14   0.000250  4.362972  0.491813  0.026082
+15   0.000200  4.344009 -4.126154 -0.006321
+16   0.000167  4.515485 -0.110589  0.057159
+17   0.000143  4.602861  1.962492  0.029126
+18   0.000125  4.570891 -2.733038 -0.010657
+19   0.000111  4.751871 -0.176652  0.060327
+20   0.000100  4.916141  1.101718  0.054757
+21   0.000091  5.213454  0.552517  0.099104
+22   0.000083  5.371422  1.882101  0.052656
+23   0.000077  5.832080  0.342919  0.153553
+24   0.000071  5.977517  3.167411  0.048479
+25   0.000067  6.262725  0.509932  0.095069
+26   0.000063  6.720108  0.623567  0.152461
+];
+
+%Force field: charmm36 | Tolerance: 1e-12 | Parameter: velocities
[…]

### 2024-05-03 report/experiments/gromacs/src/gromacs_figures.m

@@ -1,0 +1,29 @@
+% Generate figures from GROMACS experiments
+
+% Set the path to figures
+figpath='../experiments/gromacs/fig/';
+
+% Load the data
+datpath='../experiments/gromacs/dat/';
+fname=strcat(datpath,'gromacs_all.mat');
+load(fname);
+
+% Collect the structures containing energies into a cell array
+c=cell(1,4); c{1}=C3; c{2}=C6; c{3}=O3; c{4}=O6;
+
+% Loop over the 4 sets of data
+for i=1:4
+
+    % Extract fields from the structure
+    aux=c{i}.data; field=c{i}.field; tol=c{i}.tol;
+
+    % Rename the columns to reduce the chance of misprints
+    ekin=aux(:,3); epot=aux(:,4); etol=ekin+epot;
+
+    % Call specialized plotting routine
+    [fig, ax]=gromacs_plot(field, tol, n, ekin, epot, etol);
+
+    fname=strcat(field,'tol',num2str(round(-log10(tol)),'%02d'),'.eps');
+    saveas(fig,strcat(figpath,fname));
+
+end

### 2024-05-03 report/experiments/gromacs/src/gromacs_plot.m

@@ -1,0 +1,37 @@
+function [fig, ax]=gromacs_plot(field, tol, n, ekin, epot, etol)
+
+fig=figure(); fig.Position=[100 800 1800 400];
+
+subplot(1,3,1); 
+plt1=plot(n,ekin); grid; 
+xlabel('number of steps'); 
+ylabel('Kinetic energy (kJ/mol)'); 
+
+% Set the fontsize for the axes
+ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
+
+subplot(1,3,2); 
+plt2=plot(n,epot); grid;
+
+xlabel('number of steps');
+ylabel('Potential energy (kJ/mol)');
+
+% Set the fontsize for the axes
+ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
+
+subplot(1,3,3); 
+plt3=plot(n,etol); grid;
+xlabel('number of steps');
+ylabel('Total energy (kJ/mol)');
+
+% Set the fontsize for the axes
+ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
+
+% Set the linewidth for the plot
+plt1.LineWidth=2;
+plt2.LineWidth=2;
+plt3.LineWidth=2;
+
+% Set a common title
+tit=sgtitle(['Forcefield = ' field '     constraint tolerance = ' num2str(tol,'%.2e')]);
+tit.FontSize=20;

### 2024-05-03 report/experiments/integration/src/rint_mwe1.m

@@ -1,0 +1,78 @@
+% Apply Richardson's techniques to integrating a smooth function
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-04-09  Adapted from 5dv231ht23
+
+% Interval
+a=0; b=1; 
+
+% Function
+f=@(x)exp(x); 
+
+% Integration rule
+rule=@(y,a,b,N)trapezoid(y,a,b,N); 
+
+% Theoretical order of the method 
+p=2; 
+
+% Number of refinements
+kmax=21; 
+
+% True value of the integral 
+val=exp(1)-1;
+
+% Apply Richardson's techniques
+data=rint(f,a,b,rule,p,kmax,val);
+
+% Obtain table parameters
+tp=table_param('rdif',data);
+
+% Print the information nicely on the screen
+print_table(data,tp);
+
+% Define path to figures
+mypath='../experiments/integration/fig/';
+
+% ////////////////////////////////////////
+% Generate Figure rint_mwe1a
+% ////////////////////////////////////////
+
+% Plot the evolution of Richardson's fraction
+figa=figure(); plta=plot_fraction(data,p);
+
+% Set the fontsize for the axes
+ax=figa.CurrentAxes; ax.FontSize=20; ax.GridLineWidth=2;
+
+% Set the linewidth for the plot
+plta.LineWidth=2;
+
+% Save the figure
+fname='rint_mwe1a.png'; saveas(figa,strcat(mypath,fname));
+
+% ////////////////////////////////////////
+% Generate Figure rint_mwe1b
+% ////////////////////////////////////////
+
+% Plot the error and the error estimate and the comparison
+figb=figure();
+
+% Isolate the relevant columns
+k=data(:,1); err=data(:,5); rel=data(:,6);
+
+% Plot the data
+hold on;
+pltb1=plot(k,log10(abs(err)));
+pltb2=plot(k,log10(abs(rel)));
+axis([0 kmax -14 2]);
+grid; lgdb=legend('log_{10}(|E_h|)','log_{10}(|E_h-R_h|/|E_h|)', ...
+    'Location','North');
+
+% Set the fontsize for the axes
+axb=figb.CurrentAxes; axb.FontSize=20; axb.GridLineWidth=2;
+
+% Set the linewidth for the plot
+pltb1.LineWidth=2; pltb2.LineWidth=2;
+
+% Save the figure
+fname='rint_mwe1b.png';
+saveas(figb,strcat(mypath,fname));

### 2024-05-03 report/experiments/iontrap/aux/cut1.m

@@ -1,0 +1,6 @@
+function y=cut1(rho,x)
+
+% Initialize the output
+y=zeros(size(x));
+% Anything less than rho is mapped to 1
+y(x<rho)=1;

### 2024-05-03 report/experiments/iontrap/aux/cut2.m

@@ -1,0 +1,4 @@
+function y=cut2(a,b,x)
+
+% Smooth transition from 1 to 0 on [a,b]
+y=1-f3(a,b,x);

### 2024-05-03 report/experiments/iontrap/aux/f1.m

@@ -1,0 +1,8 @@
+function y=f1(x)
+
+% Initialize output
+y=zeros(size(x));
+% Identify positive x
+idx=(x>0);
+% Modify the corresponding y values
+y(idx)=exp(-1./x(idx));

### 2024-05-03 report/experiments/iontrap/aux/f2.m

@@ -1,0 +1,4 @@
+function y=f2(x)
+
+% Smooth transition function between 1 and 0
+y=f1(x)./(f1(x)+f1(1-x));

### 2024-05-03 report/experiments/iontrap/aux/f3.m

@@ -1,0 +1,3 @@
+function y=f3(a,b,x)
+
+y=f2((x-a)./(b-a));

### 2024-05-03 report/experiments/iontrap/src/iontrap_mwe1.m

@@ -1,0 +1,148 @@
+% Ion trap with dampning
+% Runge-Kutta methods
+% No cutoff function
+
+% Set the central file name
+datpath='../experiments/iontrap/dat/';
+fname=strcat(datpath,"iontrap_mwe1.mat");
+
+% Check if raw data exists
+if ~isfile(fname)
+
+    % Run experiment from scracth
+    fprintf("Generating raw data\n");
+
+    % Seed the generator
+    seed=2024; rng(seed);
+
+    % Set the number of particles
+    n=4;
+
+    % Set the spatial dimension
+    dim=3;
+
+    % Set the leading dimension of the arrays
+    ld=n*dim;
+
+    % Generate position and momenta
+    q0=rand(ld,1)-rand(ld,1); p0=zeros(ld,1);
+
+    % Set the charges
+    charge=ones(n,1)/16;
+
+    % Generate initial state
+    x0=[q0; p0];
+
+    % Spring constant
+    spring=5;
+
+    % Dampning
+    damp=2;
+
+    % Scaling factor for columb potential
+    scale=1;
+
+    % Set the forcefield: coulomb force + springs + dampning
+    f=@(q,p)coulomb_force(dim,n,charge,q,scale)-spring*q-damp*p;
+
+    % Set the potential
+    U=@(q)coulomb_potential(dim,n,charge,q,scale)-0.5*spring*norm(q)^2;
+
+    % Set the function that drives the ODE
+    g=@(t,x)[x(ld+1:2*ld,1); f(x(1:ld,1),x(ld+1:2*ld,1))];
+
+    % The length of the simulation
+    T=10;
+
+    % Number of methods
+    nm=4;
+
+    % Define list of integration methods
+    tit=cell(4,1);
+    tit{1}='rk1';
+    tit{2}='rk2';
+    tit{3}='rk3';
+    tit{4}='rk4';
+
+    % Set the order of the primary error term
+    order=[1; 2; 3; 4];
+
+    % Number of samples
+    N1=512;
+
+    % Number of simulations/approximations
+    kmax=12;
+
+    % Allocate space for the raw data
+    raw=zeros(2*ld,N1+1,kmax,nm);
+
+    % Allocate space for the processed data
+    table=zeros(kmax,4,nm);
+
+    % Loop over the different methods
+    for j=1:nm
+        % Select the method
+        method=tit{j};
+        % Set steps between samples
+        N2=1;
+        % Loop over the different timestep sizes
+        for k=1:kmax
+            % Progress
[…]

### 2024-05-03 report/experiments/iontrap/src/iontrap_mwe2.m

@@ -1,0 +1,150 @@
+% Ion trap with dampning
+% Runge-Kutta methods
+% The forcefield has a jump discontinuity.
+% It is truncated at a finite distance.
+
+% Set the central file name
+datpath='../experiments/iontrap/dat/';
+fname=strcat(datpath,"iontrap_mwe2.mat");
+
+% Check if raw data exists
+if ~isfile(fname)
+
+    % Run experiment from scracth
+    fprintf("Generating raw data\n");
+
+    % Seed the generator
+    seed=2024; rng(seed);
+
+    % Set the number of particles
+    n=4;
+
+    % Set the spatial dimension
+    dim=3;
+
+    % Set the leading dimension of the arrays
+    ld=n*dim;
+
+    % Generate position and momenta
+    q0=rand(ld,1)-rand(ld,1); p0=zeros(ld,1);
+
+    % Set the charges
+    charge=ones(n,1)/16;
+
+    % Generate initial state
+    x0=[q0; p0];
+
+    % Spring constant
+    spring=5;
+
+    % Dampning
+    damp=2;
+
+    % Scaling factor for columb potential
+    scale=1;
+
+    % Set cutoff function
+    rho=0.1462/2; cut=@(x)cut1(rho,x);
+    
+    % Set the forcefield: coulomb force + springs + dampning
+    f=@(q,p)modified_coulomb_force(dim,n,charge,q,scale,cut)-spring*q-damp*p;
+
+    % Set the potential
+    U=@(q)coulomb_potential(dim,n,charge,q,scale)-0.5*spring*norm(q)^2;
+
+    % Set the function that drives the ODE
+    g=@(t,x)[x(ld+1:2*ld,1); f(x(1:ld,1),x(ld+1:2*ld,1))];
+
+    % The length of the simulation
+    T=10;
+
+    % Number of methods
+    nm=4;
+
+    % Define list of integration methods
+    tit=cell(4,1);
+    tit{1}='rk1';
+    tit{2}='rk2';
+    tit{3}='rk3';
+    tit{4}='rk4';
+
+    % Set the order of the primary error term
+    order=[1; 2; 3; 4];
+
+    % Number of samples
+    N1=512;
+
+    % Number of simulations/approximations
+    kmax=12;
+
+    % Allocate space for the raw data
+    raw=zeros(2*ld,N1+1,kmax,nm);
+
+    % Allocate space for the processed data
+    table=zeros(kmax,4,nm);
+
[…]

### 2024-05-03 report/experiments/iontrap/src/iontrap_mwe3.m

@@ -1,0 +1,148 @@
+% Ion trap with dampning
+% Runge-Kutta methods
+% The forcefield transitions smoothly to zero 
+
+% Set the central file name
+datpath='../experiments/iontrap/dat/';
+fname=strcat(datpath,"iontrap_mwe3.mat");
+
+% Check if raw data exists
+if ~isfile(fname)
+
+    % Run experiment from scracth
+    fprintf("Generating raw data\n");
+
+    % Seed the generator
+    seed=2024; rng(seed);
+
+    % Set the number of particles
+    n=4;
+
+    % Set the spatial dimension
+    dim=3;
+
+    % Set the leading dimension of the arrays
+    ld=n*dim;
+
+    % Generate position and momenta
+    q0=rand(ld,1)-rand(ld,1); p0=zeros(ld,1);
+
+    % Set the charges
+    charge=ones(n,1)/16;
+
+    % Generate initial state
+    x0=[q0; p0];
+
+    % Spring constant
+    spring=5;
+
+    % Dampning
+    damp=2;
+
+    % Scaling factor for columb potential
+    scale=1;
+
+    % Set cutoff function
+    rho=0.1462; rho1=0.5*rho; rho2=0.75*rho; cut=@(x)cut2(rho1, rho2, x);
+    
+    % Set the forcefield: coulomb force + springs + dampning
+    f=@(q,p)modified_coulomb_force(dim,n,charge,q,scale,cut)-spring*q-damp*p;
+
+    % Set the potential
+    U=@(q)coulomb_potential(dim,n,charge,q,scale)-0.5*spring*norm(q)^2;
+
+    % Set the function that drives the ODE
+    g=@(t,x)[x(ld+1:2*ld,1); f(x(1:ld,1),x(ld+1:2*ld,1))];
+
+    % The length of the simulation
+    T=10;
+
+    % Number of methods
+    nm=4;
+
+    % Define list of integration methods
+    tit=cell(4,1);
+    tit{1}='rk1';
+    tit{2}='rk2';
+    tit{3}='rk3';
+    tit{4}='rk4';
+
+    % Set the order of the primary error term
+    order=[1; 2; 3; 4];
+
+    % Number of samples
+    N1=512;
+
+    % Number of simulations/approximations
+    kmax=12;
+
+    % Allocate space for the raw data
+    raw=zeros(2*ld,N1+1,kmax,nm);
+
+    % Allocate space for the processed data
+    table=zeros(kmax,4,nm);
+
[…]

### 2024-05-03 report/experiments/iontrap/src/iontrap_mwe4.m

@@ -1,0 +1,149 @@
+% Ion trap with dampning
+% Runge-Kutta methods
+% The forcefield transitions smoothly to zero 
+
+% Set the central file name
+datpath='../experiments/iontrap/dat/';
+fname=strcat(datpath,"iontrap_mwe4.mat");
+
+% Check if raw data exists
+if ~isfile(fname)
+
+    % Run experiment from scracth
+    fprintf("Generating raw data\n");
+
+    % Seed the generator
+    seed=2024; rng(seed);
+
+    % Set the number of particles
+    n=4;
+
+    % Set the spatial dimension
+    dim=3;
+
+    % Set the leading dimension of the arrays
+    ld=n*dim;
+
+    % Generate position and momenta
+    q0=rand(ld,1)-rand(ld,1); p0=zeros(ld,1);
+
+    % Set the charges
+    charge=ones(n,1)/16;
+
+    % Generate initial state
+    x0=[q0; p0];
+
+    % Spring constant
+    spring=5;
+
+    % Dampning
+    damp=2;
+
+    % Scaling factor for columb potential
+    scale=1;
+
+    % Set cutoff function
+    rho=0.1462; rho1=0.5*rho; rho2=0.95*rho; cut=@(x)cut2(rho1, rho2, x);
+    
+    % Set the forcefield: coulomb force + springs + dampning
+    f=@(q,p)modified_coulomb_force(dim,n,charge,q,scale,cut)-spring*q-damp*p;
+
+    % Set the potential
+    U=@(q)coulomb_potential(dim,n,charge,q,scale)-0.5*spring*norm(q)^2;
+
+    % Set the function that drives the ODE
+    g=@(t,x)[x(ld+1:2*ld,1); f(x(1:ld,1),x(ld+1:2*ld,1))];
+
+    % The length of the simulation
+    T=10;
+
+    % Number of methods
+    nm=4;
+
+    % Define list of integration methods
+    tit=cell(4,1);
+    tit{1}='rk1';
+    tit{2}='rk2';
+    tit{3}='rk3';
+    tit{4}='rk4';
+
+    % Set the order of the primary error term
+    order=[1; 2; 3; 4];
+
+    % Number of samples
+    N1=512;
+
+    % Number of simulations/approximations
+    kmax=12;
+
+    % Allocate space for the raw data
+    raw=zeros(2*ld,N1+1,kmax,nm);
+
+    % Allocate space for the processed data
+    table=zeros(kmax,4,nm);
+
[…]

### 2024-05-03 report/experiments/maxrange/src/maxrange_rk1.m

@@ -1,0 +1,151 @@
+% Compute the range of shells of different types
+
+% Clear the work space
+clear;
+
+% Load the gun
+d20;
+
+% Load all shells
+load shells
+
+% Set the number of shell types
+numshell=6;
+
+% Set the method used to integrate the trajectory
+method='rk1';
+
+% Set the order of the method
+p=1;
+
+% Set path to data and figures
+datpath='../experiments/maxrange/dat/';
+figpath='../experiments/maxrange/fig/';
+
+% Set file name
+fname='maxrange_rk1.mat'; 
+fname=strcat(datpath,fname);
+
+% Create a list of drag functions
+f=cell(numshell,1);
+f{1}=mcg1;
+f{2}=mcg2;
+f{3}=mcg5;
+f{4}=mcg6;
+f{5}=mcg7;
+f{6}=mcg8;
+
+tit=cell(numshell,1);
+tit{1}='G1';
+tit{2}='G2';
+tit{3}='G5';
+tit{4}='G6';
+tit{5}='G7';
+tit{6}='G8';
+
+% Set the solver for the final timestep
+solver=@bisection;
+
+% Set the number of approximations
+kmax=12;
+
+% Array for tolerances
+aux=4:7:53; numtol=numel(aux); tol=2.^(-aux);
+
+% Check if the raw data exists
+if ~isfile(fname)
+    % Mark the start time
+    rt=cputime;
+    disp("Generating raw data");
+    % Allocate space for the output
+    raw_range=zeros(kmax,numshell,numtol);
+    raw_theta=zeros(kmax,numshell,numtol);
+    % Loop over the different tolerance
+    for t=1:numtol
+        % Set the tolerance for the final solve
+        cntl.eps=tol(t); % eps(1)/2;
+        cntl.delta=tol(t); % eps(1)/2;
+        cntl.maxit=60;
+
+        % Loop over the different shells
+        for j=1:numshell
+            % Select the drag coefficient
+            param.drag=f{j};
+            % Set the time step and maxstep
+            dt=4; maxstep=100;
+            % Loop over the time step size
+            for i=1:kmax
+                % Display progress indicator
+                disp([(t-1)*numshell*kmax+(j-1)*kmax+i numtol*numshell*kmax]);
+                % Define the range function
+                rf=@(theta)range_rkx(param, v0, theta, method, dt,...
+                    maxstep, solver, cntl);
+                % Find the largest range
[…]

### 2024-05-03 report/experiments/maxrange/src/maxrange_rk2.m

@@ -1,0 +1,151 @@
+% Compute the range of shells of different types
+
+% Clear the work space
+clear;
+
+% Load the gun
+d20;
+
+% Load all shells
+load shells
+
+% Set the number of shell types
+numshell=6;
+
+% Set the method used to integrate the trajectory
+method='rk2';
+
+% Set the order of the method
+p=2;
+
+% Set path to data and figures
+datpath='../experiments/maxrange/dat/';
+figpath='../experiments/maxrange/fig/';
+
+% Set file name
+fname='maxrange_rk2.mat'; 
+fname=strcat(datpath,fname);
+
+% Create a list of drag functions
+f=cell(numshell,1);
+f{1}=mcg1;
+f{2}=mcg2;
+f{3}=mcg5;
+f{4}=mcg6;
+f{5}=mcg7;
+f{6}=mcg8;
+
+tit=cell(numshell,1);
+tit{1}='G1';
+tit{2}='G2';
+tit{3}='G5';
+tit{4}='G6';
+tit{5}='G7';
+tit{6}='G8';
+
+% Set the solver for the final timestep
+solver=@bisection;
+
+% Set the number of approximations
+kmax=12;
+
+% Array for tolerances
+aux=4:7:53; numtol=numel(aux); tol=2.^(-aux);
+
+% Check if the raw data exists
+if ~isfile(fname)
+    % Mark the start time
+    rt=cputime;
+    disp("Generating raw data");
+    % Allocate space for the output
+    raw_range=zeros(kmax,numshell,numtol);
+    raw_theta=zeros(kmax,numshell,numtol);
+    % Loop over the different tolerance
+    for t=1:numtol
+        % Set the tolerance for the final solve
+        cntl.eps=tol(t); % eps(1)/2;
+        cntl.delta=tol(t); % eps(1)/2;
+        cntl.maxit=60;
+
+        % Loop over the different shells
+        for j=1:numshell
+            % Select the drag coefficient
+            param.drag=f{j};
+            % Set the time step and maxstep
+            dt=4; maxstep=100;
+            % Loop over the time step size
+            for i=1:kmax
+                % Display progress indicator
+                disp([(t-1)*numshell*kmax+(j-1)*kmax+i numtol*numshell*kmax]);
+                % Define the range function
+                rf=@(theta)range_rkx(param, v0, theta, method, dt,...
+                    maxstep, solver, cntl);
+                % Find the largest range
[…]

### 2024-05-03 report/experiments/pendulum/aux/initq_mwe1.m

@@ -1,0 +1,60 @@
+% Generates a valid configuration for a chain of particles
+
+% Initialize the random generator
+rng(2022);
+
+% Set the number of atoms
+n=15;
+
+% Set the number of bonds
+m=n-1;
+
+% Set the bond lengths
+l=(1:m); l=mod(l,2)+1;
+
+% Set the dimension of the vector space
+dim=3;
+
+% Include this information in the parameters
+param.dim=dim;
+
+% Define the parameters of the chain
+param.n=n;
+param.l=l;
+
+% Define the constraint function
+g=@(q)chain(param,q);
+
+% Define the Jacobian of the constraint function
+dg=@(q)chain_jacobian(param,q);
+
+% Add the constraint function and the Jacobian to the parameters
+param.g=g;
+param.dg=dg;
+
+% Initialize the position of the atoms
+q0=rand(dim*n,1)-rand(dim*n,1);
+
+% Set the tolerance and maxit
+tol=1e-14; maxit=100;
+
+% Initialization
+[q, flag, it, his, res]=initq(q0, param, tol, maxit);
+
+if (dim==2 || dim==3)
+    % Plot the solution
+    for i=1:it+1
+        aux=reshape(his(:,i),dim,n);
+        % Plot the location of the atoms
+        if (dim==3)
+            plot3(aux(1,:),aux(2,:),aux(3,:),'-*');
+        end
+        if (dim==2)
+            plot(aux(1,:),aux(2,:),'-*');
+        end
+        % Finalize the graphics
+        s=strcat("i = ",num2str(i)," res = ",num2str(res(i))); title(s); grid; xlabel('x'); ylabel('y'); zlabel('z');
+        % Pause 1 second before the next iteration
+        pause(1);
+    end
+end

### 2024-05-03 report/experiments/pendulum/aux/initv.m

@@ -1,0 +1,35 @@
+function v=initv(q0, v0, param)
+
+% Find a velocity that is compatible with the hidden constraint
+%
+%     Dg(q0)v_0 = 0
+%
+% INPUT:
+%   q0     the initial position of the atoms
+%   v0     initial guess for a compatible velocity
+%   param  a structure such that
+%             param.dg is the Jacobian
+%
+% OUTPUT:
+%   v      a velocity vector that is compatible with the hidden constraint.  
+%
+% MINIMAL WORKING EXAMPLE: TODO
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2022-XX-YY  Initial programming and testing
+%   2022-08-03  Completed the internal documentation
+%
+% REMARK: It is the responsiblilty of the user to ensure that the
+% dimensions match.
+
+% Isolate the Jacobian
+dg=param.dg;
+
+% Evaluate the Jacobian at q0;
+A=dg(q0);
+
+% Determine an orthogonal basis for the null space of A
+Q=null(full(A));
+
+% Project v0 onto N(A);
+v=Q*(Q'*v0);

### 2024-05-03 report/experiments/pendulum/aux/pendulum1.m

@@ -1,0 +1,73 @@
+% Defines a pendulum that moves in a plane
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-04-26 Adapted from existing code
+
+
+% Define the dimension of the vector space
+dim=2;
+
+% Define the mass of the particle
+mass=1;
+
+% Define the constant of gravity
+grav=9.82;
+
+% Define the potential
+phi=@(x)mass*grav*x(dim);
+
+% Define the constant forcefield
+f=@(x)[0; -mass*grav];
+
+% Define the length of the pendulum
+l=1;
+
+% Define the point where the pendulum is attached
+s=[0; l];
+
+% Define the constraint function
+g=@(r)0.5*(l^2-norm(s-r)^2);
+
+% Define the Jacobian of the constraint function
+dg=@(r)(s-r)';
+
+% Compute theoretical period
+period=2*pi*sqrt(l/grav);
+
+% //////////////////////////////////////////////////////////////
+%  The initial condition of the system
+% //////////////////////////////////////////////////////////////
+
+% Initial position of the pendulum.
+r0=[0; 0];
+
+% Initial velocity of the pendulum.
+v0=[1; 0];
+
+% ///////////////////////////////////////////////////////////
+%  Describe the physics
+% ///////////////////////////////////////////////////////////
+
+% Set the dimension of the vector space
+param.dim=dim;
+
+% Number of constraints
+param.m=1;
+
+% Number of atoms
+param.n=1;
+
+% Mass of particles
+param.mass=mass;
+
+% Potential
+param.phi=phi;
+
+% Forcefield
+param.f=f;
+
+% Constraint function
+param.g=g;
+
+% Jacobian of constraint function
+param.dg=dg;

### 2024-05-03 report/experiments/pendulum/src/shake_pendulum_mwe1.m

@@ -1,0 +1,68 @@
+% SHAKEs pendulum
+
+% Clear the workspace
+clear;
+
+% Define the pendulum 
+pendulum1; 
+
+% /////////////////////////////////////////////////////////
+%   Define the time interval and the number of steps
+% /////////////////////////////////////////////////////////
+
+% Set the number of periods
+np=500;
+
+% Set the number of samples to record
+N1=20*np;
+
+% Set the number of time steps between recorded samples
+N2=[10 20 40];
+
+% Define the time interval
+a=0; b=N1*period;
+
+% /////////////////////////////////////////////////////////
+%   Set parameters for the constraint solver
+% /////////////////////////////////////////////////////////
+
+% Specify the methods used
+method={'newton','quasi'};
+
+% Speficy the tolerance
+tol=[1e-12,1e-4];
+
+% Number of iterations pr. equations solved
+maxit=20;
+
+% Allocate space to store the data
+data=zeros(N1+1,2,3,2);
+
+% Loop over the method
+for l=1:2
+    % Loop over the columns
+    for j=1:3
+        % Loop over the tolerances
+        for i=1:2
+            [r, v, t, lambda, it, residual, flag, energy]=...
+                shake(r0, v0, param, a, b, N1, N2(j), method{l}, tol(i), maxit);
+            data(:,i,j,l)=energy';            
+        end
+    end
+end
+
+
+% Loop over the method
+for l=1:2
+    % Loop over the columns
+    for j=1:3
+        % Loop over the tolerances
+        for i=1:2
+            % Extract energy
+            energy=data(:,i,j,l);
+            % Plot the energy drift
+            subplot(4,3,(l-1)*6+(i-1)*3+j); 
+            plot((energy-energy(1))./energy(1));
+        end
+    end
+end

### 2024-05-03 report/matlab/ballistics/artillery/d20.m

@@ -1,0 +1,34 @@
+% D20 152 mm howitzer, USSR, early 1950s
+
+% Data extracted from:
+% https://en.wikipedia.org/wiki/152_mm_towed_gun-howitzer_M1955_%28D-20%29
+
+% Load shell models
+load shells mcg7;
+
+% Set the mass of the shell
+param.mass=44;
+
+% Caliber in meters
+param.cali=0.1524;
+
+% Select projectile type G7 (based on visual comparison by CCKM)
+param.drag=@(x)mcg7(x);
+
+% Select standard atmosphere
+param.atmo=@(x)atmosisa(x);
+
+% Select standard gravity
+param.grav=@(x)9.80665;
+
+% Select muzzle velocity
+v0=655; 
+
+% Select method and time step for integration
+method='rk1'; dt=1; 
+
+% Select maximum number of timesteps
+maxstep=1000;
+
+% Print header
+fprintf('D20 152mm howitzer firing G7 shells (boat-tail) loaded\n\n');

### 2024-05-03 report/matlab/ballistics/artillery/flak36.m

@@ -1,0 +1,33 @@
+% 88 mm anti-aircraft gun, Germany, WWII
+% Data extracted from:
+% https://en.wikipedia.org/wiki/8.8_cm_Flak_18/36/37/41
+
+% Ensure that the shell models are loaded
+load shells
+
+% Set the mass of the shell
+param.mass=9.2;
+
+% Caliber in meters
+param.cali=0.088;
+
+% Select projectile type G6 (based on visual comparison by CCKM)
+param.drag=@(x)mcg6(x);
+
+% Select standard atmosphere
+param.atmo=@(x)atmosisa(x);
+
+% Select standard gravity
+param.grav=@(x)9.80665;
+
+% Select muzzle velocity
+v0=840; 
+
+% Select method and time step for integration
+method='rk1'; dt=0.25; 
+
+% Select maximum number of timesteps
+maxstep=800;
+
+% Print header
+fprintf('Flak 36 anti-aircraft gun firing G6 shells loaded\n\n');

### 2024-05-03 report/matlab/ballistics/artillery/kmk14.m

@@ -1,0 +1,34 @@
+% 420 mm heavy siege gun ("Dicke Bertha"), Germany, WWI
+
+% Data extracted from:
+% https://en.wikipedia.org/wiki/Big_Bertha_%28howitzer%29
+
+% Ensure that the shell models are loaded
+load shells
+
+% Set the mass of the shell
+param.mass=820;
+
+% Caliber in meters
+param.cali=0.420;
+
+% Select projectile type G1 (based on visual comparison by CCKM)
+param.drag=@(x)mcg1(x);
+
+% Select standard atmosphere
+param.atmo=@(x)atmosisa(x);
+
+% Select standard gravity
+param.grav=@(x)9.80665;
+
+% Select muzzle velocity
+v0=400; 
+
+% Select method and time step for integration
+method='rk1'; dt=1; 
+
+% Select maximum number of timesteps
+maxstep=1000;
+
+% Display header
+fprintf('"Dicke Bertha" firing G1 shells loaded\n\n');

### 2024-05-03 report/matlab/ballistics/artillery/m114.m

@@ -1,0 +1,34 @@
+% M114 155 mm howitzer, USA, WWII
+
+% Data extracted from
+% https://en.wikipedia.org/wiki/M114_155_mm_howitzer
+
+% Ensure that the shell models are loaded
+load shells
+
+% Set the mass of the shell
+param.mass=41.86;
+
+% Caliber in meters
+param.cali=0.15471;
+
+% Select projectile type G7 (based on visual comparison by CCKM)
+param.drag=@(x)mcg7(x);
+
+% Select standard atmosphere
+param.atmo=@(x)atmosisa(x);
+
+% Select standard gravity
+param.grav=@(x)9.80665;
+
+% Select muzzle velocity
+v0=563; 
+
+% Select method and time step for integration
+method='rk1'; dt=1; 
+
+% Select maximum number of timesteps
+maxstep=1000;
+
+% Print header
+fprintf('M114 155mm howitzer firing G7 shells (boat-tail) loaded\n\n');

### 2024-05-03 report/matlab/ballistics/artillery/pzh2000.m

@@ -1,0 +1,34 @@
+% 155 mm self-propelled howitzer, Germany, current era
+
+% Data extracted from:
+% http://www.navweaps.com/Weapons/WNGER_61-52_MONARC.htm
+
+% Ensure that the shell models are loaded
+load shells
+
+% Set the mass of the shell
+param.mass=44.5;
+
+% Caliber in meters
+param.cali=0.155;
+
+% Select projectile type G7 (based on visual comparison by CCKM)
+param.drag=@(x)mcg7(x);
+
+% Select standard atmosphere
+param.atmo=@(x)atmosisa(x);
+
+% Select standard gravity
+param.grav=@(x)9.80665;
+
+% Select muzzle velocity
+v0=945; 
+
+% Select method and time step for integration
+method='rk1'; dt=1; 
+
+% Select maximum number of timesteps
+maxstep=1000;
+
+% Print header
+fprintf('PzH-2000 firing G7 shells (boat-tail) loaded\n\n');

### 2024-05-03 report/matlab/ballistics/compute_elevation.m

@@ -1,0 +1,116 @@
+function [table, count]=compute_elevation(r,param,v0,method,dt,maxstep,sf1,cntl,sf2,sp2)
+
+% COMPUTE_ELEVATION Find pairs of firing solutions
+%
+% CALL SEQUENCE:
+%
+%   [table, count]=compute_elevation(r,param,v0,method,dt,maxstep,sf1,cntl,sf2,sp2)
+%
+% INPUT:
+%   r         array of distances to targets
+%   param     a structure describing the shell and the environment
+%              see range_rkx for details
+%   v0        the muzzle velocity of the gun
+%   method    the method used to integrate the trajectories
+%               see rk for options
+%   dt        the standard time step
+%   maxstep   the maximum number of time steps per trajectory
+%   sf1       handler to a function used to solve for the final time step
+%   cntl      a structure of parameters for the solver
+%                cntl.delta    controls the error
+%                cntl.epsilon  controls the residual
+%                cntl.maxit    controls maximum number of iterations
+%   sf2       handler to a function used to solve for elevations
+%   sp2       structure controlling the behavior of sf2
+%               sp2.delta  absolute error tolerance
+%               sp2.eps    absolute residual tolerance
+%               sp2.maxit  maximum number of iterations allowed
+%
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-03-05  Initial programming and testing
+
+% Determine the number of targets
+m=numel(r);
+
+% Sort the targets in ascending order
+r=sort(reshape(r,m,1));
+
+% Number of test trajectories
+n=9; 
+
+% Prepare for a number of test firings
+psi=linspace(0,pi/2,n);
+
+% Compute an auxiliary table that is used to obtain crude brackets
+aux=compute_range(param,v0,psi,method,dt,maxstep,sf1,cntl);
+
+% Initialize the number of trajectories computer
+count=n;
+
+% Allocate space for the table
+table=zeros(m,5);
+
+% Number of decimal digits of m 
+num=floor(log10(m)+1);
+fprintf('Computing firing solutions for %*d targets \n',num,m);
+
+% Main loop over the targets
+for i=1:m
+    fprintf('.'); 
+
+    % Save the ith range in the table
+    table(i,1)=r(i);
+    
[…]

### 2024-05-03 report/matlab/ballistics/compute_range.m

@@ -1,0 +1,94 @@
+function [table, flag]=compute_range(pp, v0, theta, method, dt, maxstep, solver, cntl)
+
+% COMPUTE_RANGE Computes the range as function of the elevation
+%
+% CALL SEQUENCE: 
+%
+%   [table, flag]=compute_range(pp,v0,theta,method,dt,maxstep,solver,cntl)
+%
+% INPUT:
+%   pp        the physical parameters
+%                pp.mass   the mass of the shell
+%                pp.cali   the caliber of the shell
+%                pp.drag   a function computing the drag coeffient
+%                pp.atmo   a function computing the atmosphere
+%                pp.grav   a function computing gravity
+%                pp.wind   a function computing the wind
+%   v0        the muzzle velocity of the gun
+%   theta     an array of elevations of dimension 1
+%   method    a string describing the method used to compute the trajectory
+%   dt        the length of all but the final time step
+%   maxstep   the maximum number of time steps allowed
+%   solver    a handler to function for solving nonlinear equations
+%   cntl      a structure of parameters for the solver
+%                cntl.delta    controls the error
+%                cntl.epsilon  controls the residual
+%                cntl.maxit    controls maximum number of iterations
+% OUTPUT:
+%   table     a two dimensional array, such that
+%               table(j,1) = theta(j)
+%               table(j,2) = the range obtained with elevation theta(j)
+%               table(j,3) = the corresponding flight time 
+%   flag      if flag =  1, then everything went smoothly
+%             if flag = -j, then execution failed using elevation j
+%
+% Failures are often caused by small values of MAXSTEP
+%
+% MINIMAL WORKING EXAMPLE: compute_range_mwe1
+%
+% See also: RANGE_RK1, RANGE_RKX
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  Fall 2014 : Initial programming and testing
+%  2015-09-22: Integrated global variables m, k, g into structure
+%  2015-09-22: Minor improvement of documentation
+%  2015-10-31: Replaced structure const with mandatory PARAM
+%  2016-06-27: Integrated displaytable()
+%  2016-09-09: Adapted function to SHELL4A
+%  2024-03-01: Removed printing
+%  2024-03-04: Introduced control over the final time step
+
+% Reshape and sort the array containing the elevations
[…]

### 2024-05-03 report/matlab/ballistics/mwe/compute_elevation_mwe1.m

@@ -1,0 +1,67 @@
+% Fast computation of firing solutions for the D-20 howitzer
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2021-05-12  Initial programming and testing
+%   2024-03-04  Essentially rewritten from scratch
+
+% Define the piece of artillery
+d20;
+
+% Distance to targets
+r=(1000:1000:14000);
+
+% ////////////////////////////////////////////////////////////////////
+%  Select solver for computing the final time step for each trajectory
+% ////////////////////////////////////////////////////////////////////
+sf1=@robust_secant;
+
+% Control parameters for the first solver
+cntl.delta=eps(1)/2;
+cntl.eps=eps(1)/2;
+cntl.maxit=53;
+
+% Define the correspoinding range function
+rf1=@(theta)range_rkx(param,v0,theta,method,dt,maxstep,sf1,cntl); disp(rf1);
+
+% ////////////////////////////////////////////////////////////////////
+% Select solver for computing the firing solution for each target
+% ////////////////////////////////////////////////////////////////////
+sf2=@robust_secant;
+
+% Tolerances and maximum number of iterations
+sp2.delta=eps(1)/2;       % Why is this parameter set to zero?
+sp2.eps=eps(1)/2;         % Why should this be the kill radius of the shell?
+sp2.maxit=20;             % Is this always a reasonable value?
+
+% Start the clock;
+start=tic;
+
+% Compute table
+[table1, count1]=compute_elevation(r,param,v0,method,...,
+    dt,maxstep,sf1,cntl,sf2,sp2);
+
+% Stop the clock
+time=toc(start);
+
+% Number of targets
+n=numel(r);
+
+% ////////////////////////////////////////////////////////////////////
+% Print the table
+% ////////////////////////////////////////////////////////////////////
+
+% Obtain predefined parameters for how to format the table
+tp=table_param('compute_elevation');
+
+% Convert the elevations from radians to degrees
+aux=180/pi; data1=table1; data1(:,[2 4])=data1(:,[2 4])*aux; 
+
+% Print the table
+print_table(data1, tp);
+
+% Statistics
+fprintf('\n');
+fprintf('SUMMARY:\n');
+fprintf('Number of targets                 : %4d\n',n);
+fprintf('Number of trajectories computed   : %4d\n',count1);
+fprintf('Wall time (seconds)               : %7.2f\n',time);

### 2024-05-03 report/matlab/ballistics/mwe/compute_elevation_mwe2.m

@@ -1,0 +1,72 @@
+% Slow computation of firing solutions for the D-20 howitzer
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2021-05-12  Initial programming and testing
+%   2024-03-04  Essentially rewritten from scratch
+
+% This scipt is solves the same problem as compute_elevation_mwe1
+% However the author has choosen poorly among the available algorithms
+
+% Define the piece of artillery
+d20;
+
+% Select a very accurate method
+method='rk4';
+
+% Distance to targets
+r=(1000:1000:14000);
+
+% ////////////////////////////////////////////////////////////////////
+%  Select solver for computing the final time step for each trajectory
+% ////////////////////////////////////////////////////////////////////
+sf1=@bisection;
+
+% Tolerances and maximum number of iterations
+cntl.delta=1e-2;  % We cannot do error estimation unless delta is O(u)
+cntl.eps=1e-2;    % We cannot do error estimation unless eps is O(u)
+cntl.maxit=60;    % Is this a reasonable value?
+
+% Define the correspoinding range function
+rf2=@(theta)range_rkx(param,v0,theta,method,dt,maxstep,sf1,cntl);
+
+% //////////////////////////////////////////////////////////////////
+% Select solver for computing the firing solution for each target
+% ////////////////////////////////////////////////////////////////////
+sf2=@bisection;
+
+% Tolerances and maximum number of iterations
+sp2.delta=eps(1)/2;  % We cannot do error estimation unless delta is O(u)
+sp2.eps=eps(1)/2;    % We cannot do error estimation unless eps is O(u)
+sp2.maxit=60;        % Is a reasonable value?
+
+% Start the clock;
+start=tic;
+
+% Compute table
+[table2, count2]=compute_elevation(r,param,v0,method,dt,maxstep,sf1,cntl,sf2,sp2);
+
+% Stop the clock
+time=toc(start);
+
+% Number of targets
+n=numel(r);
+
+% ////////////////////////////////////////////////////////////////////
+% Print the table
+% ////////////////////////////////////////////////////////////////////
+
+% Obtain predefined parameters for how to format the table
+tp=table_param('compute_elevation');
+
+% Convert the elevations from radians to degrees
+aux=180/pi; data2=table2; data2(:,[2 4])=data2(:,[2 4])*aux; 
+
+% Print the table
[…]

### 2024-05-03 report/matlab/ballistics/mwe/compute_range_mwe1.m

@@ -1,0 +1,39 @@
+% COMPUTE_RANGE_MWE1 Computes a table of ranges for the Flak36 
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2017-10-24 Extracted from compute_range
+%   2024-02-28 Adapted to the separation of computing/printing
+%   2024-03-04 Allows user-defined solution of the final timestep
+
+% Load a representation of the Flak 36
+% flak36;
+d20;
+load shells
+param.drag=@(x)mcg6(x);
+
+% Specify muzzle elevation in degrees and convert to radian
+deg=0:1:90; theta=deg*pi/180;
+
+% Define the solver used to compute the last time step of each trajectory
+solver=@robust_secant; 
+cntl.delta=eps(1)/2; 
+cntl.eps=eps(1)/2;
+cntl.maxit=60;
+
+% Finally, produce the table of ranges
+[table, flag]=compute_range(param, v0, theta, method, dt, maxstep, solver, cntl);
+
+% ////////////////////////////////////////
+% Prepare for printing
+% ////////////////////////////////////////
+
+% Convert the elevation from radians to degrees
+data=[table(:,1)*180/pi,table(:,2),table(:,3)];
+
+% Obtain predefined parameters for formating the table
+tp=table_param('compute_range');
+
+% Print the table
+print_table(data,tp);
+
+rf=@(theta)range_rkx(param, v0, theta, method, dt, maxstep, solver, cntl);

### 2024-05-03 report/matlab/ballistics/mwe/d20_mwe1.m

@@ -1,0 +1,112 @@
+% Compute the range of shells of different types
+
+% Clear the work space
+clear;
+
+% Load the gun
+d20;
+
+% Load all shells
+load shells
+
+% Set the number of shell types
+numshell=1;
+
+% Set the method used to integrate the trajectory
+method='rk1';
+
+% Set the order of the method
+p=1;
+
+% Appropriate file name
+fname="ballistics/mwe/data/d20_mwe1.mat";
+
+% Create a list of drag functions
+f=cell(numshell,1);
+f{1}=mcg7;
+f{2}=mcg2;
+f{3}=mcg5;
+f{4}=mcg6;
+f{5}=mcg1;
+f{6}=mcg8;
+
+tit=cell(numshell,1);
+tit{1}='G7';
+tit{2}='G2';
+tit{3}='G5';
+tit{4}='G6';
+tit{5}='G1';
+tit{6}='G8';
+
+% Set the solver for the final timestep
+solver=@robust_secant;
+
+% Set the number of approximations
+kmax=10;
+
+% Tolerance used to control the final solve
+tol=2.^(-53); 
+
+% Check if the raw data exists
+if ~isfile(fname)
+    % Mark the start time
+    rt=cputime;
+    disp("Generating raw data");
+    % Allocate space for the output
+    raw_range=zeros(kmax,numshell);
+    raw_theta=zeros(kmax,numshell);
+
+    % Set the tolerance for the final solve
+    cntl.eps=tol; % eps(1)/2;
+    cntl.delta=tol; % eps(1)/2;
+    cntl.maxit=10;
+
+    % Loop over the different shells
+    for j=1:numshell
+        % Select the drag coefficient
+        param.drag=f{j};
+        % Set the time step and maxstep
+        dt=4; maxstep=100;
+        % Loop over the time step size
+        for i=1:kmax
+            % Display progress indicator
+            disp([(j-1)*kmax+i numshell*kmax]);
+            % Define the range function
+            rf=@(theta)range_rkx(param, v0, theta, method, dt,...
+                maxstep, solver, cntl);
+            % Find the largest range
+            [theta,r]=gss(rf,0,pi/2,eps(1)*pi/4,53);
+            % Save the range
+            raw_range(i,j)=r;
+            raw_theta(i,j)=theta;
+            % Reduce the timestep and increase maxstep
+            dt=dt/2; maxstep=maxstep*2;
+        end
+    end
[…]

### 2024-05-03 report/matlab/ballistics/mwe/range_rk1_mwe1.m

@@ -1,0 +1,52 @@
+% Computes the range of a shell fired from the Flak-36
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2018-11-01  Adapted from existing script and including a plot
+%   2024-03-04  Reformatted during yearly review
+
+% Load a model of G7 type shell
+load shells mcg7
+
+% ////////////////////////////////////////
+% Set all parameters (SI-units)
+% ////////////////////////////////////////
+
+% This is the mass of the shell
+param.mass=10;
+
+% This is the caliber of the barrel
+param.cali=0.088;
+
+% This function controls the size of the friction
+param.drag=@(x)mcg7(x);
+
+% The standard model of Earth's atmosphere
+param.atmo=@(x)atmosisa(x);
+
+% The standard acceleration due to gravity
+param.grav=@(x)9.80665;
+
+% Set the muzzle velocity and elevation
+v0=780; theta=45*pi/180; 
+
+% Select the basic time step size 
+dt=1; 
+
+% Set the maximum number of time steps
+maxstep=100;
+
+% Compute the range of the shell
+[r, flag, t, tra]=range_rk1(param,v0,theta,...
+    dt,maxstep);
+
+% Plot the trajectory nicely
+plot(tra(1,:),tra(2,:)); 
+
+% Adjust axis to use the same scale
+axis equal;
+
+% Add a coarse and a fine grid
+grid on; grid minor;
+
+% Print the range
+fprintf('Range of shell = %.2f meter\n',r);

### 2024-05-03 report/matlab/ballistics/range_rk1.m

@@ -1,0 +1,119 @@
+function [r, flag, t, tra]=range_rk1(param,v0,theta,dt,maxstep)
+
+% RANGE_RK1 Computes the range of a shell using Euler's explicit method
+%
+% All time steps except the last have the same size.
+% The last step is choosen to put the shell exactly on the ground.
+%
+% CALL SEQUENCE: [r, flag, t, tra]=range_rk1(param,v0,theta,dt,maxstep)
+%
+% INPUT:
+%   param    a structure describing of the environment and the shell
+%                param.mass   the mass of the shell
+%                param.cali   the caliber of the shell
+%                param.drag   a function computing the drag coeffient
+%                param.atmo   a function computing the atmosphere
+%                param.grav   a function computing gravity
+%                param.wind   a function computing the wind
+%   v0       the muzzle velocity of the shell
+%   theta    the elevation of the gun in radians
+%   dt       the standard time step, the last step will be shorter
+%   maxstep  the maximum number of time steps allowed
+%
+% OUTPUT:
+%    r       the computed range if flag=1, see below.
+%    t       the time instances where the trajectory was sampled
+%    tra     the computed trajectory, tra(:,i) corresponds to time t(i)
+%               tra(1,i) is the x-component of the shells position
+%               tra(2,i) is the y-component of the shells position
+%               tra(3,i) is the x-component of the shells velocity
+%               tra(4,i) is the y-component of the shells velocity
+%    flag    flag=0 if the shell did not hit the ground
+%            flag=1 if the shell hit the ground
+%
+% MINIMAL WORKING EXAMPLE: range_rk1_mwe1.m
+%
+% See also: COMPUTE_ELEVATION, COMPUTE_RANGE, RANGE_RKX
+
+% PROGRAMMING: Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  Fall 2014   Initial programming and testing.
+%  2015-09-22  Global variables m, k, and g integrated into a structure
+%  2015-10-31  The structure was replaced by a variable PARAM.
+%  2015-10-31  Documentation and variable names improved.
+%  2016-06-22  Minor changes to the documentation, comments and formatting
+%  2016-09-09  Adapted to SHELL4A. Fixed typo in MWE.
+%  2016-10-26  Edited text in MWE to be compatible with call.
[…]

### 2024-05-03 report/matlab/ballistics/shells/make_shells.m

@@ -1,0 +1,74 @@
+% MAKE_SHELLS  Build drag functions for standard shells
+
+% Defines the drag coefficient functions 
+% for G1, G2, G5, G5, G8, G7 and G8 type shells
+%
+% The raw data has been downloaded from this address
+% https://jbmballistics.com/ballistics/downloads/downloads.shtml
+% 
+% The original source of this data is the late Robert L. McCoy
+% of US Army's Ballistics Research Laboratory
+%
+% The abbreviation McG refers to McCoy and the
+% Gâvre Commission who defined the standard shells
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   201X-YY-ZZ Initial programming and testing
+%   2024-03-04 Reformatted during yearly review
+
+% Load the ascii file
+A=importdata('mcg1.txt');
+
+% Construct the corresponding spline
+cs=spline(A(:,1),A(:,2));
+mcg1=@(x)ppval(cs,x);
+
+% ////////////////////////////////////////
+
+% Load the ascii file
+A=importdata('mcg2.txt');
+
+% Construct the corresponding spline
+cs=spline(A(:,1),A(:,2)); 
+mcg2=@(x)ppval(cs,x);
+
+% ////////////////////////////////////////
+
+% Load the ascii file
+A=importdata('mcg5.txt');
+
+% Construct the corresponding spline
+cs=spline(A(:,1),A(:,2)); 
+mcg5=@(x)ppval(cs,x);
+
+% ////////////////////////////////////////
+
+% Load the ascii file
+A=importdata('mcg6.txt');
+
+% Construct the corresponding spline
+cs=spline(A(:,1),A(:,2)); 
+mcg6=@(x)ppval(cs,x);
+
+% ////////////////////////////////////////
+
+% Load the ascii file
+A=importdata('mcg7.txt');
+
+% Construct the corresponding spline
+cs=spline(A(:,1),A(:,2)); 
+mcg7=@(x)ppval(cs,x);
+
+% ////////////////////////////////////////
+
+% Load the ascii file
+A=importdata('mcg8.txt');
+
+% Construct the corresponding spline
+cs=spline(A(:,1),A(:,2)); 
+mcg8=@(x)ppval(cs,x);
+
+% ////////////////////////////////////////
+
+% Save the spline functions
+save shells.mat mcg1 mcg2 mcg5 mcg6 mcg7 mcg8

### 2024-05-03 report/matlab/ballistics/shells/plot_shells.m

@@ -1,0 +1,54 @@
+% PLOT_SHELLS  Plot drag functions for standard shells
+%
+% Plot the drag coefficient functions for the 
+% G1, G2, G5, G5, G8, G7 and G8 type shells
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   201X-YY-ZZ Initial programming and testing
+%   2024-03-04 Updated and reformatted
+
+% Load the shell functions
+load shells;
+
+% Define a range of Mach numbers
+ma=linspace(0,5,1001);
+
+% Get coordinates of the corners of the screen
+screen=get(groot,'Screensize'); 
+
+% Get width and height of the screen (pixels)
+sw=screen(3); sh=screen(4);
+ 
+% Get a handle to the current figure
+hFig=gcf; clf;
+
+% Set the size and position of the figure
+set(hFig,'Position',[sw/4 sh/4 sw/2 sh/2]);
+
+% ////////////////////////////////////////
+% Plot the drag coefficient functions 
+% ////////////////////////////////////////
+
+subplot(2,3,1); plot(ma,mcg1(ma)); 
+title('g1 drag coefficient'); axis([0,5,0,0.7]);
+xlabel('Mach number'); grid;
+
+subplot(2,3,2); plot(ma,mcg2(ma)); 
+title('g2 drag coefficient'); axis([0,5,0,0.7]);
+xlabel('Mach number'); grid;
+
+subplot(2,3,3); plot(ma,mcg5(ma)); 
+title('g5 drag coefficient'); axis([0,5,0,0.7]);
+xlabel('Mach number'); grid;
+
+subplot(2,3,4); plot(ma,mcg6(ma)); 
+title('g6 drag coefficient'); axis([0,5,0,0.7]); 
+xlabel('Mach number'); grid;
+
+subplot(2,3,5); plot(ma,mcg7(ma)); 
+title('g7 drag coefficient'); axis([0,5,0,0.7]); 
+xlabel('Mach number'); grid;
+
+subplot(2,3,6); plot(ma,mcg8(ma)); 
+title('g8 drag coefficient'); axis([0,5,0,0.7]); 
+xlabel('Mach number'); grid;

### 2024-05-03 report/matlab/ballistics/time_step_dependence.m

@@ -1,0 +1,47 @@
+% Shows that the computed trajectory depends on the time step
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-03-09 Initial programming and testing
+
+% Load a howitzer
+d20;
+
+% Set the time step sizes, maxstep and the elevation
+dt=[1 5 10 15 20]; maxstep=200; theta=45*pi/180;
+
+% ////////////////////////////////////////////////////////////////////
+%  Select solver for computing the final time step for each trajectory
+% ////////////////////////////////////////////////////////////////////
+solver=@robust_secant;
+
+% Control parameters for this solver
+cntl.delta=eps(1)/2;
+cntl.eps=eps(1)/2;
+cntl.maxit=20;
+
+% Compute and print several approximations of the trajectory
+fig=figure(); hold on;
+
+% Plot several trajectories and construct a legend
+n=numel(dt); lgd=cell(n,1);
+for i=1:n
+    % Compute trajectory using time step dt(i)
+    [r, flag, t, tra]=range_rkx(param, v0, theta, method, dt(i), maxstep, solver, cntl);
+    % Plot the computed trajectory
+    plot(tra(1,:),tra(2,:),'LineWidth',2);
+    % Construct a entry in the legend
+    lgd{i} = strcat('dt =',num2str(dt(i))) ;
+end
+
+% Labels and grids
+xl=xlabel('x (meter)'); xl.FontSize=18;
+yl=ylabel('y (meter)'); yl.FontSize=18;
+grid on; grid minor;
+
+% Legend
+leg=legend(lgd); 
+leg.FontSize=18;
+
+% Title
+tit=title('The computed trajectory depends on the time step');
+tit.FontSize=18

### 2024-05-03 report/matlab/differentiation/mwe/rdif_mwe1.m

@@ -1,0 +1,34 @@
+% RDIF_MWE1  Analyze 1st order accurate finite difference method
+
+% Set trial function
+f=@(x)exp(x);
+
+% Set point where we want to estimate the derivative
+x=1; 
+
+% Define the finite difference approximation
+D1=@(g,x,h)(g(x+h)-g(x))./h; 
+
+% Set the theoretical order of the method
+p=1; 
+
+% Set the basic stepsize
+h0=0.125; 
+
+% Set the number of times we will reduce the stepsize by a factor of 2
+kmax=25; 
+
+% Define the real derivative
+df=@(x)exp(x);
+
+% Apply Richardson's techniques
+data=rdif(f,x,D1,p,h0,kmax,df);
+
+% Obtain table parameters
+tp=table_param('rdif',data);
+
+% Print the information nicely
+print_table(data,tp);
+
+% Plot the evolution of Richardson's fraction
+fig=plot_fraction(data,p);

### 2024-05-03 report/matlab/extrapolation/plot_fraction.m

@@ -1,0 +1,26 @@
+function plt=plot_fraction(data,p)
+
+% Plot the evolution of Richardson's fraction
+%
+% CALL SEQUENCE: flag=plot_fraction(data)
+%     
+% INPUT:
+%   data   array returned by a call to richardson
+%   p      the order of the method
+% 
+% OUTPUT
+%   a handler to a figure plotting Richardson's fraction
+%
+% MINIMAL WORKING EXAMPLE: rdif_mwe1
+
+% PROGRAMMING by Carl Christian K. Mikkelsen (spock@cs.umu.se)
+%   2024-03-08 Extracted from old version of rdif_mwe1
+
+% Plot the development of Richardson's fraction
+plt=plot(data(:,1),log2(abs(2^p-data(:,3)))); 
+
+% Labels
+xlabel('k'); ylabel('log_{2}(abs(2^p - F_h))');
+
+% Grids
+grid on; grid minor;

### 2024-05-03 report/matlab/extrapolation/richardson.m

@@ -1,0 +1,73 @@
+function data=richardson(ah,p,tar)
+
+% Applies Richardson's techniques to a set of approximations
+%
+% CALL SEQUENCE: data=richardson(ah,tar)
+%   
+% INPUT:
+%   ah      a vector of approximations of the target value
+%
+%                 ah(j) = A(h(j))
+%
+%           where 
+%
+%                h(j+1) = 0.5*h(j)
+% 
+%  p        the order of the method used to computed ah
+%  tar      the target value (optional)
+%
+% OUTPUT:
+%   data    an m by n array where m = numel(ah) and 
+%             n=4  if tar is omitted 
+%             n=6  if tar is included
+%
+% MINIMAL WORKING EXAMPLE: rdif_mwe1
+%
+% SEE ALSO: rdif, rint, rode
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-03-07  Adapted from previous implementations from 5dv231
+%   2024-03-09  Improved documentation
+
+% Determine the number of approximations
+m=numel(ah);
+
+% Reshape the array
+ah=reshape(ah,m,1);
+
+% Allocate space for the output array
+n=4;
+if exist('tar','var')
+    % If the target value is known, then 
+    %   the error can be computed
+    %   the error estimate can be compared to the error
+    n=6;
+end
+data=zeros(m,n);
+
+% Number the rows of data
+data(:,1)=(1:m)';
+
+% Include the original approximations in data
+data(:,2)=ah;
+
+% Compute Richardson's fraction
+for i=3:m
+    data(i,3)=(ah(i-1)-ah(i-2))/(ah(i)-ah(i-1));
+end
+
+% Compute Richardson's error estimate
+aux=2^p-1;
+for i=2:m
+    data(i,4)=(ah(i)-ah(i-1))/aux;
+end
+
+% Is the target value known?
+if (n==6)
+    % Compute the actual error
+    data(:,5)=tar-ah;
+    % View Richardson's error estimate as an approximation
+    % of the actual error and compute the corresponding 
+    % relative error
+    data(:,6)=(data(:,5)-data(:,4))./data(:,5);
+end

### 2024-05-03 report/matlab/integration/mwe/rint_mwe3.m

@@ -1,0 +1,40 @@
+% Apply Richardson's techniques to integrating a nonsmooth function
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-04-09  Finalized during yearly review
+
+% Interval
+a=0; b=1; 
+
+% Function
+f=@(x)sqrt(x); 
+
+% Anti-derivative
+F=@(x)(2/3)*x.^(3/2);
+
+% Integral
+val=F(b)-F(a);
+
+% Integration rule
+rule=@(y,a,b,N)trapezoid(y,a,b,N); 
+
+% Theoretical order of the method 
+% The value below is wrong!!! The function is not smooth enough!
+% p=2;
+% The value below is correct!
+p = 1.5 
+
+% Number of refinements
+kmax=26; 
+
+% Apply Richardson's techniques
+data=rint(f,a,b,rule,p,kmax,val);
+
+% Obtain table parameters
+tp=table_param('rdif',data);
+
+% Print the information nicely
+print_table(data,tp);
+
+% Plot the evolution of Richardson's fraction
+fig=plot_fraction(data,p);

### 2024-05-03 report/matlab/integration/mwe/trapezoid_mwe1.m

@@ -1,0 +1,19 @@
+% Compute a composite trapezoidal sum for a simple function
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  2024-04-09  Extracted from trapezoid
+
+% Define the domain/interval
+a=0; b=1; 
+
+% Define the function
+f=@(x)exp(x); 
+
+% Define the sample points
+N=8; x=linspace(a,b,N+1); 
+
+% Evaluate the function values
+y=f(x);
+
+% Apply the composite rapezoidal rule
+s=trapezoid(y,a,b,N); 

### 2024-05-03 report/matlab/integration/trapezoid.m

@@ -1,0 +1,32 @@
+function s=trapezoid(y,a,b,N)
+
+% TRAPEZOID Composite trapezoidal rule for a function on an interval
+%
+% CALL SEQUENCE: s=trapezoid(y,a,b,N)
+%
+% INPUT:
+%   y      the function values on N+1 equidistant points on [a,b]
+%   a, b   the interval 
+%   N      determines the stepsize, h=(b-a)/N
+% 
+% OUTPUT:
+%  s      the trapezoidal sum. 
+%
+% MINIMAL WORKING EXAMPLE: trapezoid_mwe1
+%
+% See also: RINT
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  2015-XX-YY  Initial programming and testing
+%  2016-06-28  Reformatted during yearly review
+%  2024-04-09  Minimal working example moved to separate file
+
+% Compute the length of the subintervals
+h=(b-a)/N;
+
+% Compute the trapezoidal sum.
+s=0; 
+for i=1:N
+    s=s+(y(i)+y(i+1))*h/2;
+end
+

### 2024-05-03 report/matlab/nonlinear-equations/bisection.m

@@ -1,0 +1,188 @@
+function [x, flag, it, data]=bisection(f,sp)
+
+% Solves a nonlinear equation using bisection
+%
+% Attempts to solve the non-linear equation scalar equation
+%
+%   f(x) = 0
+%
+% using the bisection algorithm.
+%
+% CALL SEQUENCE:
+%
+%   [x, flag, it, data]=bisection(f,param)
+%
+% INPUT:
+%   f        a handler to the function f
+%   sp       a structure with the following fields
+%              xval     the initial bracket
+%              fval     the corresponding function values
+%              delta    break if abs(a(it)-b(it)) < delta
+%              eps      break if abs(res(it)) < eps
+%              maxit    break if it=maxit
+%
+% OUTPUT:
+%   x        the final approximation of the root
+%   flag     a flag signaling succes/failure, if flag = 
+%              -1,  then the initial bracket was bad
+%               0,  then the iteration did not converge
+%               1,  then the last bracket has length less than delta
+%               2,  then the last residual has size less than eps
+%               3,  then the two previous conditions are both true
+%   it       the number of iterations completed
+%   data     an it by 5 array
+%               data(j,1)   = j
+%               data(j,2:3) = [a(j) b(j)], the jth bracket
+%               data(j,4)   = his(j), the jth approximation of x
+%               data(j,5)   = res(j), the the corresponding residual
+%
+% A bad bracket is a pair of points a, b, such that f(a)*f(b) >= 0. In this
+% case, we can not conclude that there is a root strictly between a and b. 
+%
+% MINIMAL WORKING EXAMPLE: bisection_mwe1
+%
+% See also: NEWTON, SECANT, ROBUST_SECANT
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  2014-05-12  Initial programming and testing
+%  2014-10-20  Call sequence and comments improved
+%  2014-11-12  All arguments are now assigned values regardless of flag
+%  2014-11-16  Program rewritten from scratch to clarify the logic
+%  2015-09-22  Minor improvements to the documentation and comments
+%  2016-06-14  Minor improvements to the documentation and comments
[…]

### 2024-05-03 report/matlab/nonlinear-equations/newton.m

@@ -1,0 +1,92 @@
+function [x, flag, it, his, res, resnorm]=newton(f,sp)
+
+% Solves a system of nonlinear equations using Newton's method
+%
+% An implementation of Newton's method for solving a non-linear equation
+%
+%        f(x) = 0
+%
+% with an initial guess x=x0. 
+%
+% CALL SEQUENCE:
+%
+% [x, flag, it, his, res, resnorm]=newton(f,param)
+%
+% INPUT:
+%   f         a handler to the function f
+%   sp        a structure with the following fields
+%               sp.xval   the initial value
+%               sp.eps    tolerance for absolute residual
+%               sp.maxit  maximum number of iterations
+%               sp.df     handler to the Jacobian
+%
+% OUTPUT:
+%   x         the final approximation
+%   flag      a flag signaling succes/failure,
+%               flag = -1, if we did not converge
+%               flag =  1, if we stagnated
+%               flag =  2, if we have a small residual
+%               flag =  3, if flag=1 and flag=2
+%   it        the number of iterations completed
+%   his       his(j) is the jth approximation
+%   res       res(j) is the jth residual
+%   resnorm   resnorm(j) is the norm of the jth residual
+%
+% MINIMAL WORKING EXAMPLE(s) newton_mwe1, newton_mwe2
+% 
+% See also: BISECTION, SECANT, ROBUST_SECANT
+
+% PROGRAMMING by Carl Christian K. Mikkelsen (spock@cs.umu.se)
+%  2014-10-20  Call sequence determined
+%  2014-10-21  Initial implementation and testing
+%  2014-11-20  More comments added in response to student questions
+%  2015-10-16  Documentation edited
+%  2016-06-27  Documentation edited. Added links to other methods
+%  2016-06-27  Change code to conform to standard
+%  2016-08-15  Minor changes to the documentation
+%  2018-11-23  Move the minimal working examples to separate files
+%  2018-11-23  Changed return codes
+%  2020-11-01  Minor update of documentation
+%  2024-04-03  Adapted to the standardized interface for all solvers
+
+% Extract parameters
+x0=sp.xval;
+eps=sp.eps;
+maxit=sp.maxit;
+df=sp.df;
+
+% Newton's method is tricky to apply, so flag=-1 is the default
+flag=-1;
+
+% Extract the dimension of the problem
+n=numel(x0);
+
+% Allocate space for approximations, residuals, and their norms
[…]

### 2024-05-03 report/matlab/nonlinear-equations/robust_secant.m

@@ -1,0 +1,205 @@
+function [x, flag, it, data]=robust_secant(f,sp)
+
+% Solves a nonlinear equation using a bisection/secant hybrid
+%
+% Attempts to solve the nonlinear scalar equation
+%
+%   f(x) = 0
+%
+% using a hybrid between the bisection method and the secant method. This 
+% method combines the robustness of the bisection method with the rapid 
+% local convergence of the secant method. The function requires a bracket
+% around the root and computes a sequence of brackets of decreasing length.
+% Secant steps are accepted only if they do not leave the current bracket. 
+% If a secant step is impossible or if it would leave the current bracket
+% a bisection step is completed instead.
+% 
+% CALL SEQUENCE:
+%
+%   [x, flag, it, data]=robust_secant(f,sp)
+%
+% INPUT:
+%   f        a handler to the function f
+%   sp       a structure with the following fields
+%              xval     the initial bracket
+%              fval     the corresponding function values
+%              delta    break if abs(a(it)-b(it)) < delta
+%              eps      break if abs(res(it)) < eps
+%              maxit    break if it=maxit
+%
+% OUTPUT:
+%   x        the final approximation
+%   flag     a flag signaling succes/failure, if flag = 
+%              -1,  then the initial bracket was bad, see below
+%               0,  then the iteration did not converge
+%               1,  then the last bracket has length less than delta
+%               2,  then the last residual has size less than eps
+%               3,  then the two previous conditions are both true
+%   data     an array of dimension it by 6 datastructure contain these fields
+%               data(j,1)   = j
+%               data(j,2:3) = [a(j) b(j)], the jth bracket
+%               data(j,4)   = his(j), the jth approximation of x
+%               data(j,5)   = res(j), the the corresponding residual
+%               data(j,6) 
+%                    = 1, if a secant step was completed
+%                    = 2, if a secant step would leave the bracket
+%                    = 3, if a secant step would cause division by zero               
+%
+% A *bad* bracket is a pair of points a, b, such that f(a)*f(b) >= 0. In 
+% such a case, we can not conclude that there is a root strictly between 
+% a and b. 
+% 
[…]

### 2024-05-03 report/matlab/ode/aux/phi1.m

@@ -1,0 +1,23 @@
+function z=phi1(f,t,y,dt)
+
+% PHI1 Kernel for Euler's explicit method of order 1
+%
+% Advances the solution of the ODE 
+%
+%     y'(t) = f(t,y(t))
+%
+% a single time step using Euler's explicit method.
+%
+% See also: RK, PHI2, PHI3, PHI4
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se) 
+%  Apr 2014   :  Initial coding
+%  Oct 2014   :  Minor changes to the text.
+%  2015-06-23 :  Reformated during yearly review.
+
+% Compute the auxiliary coefficient
+k1=f(t,y); 
+
+% Advance the solution a single time step
+z=y+dt*k1;
+

### 2024-05-03 report/matlab/ode/aux/phi2.m

@@ -1,0 +1,24 @@
+function z=phi2(f,t,y,dt)
+
+% PHI2 Kernel for a second order explicit Runge-Kutta method
+%
+% Advances the solution of the ODE
+%
+%     y'(t) = f(t,y(t))
+%
+% a single time step using Heyn's method.
+%
+% See also: RK, PHI1, PHI3, PHI4
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se) 
+%  Apr 2014   :  Initial coding and testing.
+%  Oct 2014   :  Minor changes to the text.
+%  2015-06-23 :  Reformated during yearly review.
+
+% Compute auxiliary coefficients
+k1=f(t,y); 
+k2=f(t+dt,y+dt*k1);
+
+% Advance the solution a single step
+z=y+0.5*dt*(k1+k2);
+

### 2024-05-03 report/matlab/ode/aux/phi3.m

@@ -1,0 +1,25 @@
+function z=phi3(f,t,y,dt)
+
+% PHI3 Kernel for an explicit third order Runge Kutta method 
+%
+% Advances the solution of the ODE
+%
+%     y'(t) = f(t,y(t))
+%
+% a single time step using a Runge-Kutta method of order 3.
+%
+% See also: RK, PHI1, PHI2, PHI4
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se) 
+%  Apr 2014   :  Initial coding and testing.
+%  Oct 2014   :  Minor changes to the text.
+%  2015-06-23 :  Reformated during yearly review.
+
+% Compute auxiliary coefficients
+k1=f(t,y);  
+k2=f(t+0.5*dt,y+0.5*dt*k1);
+k3=f(t+dt,y-dt*k1+2*dt*k2);
+
+% Advance the solution a single step
+z=y+(dt/6)*(k1+4*k2+k3);
+

### 2024-05-03 report/matlab/ode/aux/phi4.m

@@ -1,0 +1,26 @@
+function z=phi4(f,t,y,dt)
+
+% PHI4 Kernel for the classical explicit fourth order Runge Kutta method
+%
+% Advances the solution of the ODE
+%
+%     y'(t) = f(t,y(t))
+%
+% a single time step using the classical Runge-Kutta method of order 4.
+%
+% See also: RK, PHI1, PHI2, PHI3
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se) 
+%  Apr 2014   :  Initial coding and testing.
+%  Oct 2014   :  Minor changes to the text.
+%  2015-06-23 :  Reformated during yearly review.
+
+% Compute auxiliary coefficients
+k1=f(t,y);  
+k2=f(t+0.5*dt,y+0.5*dt*k1);
+k3=f(t+0.5*dt,y+0.5*dt*k2);
+k4=f(t+dt,y+dt*k3);
+
+% Advance the solution a single step
+z=y+(dt/6)*(k1+2*k2+2*k3+k4);
+

### 2024-05-03 report/matlab/ode/event_location.m

@@ -1,0 +1,211 @@
+function [t, y, count, idx]=event_location(f,t0,t1,y0,maxstep,method,g,K)
+
+% Event location for ordinary differential equations
+%
+% Solve the differential equation
+%
+%   y'(t) = f(t,y(t)),
+%   y(t0) = y0
+%
+% while searching for events of the type
+%
+%   g(y(t)) = 0 (*)
+%
+% where g is a real value function. 
+%
+% This function is based on the RK function and the call sequence is very 
+% similar.
+%
+% CALL SEQUENCE:
+%
+%    [t, y, count, idx]=event_location(f,t0,t1,y0,N,method,g,K)
+%
+% INPUT:
+%   f         a handler to the function f, see below.
+%   t0, t1    the time interval is [t0,t1]          
+%   y0        is the initial condition, i.e. y(t0) = y0
+%   maxstep   the basic time step is (t1-t0)/N
+%   method    a string which specifies the method
+%               'rk1' : the explicit Euler method
+%               'rk2' : Heun's method aka the improved Euler method
+%               'rk3' : a third order accurate Runge-Kutta method.
+%               'rk4' : the classical 4th order Runge-Kutta method.
+%   g         a handler to a real valued function
+%   K         return at most K solutions of (*)
+% 
+% OUTPUT:
+%    t      a vector containing at most N+K+1 points in time
+%    y      y(:,i) is the approximation of the solution at time t(i).
+%    count  the number of events recorded
+%    idx    the indices in t and y of the events recorded
+%
+% NOTES:
+%   a) It is assumed that the function f returns a COLUMN vector!
+%
+% MINIMAL WORKING EXAMPLE: event_location_mwe1
+%
+% See also: RK
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se) 
+%  2015-09-21  Initial programming
+%  2015-09-22  Improved inline documentation
+%  2015-11-22  Extended call sequence and added minimum working example
+%  2016-05-22  Improved inline documentation and formatting
+%  2016-05-22  Adapted routine to improved bisection method
+
+% TODO: 
+% The game SCORCHED reveals that there are unresolved problems as shells can
+% pass straight through the narrow hills. The problem is that the event
+% function has the same sign on both sides of the hill, so there is no sign
+% change to trigger a search for a zero. 
+%   1) Obtain a smooth representation of the solution between data points
+%   2) Determine the range of the event function between data points
[…]

### 2024-05-03 report/matlab/ode/mwe/event_location_mwe1.m

@@ -1,0 +1,40 @@
+% Find the point of impact of a shell fired from a gun
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-03-07 Initial programing and testing
+
+% Load the gun
+d20;
+
+% Set the elevation of the barrel of the gun
+theta=74*pi/180; 
+
+% Redefine the basic time step for a clean graph
+dt=5;
+
+% Define the initial condition
+x0=[0; 0; v0*cos(theta); v0*sin(theta)];
+    
+% Define the function that drives the differential equation
+f=@(t,x)shell4a(param,t,x);
+
+% Define the time interval
+t0=0; t1=maxstep*dt;
+
+% Define the event function, i.e., isolate the y coordinate
+g=@(x)x(2);
+
+% Set the maximum number of events
+K=2;
+
+% Fire the shell and search for the point of impact
+[t, y, count, idx]=event_location(f,t0,t1,x0,maxstep,method,g,K);
+
+% Plot the entire trajectory
+plot(y(1,:),y(2,:),y(1,:),y(2,:),'*');
+
+% Labels
+xlabel('x (meter)'); ylabel('y (meter)');
+
+% Grids
+grid on; grid minor

### 2024-05-03 report/matlab/ode/mwe/rode_mwe2.m

@@ -1,0 +1,38 @@
+% Demonstrate Richardson's technique for a simple 2D problem
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-03-07 Adapted to the use of print_rode_table
+
+% Define the interval
+a=0; b=1;
+
+% Select the function which defines the differential equation
+f=@(t,y)[-y(2); y(1)]; 
+
+% Specify the initial condition
+y0=[1; 0];
+
+% Select the number of timesteps
+N1=10; N2=10; 
+
+% Select the number of approximations
+kmax=7;
+
+% Specify the method to use and the *correct* order
+method='rk2'; p=2; 
+
+% The exact solution can be computed 
+sol=@(t)[cos(t); sin(t)];
+
+% Apply Richardson's techniques to our initial value problem
+[s, table]=rode(f,a,b,y0,N1,N2,kmax,method,p,sol);
+
+% These parameters determine the layout of the table
+aux.kmax=kmax;
+aux.n=size(table,2);
+
+% Obtain parameters suitable for printing
+tp=table_param('rode',aux);
+
+% Print the table(s) nicely
+print_rode_table(table,tp);

### 2024-05-03 report/matlab/ode/mwe/rode_mwe3.m

@@ -1,0 +1,44 @@
+% Demonstrate Richardson's technique for a simple artillery trajectory
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-03-07 Adapted to the use of print_rode_table
+
+% Select mass, CONSTANT drag coefficient function and constant gravity
+param=struct('mass',10,'cali',0.088,'drag',@(x)0.1879,'atmo',@(x)atmosisa(0),'grav',@(x)9.82);
+
+% Select a slight head wind (swedish: motvind, tail wind = medvind)
+param.wind=@(t,x)[-5; 0];
+
+% Set the muzzle velocity and elevation
+v0=780; theta=45*pi/180;
+
+% Define the initial condition
+y0=[0; 0; v0*cos(theta); v0*sin(theta)];
+
+% Select the function which defines the differential equation
+f=@(t,y)shell4a(param,t,y);
+
+% Define the time interval we wish to investigate
+a=0; b=50;
+
+% Select the number of timesteps
+N1=10; N2=4; 
+
+% Select the number of approximations
+kmax=5;
+
+% Specify the method to use and the *correct* order
+method='rk4'; p=4; 
+
+% Apply Richardson's technique to our initial value problem
+[s, table]=rode(f,a,b,y0,N1,N2,kmax,method,p);
+
+% These parameters determine the layout of the table
+aux.kmax=kmax;
+aux.n=size(table,2);
+
+% Obtain parameters suitable for printing
+tp=table_param('rode',aux);
+
+% Print the table(s) nicely
+print_rode_table(table,tp);

### 2024-05-03 report/matlab/ode/rk.m

@@ -1,0 +1,89 @@
+function [t, y]=rk(f, t0, t1, y0, N1, N2, method)
+
+% Solves an IVP for an ODE using an explicit Runge-Kutta method
+%
+% An implementation of some common methods for the initial value problem
+%
+%   y'(t) = f(t,y(t)),
+%   y(t0) = y0
+%
+% The routine can handle systems of ODE, so f and y can be vectors.
+%
+% CALL SEQUENCE: [t y]=rk(f,t0,t1,y0,N1,N2,method)
+%
+% INPUT:
+%   f         a handler to the function f, see below.
+%   t0, t1    the time interval is [t0,t1]          
+%   y0        is the initial condition, i.e. y(t0) = y0
+%   N1, N2    the total number of time steps is N1*N2, see below.
+%   method    a string that specifies the method
+%               'rk1'     : the explicit Euler method
+%               'rk2'     : Heun's method aka the improved Euler method
+%               'rk3'     : a third order accurate Runge-Kutta method.
+%               'rk4'     : the classical 4th order Runge-Kutta method.
+%               'testphi' : user-defined method 
+% OUTPUT:
+%    t    a vector of length N1+1 such that t(i)=t0+(i-1)*h.
+%    y    y(:,i) is the approximation of the solution at time t(i).
+%
+% NOTES:
+% a) It is assumed that the function f returns a COLUMN vector!
+% b) We compute N=N1*N2 time steps
+% c) We retain only N1 equally spaced samples
+% d) The stepsize dt used to compute the solution is
+% 
+%      dt = length of interval / total number of steps = (t1-t0)/N
+%
+% MINIMAL WORKING EXAMPLE: rk_mwe1, rk_mwe2
+%
+% See also: PHI1, PHI2, PHI3, PHI4
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se) 
+%   Apr 2014    Initial programming.
+%   Oct 2014    Minor streamlining. 
+%   2014-12-18  A third order Runge-Kutta method has been added
+%   2014-12-18  Documentation reformatted to conform with the standard
+%   2024-03-09  Updated the documentation during the yearly review
+ 
+% Force the initial condition to be a COLUMN vector; 
+s=numel(y0); y0=reshape(y0,s,1); 
+
+% Define the total number of steps N and the stepsize dt;
[…]

### 2024-05-03 report/matlab/ode/rode.m

@@ -1,0 +1,145 @@
+function [s, table]=rode(f,a,b,y0,N1,N2,kmax,method,p,sol)
+
+% RODE  Applies Richardson's techniques to ODEs
+%
+% This function uses Richardson's technique to estimate the error when
+% solving the initial value problem
+%
+%    y'(t) = f(t, y),     a <= t <=b
+%     y(a) = y0
+%
+% using one of the standard Runge-Kutta methods.
+%
+% CALL SEQUENCE: 
+%
+%   [s, table]=rode(f, a, b, y0, N1, N2, kmax, method, p, sol)
+%
+% INPUT:
+%   f       a user defined function
+%   a,b     specifies the interval
+%   y0      the initial condition
+%   N1, N2  controls the total number of timesteps, see below
+%   kmax    controls the number of refinements, see below
+%   method  a string controlling the method used, see rk.m
+%   p       the order of the method used
+%   sol     an optional function which computes the true solution.
+%             
+% OUTPUT:
+%   s       s(i,j,k) is the kth approxmation of the i component of y(t(j))
+%   table   tables containing all information extracted by richardson,
+%           suitable for printing with print_ode_table
+%
+% MINIMAL WORKING EXAMPLE(s): rode_mwe1, rode_mwe2, rode_mwe3, rode_mwe4
+% 
+% See also: RDIF, RINT, RK, PRINT_RODE_TABLE
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%    Dec 2012    Initial programming and testing
+%    Apr 2014    Streamlining and better comments
+%    Oct 2014    Modified to fit with the new Runge-Kutta code
+%    2014-29-10  Optional exact solution enabled
+%    2014-12-04  Variable number of fractions allowed to check behaviour
+%    2014-12-05  Improved documentation prior to release of Project 3
+%    2014-12-16  Bug found and fixed by several students in MWE
+%    2016-06-28  Major revision during yearly review
+%                  1) New minimal working examples
+%                  2) New documentation
+%                  3) Integration of displaytable()
+%                  4) Automatic support for multiple dimensions
+%                  5) Improved datalayout
+%    2022-12-12  Trimmed the output to fit smaller screens.
+%    2024-03-08  Simplification during yearly review
+%                  1) The analysis of s has been moved to richardson
+%                  2) The printing of tables moved to print_rode_table
+
+flag=0;
[…]

### 2024-05-03 report/matlab/tables/table_param.m

@@ -1,0 +1,190 @@
+function tp=table_param(func, param)
+
+% Returns parameters for printing tables
+% 
+% CALL SEQUENCE: tp=table_param(func, param)
+% 
+% INPUT:
+%   func    a string naming the function that generated the data
+%   param   a parameter that may or may not be necessary
+%
+% OUTPUT:
+%   tp      parameters suitable for printing the table using
+%             print_table
+%
+% MINIMAL WORKING EXAMPLE: rdif_mwe1
+%
+% SEE ALSO: print_table, print_rode_table
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%   2024-03-XX Initial programming and testing
+
+% Dummy initialization of all parameters
+tp.headers=[];
+tp.wids=[];
+tp.fms=[];
+tp.rowheadings=[];
+tp.fid=1;
+tp.colsep=[];
+tp.rowending=[];
+
+switch lower(func)
+    case {'bisection'}
+        % Set column headers
+        headers = {'iter', 'a','b','approximation','residual'};
+        
+        % Set column widhs
+        wids=[6 22 22 22 22];
+        
+        % Set column formats
+        fms={'d','.15e','.15e','.15e','.15e'};
+        
+    case {'compute_range'}
+        % Define column headers
+        headers = {'Elevation (deg)','Range (m)',...
+            'Flight time (s)'};
+        
+        % Set the widths of the columns
+        wids=[15 15 15];
+        
+        % Define the format specification
+        fms={'.2f','.2f','.2f'};
+        
+    case {'compute_elevation'}
+        % Define column headers
+        headers = {'Distance (m)',...
+            'Low (deg) ','Flight time (s)',...
+            'High (deg)','Flight time (s)'};
+        
+        % Set the widths of the columns
+        wids=[12 10 15 10 15];
+        
+        % Define the format specification
+        fms={'.2f','.2f','.2f','.2f','.2f'};
+        
+    case {'dif0'}
+        % Set column headers
+        headers={'k','h','d=f(x+h)-f(x)','r=d/h'};
+        
+        % Set column widths
+        wids=[6 20 20 20];
+        
+        % Set column formats
+        fms={'d','.12e','.12e','.12e'};
+        
+    case {'harm_sum'}
+        
+        switch param
+            case 0
+                % Set column headings
+                headers={'','raw data', 'successive differences'};
+                
+                % Set column widths
+                wids=[4 53 61];
+                
[…]

### 2024-05-03 report/scripts/compile.sh

@@ -1,0 +1,4 @@
+cd src
+pdflatex draft.tex
+pdflatex draft.tex
+cd ..

### 2024-05-03 report/scripts/gatherfig.sh

@@ -1,0 +1,3 @@
+find experiments -name \*.png -exec cp {} fig \;
+find experiments -name \*.eps -exec cp {} fig \;
+

### 2024-05-07 report/src/draft.tex
Manuscript is shaping up. :)

@@
 Our data and software is freely available from our GitHub along with every script needed to generate every number, table and figure from scratch.
+The names of MATLAB function shall be written using a typewriter font, ex.
+{\tt plot\_shells}.
 
@@
 We say that the error $E_h$ satisfies an asymptotic error expansion.
-Frequently, the exponents $(p,q,r)$ are all integers, but since we shall allow exponents that are not integers, we must insist that $h$ is strictly positive.
+Frequently, the exponents $(p,q,r)$ are all integers, but since we shall allow exponents that are not integers, we insist that $h$ is strictly positive.
 
@@
 Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T - \hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe1b}.
-We observe that computed value of Richardson's error estimate is a good approxmation of the error $T - \hat{A}_h$.
-
-% tiny scriptsize footnotesize small normalsize large Large LARGE
-
-% \begin{myverbbox}[\scriptsize]{\valpha}
-%  k |                A_h |         F_h |         R_h |         E_h | (E_h-R_h)/R_h
-%  0 | 1.859140914230e+00 |  0.00000000 |  0.0000e+00 | -1.4086e-01 |    1.0000e+00
-%  1 | 1.753931092465e+00 |  0.00000000 | -3.5070e-02 | -3.5649e-02 |    1.6251e-02
-%  2 | 1.727221904558e+00 |  3.93908726 | -8.9031e-03 | -8.9401e-03 |    4.1402e-03
-%  3 | 1.720518592164e+00 |  3.98447608 | -2.2344e-03 | -2.2368e-03 |    1.0400e-03
-%  4 | 1.718841128580e+00 |  3.99610010 | -5.5915e-04 | -5.5930e-04 |    2.6031e-04
-%  5 | 1.718421660316e+00 |  3.99902383 | -1.3982e-04 | -1.3983e-04 |    6.5098e-05
-%  6 | 1.718316786850e+00 |  3.99975588 | -3.4958e-05 | -3.4958e-05 |    1.6276e-05
-%  7 | 1.718290568083e+00 |  3.99993897 | -8.7396e-06 | -8.7396e-06 |    4.0690e-06
-%  8 | 1.718284013367e+00 |  3.99998474 | -2.1849e-06 | -2.1849e-06 |    1.0167e-06
-%  9 | 1.718282374686e+00 |  3.99999619 | -5.4623e-07 | -5.4623e-07 |    2.5759e-07
-% 10 | 1.718281965016e+00 |  3.99999904 | -1.3656e-07 | -1.3656e-07 |    6.5583e-08
-% 11 | 1.718281862598e+00 |  3.99999982 | -3.4139e-08 | -3.4139e-08 |    8.2385e-08
-% 12 | 1.718281836994e+00 |  3.99999982 | -8.5348e-09 | -8.5348e-09 |    1.4743e-07
-% 13 | 1.718281830593e+00 |  3.99999566 | -2.1337e-09 | -2.1337e-09 |   -3.7464e-06
-% 14 | 1.718281828992e+00 |  3.99999681 | -5.3343e-10 | -5.3342e-10 |   -1.8177e-05
-% 15 | 1.718281828592e+00 |  4.00004551 | -1.3335e-10 | -1.3335e-10 |   -2.7197e-05
-% 16 | 1.718281828492e+00 |  4.00022202 | -3.3337e-11 | -3.3341e-11 |    1.1322e-04
-% 17 | 1.718281828467e+00 |  4.00046186 | -8.3333e-12 | -8.3409e-12 |    9.1400e-04
-% 18 | 1.718281828461e+00 |  3.98094194 | -2.0933e-12 | -2.0610e-12 |   -1.5658e-02
-% 19 | 1.718281828460e+00 |  4.13782004 | -5.0589e-13 | -5.4334e-13 |    6.8928e-02
-% 20 | 1.718281828459e+00 |  3.95086705 | -1.2805e-13 | -1.5921e-13 |    1.9572e-01
-% \end{myverbbox}
-
-% \begin{figure}
-%   \centering
-%   \valpha
-%   \caption{Numerical integration of a function that is everywhere smooth.} \label{fig:rint_mwe1}
-% \end{figure}
+We observe that computed value of Richardson's error estimate is a good approxmation of the error $T - \hat{A}_{h_h}$.
+In fact, the corresponding relative error decreases when $k$ increases and we remain inside the asymptotic region.
 
@@
 
-We mention in passing that low order methods are more pratical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
+We mention in passing that low order methods are more practical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
 This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small and this issue is more acute for high order methods than for low order methods.
@@
 In this section we present the results of more elaborate examples that highlight the practical limitations of Richardson's extrapolation as well as some limitations of contemporary software for doing molecular dynamics.
-Every function or script necessary to replicate the numbers needed to reproduce the figures is freely available from the authors GitHub.
 
@@
 
-The script {\tt maxrange\_rk1} models a shells point particle moving in a plane subject to Earth's standard gravity and the international standard atmosperic model.
+The script {\tt maxrange\_rk1} models a shell as a point particle moving in a plane subject to Earth's standard gravity and the international standard atmosperic model.
+This is a crude model that is suitable only for short range artillery, because it does not account for the rotation of the Earth or the oriention and spin of the projectile.
 Each trajectory is integrated using Euler's explicit method ({\tt 'rk1'}) and all but the final step has the same size $h$.
@@
 The drag function are interpolated from tables using cubic spline interpolation.
+The function {\tt plot\_shells} will plot the drag coefficients for the different shells as a function of the Mach number.
 
@@
 For each drag coefficient and for each value of the time step $h$, a range function is defined which returns the range of the shell as a function of the howitzers elevation $\theta$.
-The range functions are unimodal and the maximum range is found using the golden section search algorithm.
-The initial search bracket is $[0,\pi/2$] and this bracket is systematically reduced in length until it is shorter than $\frac{pi}{2}u$.
+The range functions are unimodal and the maximum range is found using the golden section algorithm.
+The initial search bracket is $[0,\pi/2$] and this bracket is systematically reduced in length until it is shorter than $\frac{\pi}{2}u$.
 
@@
 In each case the error estimate suggests that the computed range is exact to number of figures shown.
-In particular, we see while a G7 type shell achieves a maximum range of 17.4 km, all other shells have ranges that are less than 16.9 km.
+In particular, we see that while a G7 type shell achieves a maximum range of 17.4 km, all other shells have ranges that are less than 16.9 km.
 However, it is a fallacy to conclude anything on the basis of the computed ranges and the associated error estimates.
-We need to assert that we are inside the asymptotic range and that the error estimates are reliable.
+In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable.
 To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}.
 For each of the 6 different drag coefficients we find that evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$.
-This result is entirely consistent with the use of Euler's explicit method.
-We observe that for each drag coefficient, $k=12$ is still inside the asymtotic range and we have no reason to doubt the quality of the error estimate.
-We conclude that the best shell model for the D-20 howitzer is the G7 shell.
+This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step.
+We observe that for each drag coefficient, $k=12$ is still inside the asymtotic range and we have no reason to doubt the magnitude of the error estimate.
+We conclude that the best model for the D-20 howitzer is the G7 shell.
 
@@
 
-In this section we present the experiment that triggered the production of the present paper.
-We attempted to apply Richardson's techniques in the context of molecular dynamics with constraints.
-It was a spectacular failure with no obvious pattern controlling the computed numbers.
-While we cannot claim to fully understand our failure, we have identified two key differences between the our two test cases and we have through a sequence of experiments verified that each difference is enough to destroy any hope of having an asymptotic error expansion.
-
-We utilized GROMACS v2021 to conduct experiments on the behavior of hen egg white lysozyme submerged in water within a cubic simulation box, following Justin Lemkul's Lysozyme in Water GROMACS Tutorial.
+We utilized GROMACS v2021 to conduct experiments on the behavior of hen egg white lysozyme submerged in water within a cubic simulation box, following Justin Lemkul's Lysozyme in Water GROMACS Tutorial \cite{}.
 Several steps were taken to prepare the system for production simulation: first, ions were introduced to achieve electrical neutrality.
@@
 The described process was replicated using two different force fields, OPLS-AA/L and CHARMM36.
-We conducted production simulations of 1 ps for both force fields, varying the number of executed steps from 250 to 16000 and employing two different tolerances for the SHAKE algorithm (1E-4 and 1E-12).
-For each experiment we computed the total kinetic and potential energy of the system at the end of the simulation.
+We conducted production simulations of 1 ps for both force fields, using $n$ steps to cover this interval where
+\begin{equation}
+  n \in \{250, 500, 1000, 1100:100:2000, 3000:1000:16000\}.
+\end{equation}
+Moreover, we used two different tolerances {\tt tol} for the SHAKE algorithm, namely
+\begin{equation}
+  \text{tol} \in \{10^{-4}, 10^{-12}\}.
+\end{equation}
+For each experiment, we computed the total kinetic and potential energy of the system at the end of the simulation.
+The function {\tt gromacs\_figures} will generate Figure~\ref{fig:oplsaaltol04} and Figure~\ref{fig:oplsaaltol12} for the OPLS-AA/L force-file and similar figures for CHARMM36.
+These figures display the mechanical energies of the molecule at the end of the simulation as a function of the total number $n$ of time steps used to cover the interval.
+The figures have several features of interest.
+
+\begin{enumerate}
+\item The kinitic energy exhibits violent oscilations for ${\tt tol} = 10^{-4}$.
+Rapid oscillations are also shown by the potential energy, but they are less extreme.
+The applitude of all oscillations is significantly reduced when the constraint equations are solved accurately, i.e., ${\tt tol} = 10^{-12}$.
+This reduction is not unexpected.
+\item Apart from the oscillations, the total energy exhibits what appears to be linear growth with the number of time steps.
+This is not surprising as we expect the rounding error to grown with the number of operations.
+The function {\tt gromacs\_fit} does a linear least squares fit of the total energy {\tt etol} to two models $M_1$ and $M_2$ where model $M_p$ assumes that ${\tt etol} = T + C_1n + C_2/n^p$ for suitable constants $T, C_1, C_2$.
+Since the SHAKE algorithm is 2nd order accurate in the time step, one would expect model $M_2$ to provide the best fit, but for this data model $M_1$ is a marginally better.
+While mildly interesting, this issue is vastly less important than the final point.
+\item If the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} in the asymptotic range.
+Why is this?
+If we where in the asymptotic range, then the computed energies $A_h$ would follows follow a power law, i.e.
+$A_h \approx T - \alpha h^p$ would be a good approximation for some power $p>0$.
+In particular, the behaviour would be \emph{monotone} and the tiny oscillations that we have recorded would not be present.
+\end{enumerate}
+
+We are left to make the following conclusions:
+\begin{enumerate}
+  \item If we do not solve the constraint equations accurately, then we cannot expect the computed energies to change slowly with the number of time steps.
+  \item If we solve the constraint equations accurately, then there is no reason to believe that we have an asymptotic error expansion or that our time is in the asymptotic range.
+  \end{enumerate}
+  This is more serious than one might think.
+It is easy to verify that changing the force-field changes the final result, but how much this change is due to computational errors?
+If we cannot detect, when the computational error is irrelevant and estimate the discretization error, then we cannot compute the modelling error.
+
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=12cm]{charmm36tol04.pdf}
+%   \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
+The force-field is CHARMM36 and the constraint equations are not solved with great accuracy, i.e.
+${\tt tol} = 10^{-4}$.
+} \label{fig:charmm36tol04}
+% \end{figure}
+
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=12cm]{charmm36tol12.pdf}
+%    \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
+The force-field is CHARMM36 and the constraint equations are solved with great accuracy, i.e.
+${\tt tol} = 10^{-12}$.} \label{fig:charmm36tol12}
+% \end{figure}
 
@@
   \centering
-  \includegraphics[width=12cm]{charmm36tol04.pdf}
+  \includegraphics[width=12cm]{oplsaaltol04.pdf}
+  \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
+The force-field is OPLSAAL and the constraint equations are not solved with great accuracy, i.e.
+${\tt tol} = 10^{-4}$.
+} \label{fig:oplsaaltol04}
 \end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=12cm]{oplsaaltol12.pdf}
+   \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
+The force-field is OPLSAAL and the constraint equations are solved with great accuracy, i.e.
+${\tt tol} = 10^{-12}$.} \label{fig:oplsaaltol12}
+\end{figure}
+
+\section{The difference between succes and failure}
+
+In this section we contemplate the difference between the succesful and unsuccesful application of the theory.
+We identify two key conditions that are not always satisfied by the GROMACS library and we demonstrate that they are necessary for the succesful application of Richardson's techniques.
 
@@
 
-There are reasons beyond error estimation for solving the constraint equations accurately.
-The chemistry literature contains several papers that bemoan the fact that the energy drift is unacceptably large unless the constraint equations are solved with sufficient accuracy.
-However, what is perhaps less well known is \emph{direction} of the energy drift can depend on the choice of algorithm used to solve the constraint equations!
-We shall now present two examples that illustrate this point.
-We shall simulate the motion of a simple pendulum over a large number of periods.
-The only difference between the two simulation shall be the choice of algorithm used solve the scalar constraint equation.
-
-\paragraph{Example: A pendulum that gains energy through numerical error}
-
-\paragraph{Example: A pendulum that loses energy through numerical error}
-
-\section{The need for sufficient smoothness}
+When
 

### 2024-05-07 report/scripts/gatherfig.sh

@@ -1,3 +1,10 @@
 find experiments -name \*.png -exec cp {} fig \;
 find experiments -name \*.eps -exec cp {} fig \;
 
+# Crop white space from MATLAB eps files
+cd fig
+ps2pdf -DEPSCrop charmm36tol04.eps charmm36tol04.pdf
+ps2pdf -DEPSCrop charmm36tol12.eps charmm36tol12.pdf
+ps2pdf -DEPSCrop oplsaaltol04.eps oplsaaltol04.pdf
+ps2pdf -DEPSCrop oplsaaltol12.eps oplsaaltol12.pdf
+cd ..

### 2024-05-08 report/src/draft.tex
Significant progress on the draft

@@
 \end{align}
-In reality, the constraint equation is a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
-Let $T$ denote any value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
-Superficially, our task is to estimate $T - A_h$, but the presence of truncation and rounding errors ensure that we cannot hope to compute the exact value of $A_h$. 
-Let $\hat{A}_h(\tau,u)$ denote the value obtained from running the SHAKE algorithm using floating point arithmetic with unit roundoff $u$ and solving the constraint equations with a relative error that is bounded by $\tau$.
-This is the value returned by a computer. 
-We have the following trivial identity
+The constraint equation is a usually a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
+Let $T$ denote any real target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
+It is clear that $T = T(f)$ is a function of the forcefield $f$ and if $f_1 \not = f_2$, then it is relevant to compare $T(f_1)$ to $T(f_2)$.
+Can we estimate $T(f_1) - T(f_2)$?
+Can we even assert that $T(f_1) \not T(f_2)$?
+The fundamental problem is that we cannot compute the exact value $T(f)$ and we compute the exact value of $A_h(f)$.
+We must contend the fact that we cannot solve the constraint equations exactly and we cannot avoid rounding errors.
+Let $\hat{A}_h(f, \tau, u)$ denoted the value returned by the computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$.
+Suppose that the two values $\hat{A}_1 = \hat{A}_h(f_1, \tau, u)$ and $\hat{A}_2 = \hat{A}_h(f_2, \tau, u)$ are different.
+Does it follow that $T_1 = T(f_1)$ and $T_2 = T(f_2)$ are different?
+The answer is no, unless we are certain that
 \begin{equation}
-  T - \hat{A}_h(\tau,u) = T - A_h + A_h - \hat{A}_h(\tau,u)
+  | T_1 - \hat{A}_1| + |T_2 - \hat{A}_2| < |\hat{A}_1 - \hat{A}_2|
 \end{equation}
-We see that the error $T-\hat{A}_h(\tau,u)$ can be written as sum of two terms.
-The discretization error $E_h = T - A_h$ and the computational error $A_h - \hat{A}_h(\tau,u)$.
+in which case the triangle inequality implies that $T_1 \not = T_2$. 
+It is hard to estimate the error $T - \hat{A}_h(f, \tau, u)$ directly, but we can write
+\begin{equation}
+  T - \hat{A}_h)(f, \tau, u) = T - A_h + A_h - \hat{A}_h(f, \tau, u)
+\end{equation}
+and attempt to minimize the computational error $A_h - \hat{A}_h(f, \tau, u)$ and estimate the discretization error $T - A_h$.
 
-We can reduce the computational error by running the constraint solver until stagnation and by reducing the unit roundoff.
-It is frequently possible to use Richardson extrapolation to determine when computational errors are insignificant and estimate the discretization error.
-In the past, the authors have sought to accelerate the solution of nonlinear constraint equations in the context of molecular dynamics by integrating Newton's method into the GROMACS library for constrained molecular dynamics.
+As we shall demonstrate, Richardson extrapolation \cite{richardson1927} can sometimes be used to determine when computational errors are insignificant and then estimate the discretization error.
+In the past, we have sought to accelerate the solution of nonlinear constraint equations in the context of molecular dynamics by integrating Newton's method into the GROMACS library. 
 We have now discovered that the output of GROMACS is not always amenable to Richardson's analysis.
@@
 We identify two conditions that are not always satified by GROMACS and we demonstrate that each condition is necessary for the succesful application of Richardson's techniques.
-Our data and software is freely available from our GitHub along with every script needed to generate every number, table and figure from scratch.
-The names of MATLAB function shall be written using a typewriter font, ex.
-{\tt plot\_shells}.
+Our data and software is freely available from our GitHub along with every script and function needed to regenerate every number, table and figure from scratch.
+We encourage the reader to clone our repository, update their MATLAB path and execute all MATLAB functions that are explicitly named in this manuscript, such as {\tt plot\_shells}.
 
@@
 
-Consider the problem of approximation a target value $T$ using a numerical method $A = A_h$ that depends on a single real parameter $h$.
+Consider the problem of approximating a target value $T$ using a numerical method $A = A_h$ that depends on a single real parameter $h$.
 We shall assume there exists nonzero real constants $\alpha$ and $\beta$ and exponents
@@
 
-The theory applies to the difference $E_h = T - A_h$between the target value $T$ and the exact value of the approximation $A_h$.
-In practice, rounding errors ensure that the computed value $\hat{A}_h$ is different from the exact error $A_h$.
-However, it is frequently possible to assert that $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the computed error $\hat{E}_h = T - \hat{A}_h$.
+The theory applies to the difference $T - A_h$ between the target value $T$ and the exact value of the approximation $A_h$.
+In practice, the computed value $\hat{A}_h$ is different from the exact error $A_h$.
+However, it is frequently possible to assert that $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the error $T - \hat{A}_h$.
 We shall now demonstrate the procedure.
-To this end, we consider the problem of computing definite integrals
+To this end, we consider the familiar problem of computing definite integrals
 \begin{equation} \label{equ:integral}
@@
 Then $T = \frac{2}{3}$. Since $f$ is not differentible at $x=0$ we have no guarantee that there exists an asymptotic error expansion.
-The script {\tt rint\_mwe2} computes $A_h$ using $h_k = 2^{-k}$ for $k \in {0,1,\dots,25}$ and generates Figures \ref{fig:rint_mwe2a} and \ref{fig:rint_mwe2b}.
-From the raw data it is clear $p = 2$ cannot be true, but it is plausible that $p = \frac{3}{2}$.
+The script {\tt rint\_mwe2} computes $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,25\}$ and generates Figures \ref{fig:rint_mwe2a} and \ref{fig:rint_mwe2b}.
+From the raw data it is clear that $p = 2$ cannot be true, but it is plausible that $p = \frac{3}{2}$.
 Figure \ref{fig:rint_mwe1a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
 We observe that $k \rightarrow \log_2|\hat{F}_{h_k} - 2^{3/2}|$ is essentially a linear function of $k$ with slope $-\frac{1}{2}$ for $k \in \{2,3,\dots,18\}$.
-This is the so-called asymptotic range where the computed numbers $\hat{A}_h$ behave in a manner that is indistinguishable from the exact value $A_h$.
+This is the asymptotic range where the computed numbers $\hat{A}_h$ behave in a manner that is similar to the exact value $A_h$. 
 We conclude that the experiment is consistent with an asymptotic error expansion with $(p,q)=(\tfrac{3}{2},2)$.
@@
 
+% This table of values is generated by the function rint_mwe2.
+% It is omitted due to the page limitation.
+% The table is left in .tex as an example of how to use center verbatim tables.
 % \begin{myverbbox}[\scriptsize]{\vbeta}
@@
 We mention in passing that low order methods are more practical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
-This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small and this issue is more acute for high order methods than for low order methods.
+This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small and this issue is more acute for high order methods than for low order methods because $A_h$ tends to $T$ more rapidly for high order methods than for low order methods.
 
@@
 We have access to tables that describe the drag coefficient of 6 different shell types.
-Given the mass, diameter and muzzle velocity of the shells, can we determine the best possible match with any degree of confidence?
+Given all remaining parametes, such as the  mass, diameter and muzzle velocity of the shells, can we determine the drag coefficient function that provides the best possible match with any degree of confidence?
 
 The script {\tt maxrange\_rk1} models a shell as a point particle moving in a plane subject to Earth's standard gravity and the international standard atmosperic model.
-This is a crude model that is suitable only for short range artillery, because it does not account for the rotation of the Earth or the oriention and spin of the projectile.
+This is a crude model that is suitable only for short range artillery, as it does not account for the rotation of the Earth or the oriention and spin of the projectile.
 Each trajectory is integrated using Euler's explicit method ({\tt 'rk1'}) and all but the final step has the same size $h$.
 The final step is adjusted to place the shell directly on the ground.
-While the relevant equation is linear and could be solved to machine precision, the bisection method is deliberately used to compute the final step with an error that is bounded by ${\tt tol} \cdot h$, where ${\tt tol}$ is set by the script.
 The drag function are interpolated from tables using cubic spline interpolation.
@@
 
-For each drag coefficient our target value $T$ is the maximum range of the shell as the elevation varies continuously from $0$ to $\frac{\pi}{2}$.
+For each drag coefficient our target value $T$ is the maximum range of the shell as the elevation of the howitzer varies continuously from $0$ to $\frac{\pi}{2}$.
 For each drag coefficient, we compute $12$ different approximations $A_{h_k}$ of $T$ using the step size $h_k = 2^{3-k}$ seconds, where $k \in \{1,2,\dots,12\}$.
@@
 In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}.
-These figures represent calculations where the shell is placed on the ground with ``excessive'' accuracy in the sense that the final step is computed with an error that is bounded $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
-Figure \ref{fig:maxrange_rk1_table_tol53} list the maximum range and the corresponding error estimate for each of the shells know to us computed using a time step of $h = 2^{-9} s$.
+These two figures represent calculations where the shell is placed on the ground with excessive accuracy in the sense that the final step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
+Figure \ref{fig:maxrange_rk1_table_tol53} list the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
 In each case the error estimate suggests that the computed range is exact to number of figures shown.
-In particular, we see that while a G7 type shell achieves a maximum range of 17.4 km, all other shells have ranges that are less than 16.9 km.
-However, it is a fallacy to conclude anything on the basis of the computed ranges and the associated error estimates.
+In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km.
+However, it is a fallacy to conclude anything on the basis of this table.
 In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable.
@@
 We observe that for each drag coefficient, $k=12$ is still inside the asymtotic range and we have no reason to doubt the magnitude of the error estimate.
-We conclude that the best model for the D-20 howitzer is the G7 shell.
+We conclude that the best model for the D-20 howitzer is in fact the G7 shell.
 
@@
 
-When
+The output of GROMACS is clearly dependent how accurate we solve the constraint equations.
+When computing the range of a howitzer it necessary to adjust the final time-step to place the shell directly on the ground.
+In the case of Euler's explicit method, the relevant equation is linear, but in general this is nonlinear equation.
+The function {\tt maxrange\_rk1\_mwe1} and it is companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h \text{tol}$ using a wide range of tolerance {\tt tol}.
+It is instructive to observe the consequences of solving this equation inaccurately.
+Figures \ref{fig:maxrange_tol11} ($\text{tol} = 2^{-11}$) and Figures \ref{fig:maxrange_tol25} ($\text{tol} = 2^{-25}$) show the evolution of Richardson's fraction for two different values of the tolerance {\tt tol}.
+When the tolerance is large, there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
+When the tolerance is small, there is evidence to suggest that asymptotic error expansion exists and an asymptotic range can easily be identified for shells of type G1, G5 and G8.
+If we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is solve all central equations as accurately as the hardware will allow.
+
+\section{The need for sufficient smoothness}
+
+The GROMACS library is highly optimized in order to reduced the time-to-solution.
+In the case of the force-field calculation it is common to ignore the interaction between particles that are judged to be sufficiently far away.
+In practice, this is done by setting individual force-fields to zero outside of a sufficiently large ball.
+There is more than way to achieve this goal and the online documentation suggest that GROMACS operates with force-fields that are of class $C^1$.
+Why is this a concern in the context of practical error estimation?
+Proving that an asymptotic error expansion exist is often an exercise in manipulating Taylor expansions and this is only possible when the functions are sufficiently differentiable.
+We have designed a sequence of numerical experiment that illustrate this exact point.
+We consider a system of $m$ identical ions moving in a liquid.
+They repel each other according to Coulumb's law and they are pulled towards the origo according to Hooke's law.
+The friction between each ion and the fluid is proportional to its current velocity.
+This is a simple setup that ensure that particles will ultimate come to rest in a position where all forces balances each other.
+Can we estimate the value of the electrostatic energy at at specific time $T$?
+This is the question that we explore in a sequence of 4 scripts.
+In {\tt iontrap_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$.
+In {\tt opmtrap_mwe2} we modify the force-field $f = f(r)$ generated by an ion, by retain the exact value of the force-field inside a ball centered on the ion with radius $\rho/2$ and settting it to zero outside the ball.
+This ensures introduces jump-discontinuites into the simulation.
+In {\tt iontrap_mwe4} the transition to zero is done smoothly in the region from $0.5 \rho < r < 0.95 r$.
+It is clear that changing the force-fields is equivalent to physical changes, but can we estimate the discretization error?
+Figure \ref{}, \ref{}, \ref{} show the evolution of Richardson's fraction for the total kinetic energy at the end of
 

### 2024-05-08 report/src/draft.tex
Fixed misprints preventing compilation

@@
 This is the question that we explore in a sequence of 4 scripts.
-In {\tt iontrap_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$.
-In {\tt opmtrap_mwe2} we modify the force-field $f = f(r)$ generated by an ion, by retain the exact value of the force-field inside a ball centered on the ion with radius $\rho/2$ and settting it to zero outside the ball.
+In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$.
+In {\tt ointrap\_mwe2} we modify the force-field $f = f(r)$ generated by an ion, by retain the exact value of the force-field inside a ball centered on the ion with radius $\rho/2$ and settting it to zero outside the ball.
 This ensures introduces jump-discontinuites into the simulation.
-In {\tt iontrap_mwe4} the transition to zero is done smoothly in the region from $0.5 \rho < r < 0.95 r$.
+In {\tt iontrap\_mwe4} the transition to zero is done smoothly in the region from $0.5 \rho < r < 0.95 r$.
 It is clear that changing the force-fields is equivalent to physical changes, but can we estimate the discretization error?

### 2024-05-09 report/src/draft.tex
Good progress on the paper. Details need to be added and some references, but the core text is there.

@@
 
-\begin{figure}
+\begin{figure}[t!]
   \centering
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=12cm]{oplsaaltol04.pdf}
+  \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
+The force-field is OPLSAAL and the constraint equations are not solved with great accuracy, i.e.
+${\tt tol} = 10^{-4}$.
+} \label{fig:oplsaaltol04}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=12cm]{oplsaaltol12.pdf}
+   \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
+The force-field is OPLSAAL and the constraint equations are solved with great accuracy, i.e.
+${\tt tol} = 10^{-12}$.} \label{fig:oplsaaltol12}
+\end{figure}
+
 \begin{enumerate}
-\item The kinitic energy exhibits violent oscilations for ${\tt tol} = 10^{-4}$.
+\item The kinetic energy exhibits violent oscilations for ${\tt tol} = 10^{-4}$.
 Rapid oscillations are also shown by the potential energy, but they are less extreme.
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=12cm]{oplsaaltol04.pdf}
-  \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
-The force-field is OPLSAAL and the constraint equations are not solved with great accuracy, i.e.
-${\tt tol} = 10^{-4}$.
-} \label{fig:oplsaaltol04}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=12cm]{oplsaaltol12.pdf}
-   \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
-The force-field is OPLSAAL and the constraint equations are solved with great accuracy, i.e.
-${\tt tol} = 10^{-12}$.} \label{fig:oplsaaltol12}
-\end{figure}
-
 \section{The difference between succes and failure}
@@
 This is a simple setup that ensure that particles will ultimate come to rest in a position where all forces balances each other.
-Can we estimate the value of the electrostatic energy at at specific time $T$?
-This is the question that we explore in a sequence of 4 scripts.
-In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$.
-In {\tt ointrap\_mwe2} we modify the force-field $f = f(r)$ generated by an ion, by retain the exact value of the force-field inside a ball centered on the ion with radius $\rho/2$ and settting it to zero outside the ball.
-This ensures introduces jump-discontinuites into the simulation.
+Can we estimate the value of the kinetic energy at at specific time $T$?
+This is the question that we explore in a sequence of scripts.
+Let $f$ denote the force-field emenating from an particle with charge $q$, i.e
+\begin{equation}
+  f(\bm{r}) = - c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
+\end{equation}
+In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$. In {\tt iontrap\_mwe2} we replace $f$ with
+\begin{equation}
+  f_1(\bm{r}) = f(\bm{r}) g_1(r), \quad g_1(r) = \begin{cases} 1  & 0 \leq r < \rho \\ 0 & \rho \leq r \end{cases}
+\end{equation}
+This introduces jump discontinuites into the simulation.
 In {\tt iontrap\_mwe4} the transition to zero is done smoothly in the region from $0.5 \rho < r < 0.95 r$.
 It is clear that changing the force-fields is equivalent to physical changes, but can we estimate the discretization error?
-Figure \ref{}, \ref{}, \ref{} show the evolution of Richardson's fraction for the total kinetic energy at the end of
+Figure \ref{fig:iontrap_mwe1}, \ref{fig:iontrap_mwe2}, \ref{fig:iontrap_mwe4} show the evolution of Richardson's fraction for the total kinetic energy at the end of each simulation.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=10cm]{iontrap_mwe1.pdf}
+  \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$}
+  \label{fig:iontrap_mwe1}
+\end{figure}
+
+When the force-fields are not perturbed, Figure~\ref{fig:iontrap_mwe1}, there experiments supports the conjecture that the exists an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate the dynamical system.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=10cm]{iontrap_mwe2.pdf}
+  \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$.
+The electrostatic force-fields are zero outside a small sphere and exhibit jump discontinuities.}
+  \label{fig:iontrap_mwe2}
+\end{figure}
+
+When the force-fields are truncated and jump-discontinuities are introduced, Figure~\ref{fig:iontrap_mwe2}, there is no evidence to suggest that there exists an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=10cm]{iontrap_mwe4.pdf}
+  \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$.
+The electrostatic force-fields are zero outside a small sphere, but the transition to zero is done smoothly.}
+  \label{fig:iontrap_mwe4}
+\end{figure}
+
+When the force-field vanish outside a small ball, but retain their smoothness, Figure~\ref{fig:iontrap_mwe4}, we retain the ability to judge when Richardson's error estimates are reliable.
 

### 2024-05-09 report/experiments/iontrap/src/iontrap_mwe1.m

@@ -1,15 +1,26 @@
 % Ion trap with dampning
 % Runge-Kutta methods
-% No cutoff function
+% The force-fields are smooth and have infinite range
+% 
 
-% Set the central file name
-datpath='../experiments/iontrap/dat/';
-fname=strcat(datpath,"iontrap_mwe1.mat");
+% Clear the work space
+clear;
+
+% Set the main path
+mpath='../experiments/iontrap/';
+
+% Set the data path
+dpath=strcat(mpath,'dat/');
+dname=strcat(dpath,'iontrap_mwe1.mat');
+
+% Set the figure path
+fpath=strcat(mpath,'fig/');
+fname=strcat(fpath,'iontrap_mwe1.eps');
 
 % Check if raw data exists
-if ~isfile(fname)
+if ~isfile(dname)
 
-    % Run experiment from scracth
+    % Run experiment from scratch
     fprintf("Generating raw data\n");
 
     % Seed the generator
@@ -24,11 +35,14 @@ if ~isfile(fname)
     % Set the leading dimension of the arrays
     ld=n*dim;
 
-    % Generate position and momenta
+    % Generate position and velocities
     q0=rand(ld,1)-rand(ld,1); p0=zeros(ld,1);
 
     % Set the charges
     charge=ones(n,1)/16;
+    
+    % Set the mass of the particles
+    mass=ones(n,1);
 
     % Generate initial state
     x0=[q0; p0];
@@ -42,6 +56,9 @@ if ~isfile(fname)
     % Scaling factor for columb potential
     scale=1;
 
+    % Dummy declarations. There is not cut-off function in this simulation
+    rho=[]; rho1=[]; rho2=[]; cut=[];
+    
     % Set the forcefield: coulomb force + springs + dampning
     f=@(q,p)coulomb_force(dim,n,charge,q,scale)-spring*q-damp*p;
 
@@ -52,7 +69,7 @@ if ~isfile(fname)
     g=@(t,x)[x(ld+1:2*ld,1); f(x(1:ld,1),x(ld+1:2*ld,1))];
 
     % The length of the simulation
-    T=10;
+    T=5;
 
     % Number of methods
     nm=4;
@@ -70,7 +87,7 @@ if ~isfile(fname)
     % Number of samples
     N1=512;
 
-    % Number of simulations/approximations
+    % Number of approximations A_h
     kmax=12;
 
     % Allocate space for the raw data
@@ -79,6 +96,9 @@ if ~isfile(fname)
     % Allocate space for the processed data
[…]

### 2024-05-09 report/experiments/iontrap/src/iontrap_mwe2.m

@@ -1,16 +1,26 @@
 % Ion trap with dampning
 % Runge-Kutta methods
-% The forcefield has a jump discontinuity.
-% It is truncated at a finite distance.
+% The electrostatic forcefields have a jump discontinuity.
+% They are truncated at a finite distance.
 
-% Set the central file name
-datpath='../experiments/iontrap/dat/';
-fname=strcat(datpath,"iontrap_mwe2.mat");
+% Clear the work space
+clear;
+
+% Set the main path
+mpath='../experiments/iontrap/';
+
+% Set the data path
+dpath=strcat(mpath,'dat/');
+dname=strcat(dpath,'iontrap_mwe2.mat');
+
+% Set the figure path
+fpath=strcat(mpath,'fig/');
+fname=strcat(fpath,'iontrap_mwe2.eps');
 
 % Check if raw data exists
-if ~isfile(fname)
+if ~isfile(dname)
 
-    % Run experiment from scracth
+    % Run experiment from scratch
     fprintf("Generating raw data\n");
 
     % Seed the generator
@@ -25,12 +35,15 @@ if ~isfile(fname)
     % Set the leading dimension of the arrays
     ld=n*dim;
 
-    % Generate position and momenta
+    % Generate position and velocities
     q0=rand(ld,1)-rand(ld,1); p0=zeros(ld,1);
 
     % Set the charges
     charge=ones(n,1)/16;
 
+    % Set the mass of the particles
+    mass=ones(n,1);
+    
     % Generate initial state
     x0=[q0; p0];
 
@@ -44,7 +57,7 @@ if ~isfile(fname)
     scale=1;
 
     % Set cutoff function
-    rho=0.1462/2; cut=@(x)cut1(rho,x);
+    rho=0.1462/2; rho1=[]; rho2=[]; cut=@(x)cut1(rho,x);
     
     % Set the forcefield: coulomb force + springs + dampning
     f=@(q,p)modified_coulomb_force(dim,n,charge,q,scale,cut)-spring*q-damp*p;
@@ -56,7 +69,7 @@ if ~isfile(fname)
     g=@(t,x)[x(ld+1:2*ld,1); f(x(1:ld,1),x(ld+1:2*ld,1))];
 
     % The length of the simulation
-    T=10;
+    T=5;
 
     % Number of methods
     nm=4;
@@ -74,7 +87,7 @@ if ~isfile(fname)
     % Number of samples
     N1=512;
 
-    % Number of simulations/approximations
+    % Number of approximations A_h
     kmax=12;
 
[…]

### 2024-05-09 report/experiments/iontrap/src/iontrap_mwe4.m

@@ -1,15 +1,26 @@
 % Ion trap with dampning
 % Runge-Kutta methods
 % The forcefield transitions smoothly to zero 
+% 
 
-% Set the central file name
-datpath='../experiments/iontrap/dat/';
-fname=strcat(datpath,"iontrap_mwe4.mat");
+% Clear the work space
+clear;
+
+% Set the main path
+mpath='../experiments/iontrap/';
+
+% Set the data path
+dpath=strcat(mpath,'dat/');
+dname=strcat(dpath,'iontrap_mwe4.mat');
+
+% Set the figure path
+fpath=strcat(mpath,'fig/');
+fname=strcat(fpath,'iontrap_mwe4.eps');
 
 % Check if raw data exists
-if ~isfile(fname)
+if ~isfile(dname)
 
-    % Run experiment from scracth
+    % Run experiment from scratch
     fprintf("Generating raw data\n");
 
     % Seed the generator
@@ -24,12 +35,15 @@ if ~isfile(fname)
     % Set the leading dimension of the arrays
     ld=n*dim;
 
-    % Generate position and momenta
+    % Generate position and velocities
     q0=rand(ld,1)-rand(ld,1); p0=zeros(ld,1);
 
     % Set the charges
     charge=ones(n,1)/16;
 
+    % Set the mass of the particles
+    mass=ones(n,1);
+    
     % Generate initial state
     x0=[q0; p0];
 
@@ -55,7 +69,7 @@ if ~isfile(fname)
     g=@(t,x)[x(ld+1:2*ld,1); f(x(1:ld,1),x(ld+1:2*ld,1))];
 
     % The length of the simulation
-    T=10;
+    T=5;
 
     % Number of methods
     nm=4;
@@ -73,7 +87,7 @@ if ~isfile(fname)
     % Number of samples
     N1=512;
 
-    % Number of simulations/approximations
+    % Number of approximations A_h
     kmax=12;
 
     % Allocate space for the raw data
@@ -82,6 +96,9 @@ if ~isfile(fname)
     % Allocate space for the processed data
     table=zeros(kmax,4,nm);
 
+    % Compute the number of experiments
+    nexp=nm*kmax;
+    
     % Loop over the different methods
     for j=1:nm
         % Select the method
@@ -91,8 +108,9 @@ if ~isfile(fname)
         % Loop over the different timestep sizes
         for k=1:kmax
             % Progress
-            cnt=k+(j-1)*kmax; disp([cnt nm*kmax]);
[…]

### 2024-05-09 report/scripts/gatherfig.sh

@@ -7,4 +7,7 @@ ps2pdf -DEPSCrop charmm36tol04.eps charmm36tol04.pdf
 ps2pdf -DEPSCrop charmm36tol12.eps charmm36tol12.pdf
 ps2pdf -DEPSCrop oplsaaltol04.eps oplsaaltol04.pdf
 ps2pdf -DEPSCrop oplsaaltol12.eps oplsaaltol12.pdf
+ps2pdf -DEPSCrop iontrap_mwe1.eps iontrap_mwe1.pdf
+ps2pdf -DEPSCrop iontrap_mwe2.eps iontrap_mwe2.pdf
+ps2pdf -DEPSCrop iontrap_mwe4.eps iontrap_mwe4.pdf
 cd ..

### 2024-05-11 report/src/draft.tex
Good progress. References added. The page limit is being exceeded, but there many options left.

@@
 %
-\usepackage{amsmath, amssymb, lineno, lscape,hyperref,subcaption, verbatimbox, bm}
+\usepackage{amsmath, amssymb, lineno, lscape,hyperref,subcaption, verbatimbox, bm, enumitem}
 \usepackage[T1]{fontenc}
@@
 \begin{abstract} We consider the problem of estimating the error when solving a system of differential algebraic equations.
-Richardson extrapolation is a standard technique in numerical analysis that can be used to judge when rounding errors are irrelevant and estimate the discretization error.
-We have simulated molecular dynamics with constraints using GROMACS-library and found that the output is not always ameable to Richardson's analysis.
+Richardson extrapolation is a standard technique that can be used to judge when rounding errors are irrelevant and estimate the discretization error.
+We have simulated molecular dynamics with constraints using the GROMACS library and found that the output is not always ameable to Richardson's analysis.
 We derive and illustrate Richardson's technique using a variety of numerical experiments.
-We identify two necessary conditions that are not always satisfied by GROMACS.
+We identify two necessary conditions that are not always satisfied by the GROMACS library.
 \keywords{}
@@
 
-\section{Motivation} Consider the problem of solving the following system of differential algebraic equations
+\section{Motivation} Consider the problem resolving the motion of a system of atoms moving in a force-field $f$ subject to a set of constraints $g$.
+In this case, Newton's 2nd law takes the form the following system of differential algebraic equations
 \begin{align}
@@
 \end{align}
-This is Newton's law for a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints specified by the function $\bm{g}$.
 The vector $\bm{q}$ represents the position of the atoms.
-The vector $\bm{v}$ represents the velocities of the atoms.
+The vector $\bm{v}$ represents the velocities of the atoms. The function $\bm{f}$ represents the force-field acting on the atoms.
 The nonsingular diagonal matrix $\bm{M}$ lists the masses of the atoms.
-The function $\bm{G}$ is the Jacobian of the function $\bm{g}$.
+The function $\bm{G}$ is the Jacobian of the constraint function $\bm{g}$.
 In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm.
@@
 The constraint equation is a usually a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
-Let $T$ denote any real target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
-It is clear that $T = T(f)$ is a function of the forcefield $f$ and if $f_1 \not = f_2$, then it is relevant to compare $T(f_1)$ to $T(f_2)$.
-Can we estimate $T(f_1) - T(f_2)$?
-Can we even assert that $T(f_1) \not T(f_2)$?
-The fundamental problem is that we cannot compute the exact value $T(f)$ and we compute the exact value of $A_h(f)$.
-We must contend the fact that we cannot solve the constraint equations exactly and we cannot avoid rounding errors.
-Let $\hat{A}_h(f, \tau, u)$ denoted the value returned by the computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$.
-Suppose that the two values $\hat{A}_1 = \hat{A}_h(f_1, \tau, u)$ and $\hat{A}_2 = \hat{A}_h(f_2, \tau, u)$ are different.
-Does it follow that $T_1 = T(f_1)$ and $T_2 = T(f_2)$ are different?
-The answer is no, unless we are certain that
+Now let $T \in \mathbb{R}$ denote any target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
+It is clear that $T$ and $A_h$ are both functions of the force-field $\bm{f}$ and the question of adjusting $\bm{f}$ to match the outcome of a physical experiment naturally suggests itself.
+Let therefore $T_0 \in \mathbb{R}$ be given and consider the problem of solving the equation
 \begin{equation}
-  | T_1 - \hat{A}_1| + |T_2 - \hat{A}_2| < |\hat{A}_1 - \hat{A}_2|
+  T_0 = T(\bm{f})
 \end{equation}
-in which case the triangle inequality implies that $T_1 \not = T_2$. 
-It is hard to estimate the error $T - \hat{A}_h(f, \tau, u)$ directly, but we can write
+with respect to $\bm{f}$.
+The fundamental problem is that we cannot compute the exact values of $T(\bm{f})$ and $A_h(\bm{f})$.
+We must contend with the fact that we cannot solve the constraint equations exactly nor can we avoid rounding errors.
+Let $\hat{A}_h(\bm{f}, \tau, u)$ denote the value returned by our computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$.
+Suppose that $T_0 - A_h(\bm{f}, \tau, u)$ is small.
+Can we show that $T_0 - T(\bm{f})$ is small?
+The triangle inequality implies the following inequality
 \begin{equation}
-  T - \hat{A}_h)(f, \tau, u) = T - A_h + A_h - \hat{A}_h(f, \tau, u)
+  |T_0 - T(\bm{f}) | \leq  |T_0 - A_h(\bm{f},\tau,u)| + |A_h(\bm{f}) - A_h(\bm{f},\tau,u)| + |T(\bm{f}) - A_h(\bm{f})|.
 \end{equation}
-and attempt to minimize the computational error $A_h - \hat{A}_h(f, \tau, u)$ and estimate the discretization error $T - A_h$.
+We conclude that $T_0 - T(f)$ is small, if the computational error $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the discretization error $T(\bm{f}) - A_h(\bm{f})$ are both small.
+If we cannot assert control over these errors, then we cannot say with certainty that our model is a good approximation of the physical reality.
 
-As we shall demonstrate, Richardson extrapolation \cite{richardson1927} can sometimes be used to determine when computational errors are insignificant and then estimate the discretization error.
-In the past, we have sought to accelerate the solution of nonlinear constraint equations in the context of molecular dynamics by integrating Newton's method into the GROMACS library. 
-We have now discovered that the output of GROMACS is not always amenable to Richardson's analysis.
-In this paper we derive and illustrate the use of Richardson's technique using a variety of numerical experiments.
-We identify two conditions that are not always satified by GROMACS and we demonstrate that each condition is necessary for the succesful application of Richardson's techniques.
+As we shall demonstrate, Richardson extrapolation \cite{richardson1927} can often be used to determine when computational errors are insignificant and estimate the discretization error.
+In the past, we have sought to improve the solution of nonlinear constraint equations in the context of molecular dynamics by integrating Newton's method into the GROMACS library \cite{villelas2023accurate}
+Our goal has been to reduce the computational error at low cost in order to improve the long-term stability of constrained simulations of molecular dynamics.
+We have now found that the output of GROMACS is not always amenable to Richardson's methods.
+In this paper we shall derive and illustrate the use of Richardson's method using a variety of numerical experiments.
+We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson's method.
 Our data and software is freely available from our GitHub along with every script and function needed to regenerate every number, table and figure from scratch.
-We encourage the reader to clone our repository, update their MATLAB path and execute all MATLAB functions that are explicitly named in this manuscript, such as {\tt plot\_shells}.
+We encourage the reader to clone our repository and execute all MATLAB functions that are explicitly named in this manuscript, such as {\tt plot\_shells}.
 
@@
 Consider the problem of approximating a target value $T$ using a numerical method $A = A_h$ that depends on a single real parameter $h$.
-We shall assume there exists nonzero real constants $\alpha$ and $\beta$ and exponents
+We shall assume there exists nonzero real constants $\alpha$ and $\beta$ and real exponents
 \begin{equation}
@@
 \end{proof}
-We shall now demonstrate how to demonstrate the existence of an asymptotic error expansion experimentally.
+We shall now demonstrate how to investigate the existence of an asymptotic error expansion experimentally.
 We define Richardson's fraction $F_h$ using the expression
@@
   \begin{equation}
-    f(h) = 2^m \nu h^m + \phi(h), \quad g(h) = \nu h^m + \phi(h)
+    f(h) = 2^m \nu h^m + \phi(h), \quad g(h) = \nu h^m + \phi(2h)
   \end{equation}
@@
 \end{proof}
-We conclude that if the error $E_h$ satisfies equation \eqref{equ:aex}, then the order of the primary error term can be determined as the limit
+We conclude that if $T-A_h$ satisfies the asymptotic error expansion \eqref{equ:aex}, then the order of the primary error term can be determined as the limit
 \begin{equation}
@@
 \end{equation}
-is a good approximation for $h$ sufficiently small.
-In particular, we note that the left-hand side of equation \ref{equ:asymtotic-behavior-of-Fh} is essentially a linear function of $\log(h)$ with slope $m$.
+is a good approximation for $h$ sufficiently small. In particular, we note that the high-hand side of equation \ref{equ:asymtotic-behavior-of-Fh} is a linear function of $\log(h)$ with slope $m$.
 \section{Elementary examples}
@@
 The theory applies to the difference $T - A_h$ between the target value $T$ and the exact value of the approximation $A_h$.
-In practice, the computed value $\hat{A}_h$ is different from the exact error $A_h$.
-However, it is frequently possible to assert that $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the error $T - \hat{A}_h$.
-We shall now demonstrate the procedure.
+In practice, the computed value $\hat{A}_h$ is different from the exact value $A_h$.
+However, it is often possible to assert that $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the error $T - \hat{A}_h$.
+We shall now demonstrate this procedure.
 To this end, we consider the familiar problem of computing definite integrals
@@
 \end{equation}
-It is well-known that if the function $f : [a,b] \rightarrow \mathbb{R} $ is \emph{everywhere} smooth, then there exists a sequence $\{\alpha_j\}_{j=1}^\infty \subset \mathbb{R}$ such that 
-\begin{equation}
-  E_h = \sum_{j=1}^k \alpha _j h^{2j} + O(h^{2k+1}), \quad h \rightarrow 0_+
+It is well-known that if $f \in C^{\infty}([a,b], \mathbb{R})$, then there exists a sequence $\{\alpha_j\}_{j=1}^\infty \subset \mathbb{R}$ such that 
+\begin{equation} \label{equ:trapezoidal-rule:aex}
+  E_h = \sum_{j=1}^k \alpha _j h^{2j} + O(h^{2k+1}), \quad h \rightarrow 0_+.
 \end{equation}
+In particular, $(p,q,r) = (2,4,6)$ when $f$ is everywhere smooth.
+\paragraph{Integration of a function that is everywhere smooth.}
 
-\paragraph{Integration of a function that is everywhere smooth}
-
-Let $[a,b]= [0,1]$, let $f : [a, b] \rightarrow \mathbb{R}$ be given by $f(x) = e^x$ and $T$ be given by equation \eqref{equ:integral}.
+Let $f : [0, 1] \rightarrow \mathbb{R}$ be given by $f(x) = e^x$ and $T$ be given by equation \eqref{equ:integral}.
 The script {\tt rint\_mwe1} computes the composite trapezoidal sum $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,20\}$ and generates \ref{fig:rint_mwe1a} and \ref{fig:rint_mwe1b}.
-The raw data shows that $\hat{A}_{h_k}$ approach $4 = 2^2$ as $k$ increases and $k \in \{2,3,\dots,14\}$.
+The raw data shows that $\hat{A}_{h_k}$ approaches $4 = 2^2$ as $k$ increases and $k \in \{2,3,\dots,14\}$.
 This suggests that $p=2$.
 Figure \ref{fig:rint_mwe1a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
-We observe that $k \rightarrow \log_2|\hat{F}_{h_k} - 4|$ is essentially a linear function of $k$ with slope $-2$ for $k \in \{2,3,\dots,10\}$.
+We observe that $k \rightarrow \log_2|\hat{F}_{h_k} - 4|$ is essentially a linear function of $k$ with slope $-m = -2$ for $k \in \{2,3,\dots,10\}$.
 This is the so-called asymptotic range, where the computed value $\hat{A}_h$  behaves in a manner that is indistinguishable from the exact value $A_h$.
-We conclude that the experiments supports the existence of an asymptotic error expansion with $(p,q) = (2,4)$.
+We conclude that the experiments supports the existence of an asymptotic error expansion with $(p,q) = (p,p+m) = (2,4)$.
 Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T - \hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe1b}.
-We observe that computed value of Richardson's error estimate is a good approxmation of the error $T - \hat{A}_{h_h}$.
-In fact, the corresponding relative error decreases when $k$ increases and we remain inside the asymptotic region.
+We observe that computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_{h_h}$.
+In fact, the corresponding relative error decreases when $k$ increases as long as we also remain inside the asymptotic region.
 
@@
 \end{subfigure}%
-\caption{The evolution of $F_h$ and the accuracy of $R_h$ for a method with $(p,q) = (2,4)$.}
+\caption{The behavior of $F_h$, $E_h$ and $R_h$ for a method with $(p,q) = (2,4)$.}
 \end{figure}
 
-\paragraph{Integration of a function that is smooth in all but one point}
+\paragraph{Integration of a function that is smooth in all but one point.}
 
 Let $f : [0,1] \rightarrow \mathbb{R}$ be given by $f(x) = \sqrt{x}$ and let $T$ be given by equation \eqref{equ:integral}.
-Then $T = \frac{2}{3}$. Since $f$ is not differentible at $x=0$ we have no guarantee that there exists an asymptotic error expansion.
+Then $T = \frac{2}{3}$. Since $f$ is not differentible at $x=0$ we have no guarantee that there exists an asymptotic error expansion of the form given by equation \eqref{equ:trapezoidal-rule:aex}.
 The script {\tt rint\_mwe2} computes $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,25\}$ and generates Figures \ref{fig:rint_mwe2a} and \ref{fig:rint_mwe2b}.
-From the raw data it is clear that $p = 2$ cannot be true, but it is plausible that $p = \frac{3}{2}$.
+From the raw data it is clear that $p = 2$ cannot be true, but it is plausible that $p \approx \frac{3}{2}$.
 Figure \ref{fig:rint_mwe1a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
@@
 We mention in passing that low order methods are more practical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
-This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small and this issue is more acute for high order methods than for low order methods because $A_h$ tends to $T$ more rapidly for high order methods than for low order methods.
+% This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small.
+This issue is more acute for high order methods than for low order methods, because $A_h$ tends to $T$ more rapidly for high order methods than for low order methods.
 
@@
 
-In this section we present the results of more elaborate examples that highlight the practical limitations of Richardson's extrapolation as well as some limitations of contemporary software for doing molecular dynamics.
+In this section we present the results of more elaborate experiments that highlight the practical limitations of Richardson's extrapolation.
 
@@
 
-Consider the D-20 howitzer whose effective range is known to be about 17.4 km.
-We have access to tables that describe the drag coefficient of 6 different shell types.
-Given all remaining parametes, such as the  mass, diameter and muzzle velocity of the shells, can we determine the drag coefficient function that provides the best possible match with any degree of confidence?
+Consider the D-20 howitzer whose maximum range is known to be about 17.3 km \cite{foss1976artillery}.
+We have access to tables of the drag coefficient for 6 different shells types \cite{jbmballistics}.
+Can we determine the drag coefficient that provides the best match to the physical reality?
 
-The script {\tt maxrange\_rk1} models a shell as a point particle moving in a plane subject to Earth's standard gravity and the international standard atmosperic model.
-This is a crude model that is suitable only for short range artillery, as it does not account for the rotation of the Earth or the oriention and spin of the projectile.
+The script {\tt maxrange\_rk1} models a shell as a point particle moving in a plane subject to Earth's standard gravity and the international standard atmospheric model.
+This model is suitable only for short range artillery, as it does not account for the rotation of the Earth or the oriention and spin of the projectile.
 Each trajectory is integrated using Euler's explicit method ({\tt 'rk1'}) and all but the final step has the same size $h$.
-The final step is adjusted to place the shell directly on the ground.
+The final step is adjusted to place the shell on the ground.
 The drag function are interpolated from tables using cubic spline interpolation.
@@
 For each drag coefficient and for each value of the time step $h$, a range function is defined which returns the range of the shell as a function of the howitzers elevation $\theta$.
-The range functions are unimodal and the maximum range is found using the golden section algorithm.
+The range functions are unimodal and the maximum range is found using the golden section search algorithm.
 The initial search bracket is $[0,\pi/2$] and this bracket is systematically reduced in length until it is shorter than $\frac{\pi}{2}u$.
@@
 In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}.
-These two figures represent calculations where the shell is placed on the ground with excessive accuracy in the sense that the final step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
+These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
 Figure \ref{fig:maxrange_rk1_table_tol53} list the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
@@
 In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km.
-However, it is a fallacy to conclude anything on the basis of this table.
+It is however, a fallacy to conclude anything on the basis of this table alone.
 In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable.
@@
 
-\begin{myverbbox}[\normalsize]{\vgamma}
-Shell type  Maximum range (m) Error estimate (m)
-    G1           12832                0.4 
-    G2           16857                0.1 
-    G5           15918                0.2 
-    G6           15556                0.2 
-    G7           17461                0.1 
-    G8           15914                0.1 
-\end{myverbbox}
-
 \begin{figure}
   \centering
-  \vgamma
-  \caption{The computed maximum range for 6 different shell types fired from the D-20 howitzer.}
+  \begin{tabular}{c|c|c}
+    Shell type & Maximum range (m) & Error estimate (m) \\ \hline
+    G1      &     12832    &            0.4 \\
+    G2      &     16857    &            0.1 \\
+    G5      &     15918    &            0.2 \\
+    G6      &     15556    &            0.2 \\
+    G7      &     17461    &            0.1 \\
+    G8      &     15914    &            0.1 \\
+  \end{tabular}
+  \caption{The computed maximum range for 6 shells fired from a D-20 howitzer.}
   \label{fig:maxrange_rk1_table_tol53} 
@@
   \centering
-  \includegraphics[width=12cm]{maxrange_rk1_tol53.png}
+  \includegraphics[width=\linewidth]{maxrange_rk1_tol53.png}
   \caption{The evolution of Richardson's fraction corresponding to the maximum range of 6 different shells fired from the D-20 howitzer.} \label{fig:maxrange_rk1_fraction_tol53}
@@
 
-We utilized GROMACS v2021 to conduct experiments on the behavior of hen egg white lysozyme submerged in water within a cubic simulation box, following Justin Lemkul's Lysozyme in Water GROMACS Tutorial \cite{}.
+We utilized GROMACS v2021 to conduct experiments on the behavior of hen egg white lysozyme submerged in water within a cubic simulation box, following Justin Lemkul's Lysozyme in Water GROMACS Tutorial \cite{lemkul2019from}.
 Several steps were taken to prepare the system for production simulation: first, ions were introduced to achieve electrical neutrality.
@@
 The described process was replicated using two different force fields, OPLS-AA/L and CHARMM36.
-We conducted production simulations of 1 ps for both force fields, using $n$ steps to cover this interval where
-\begin{equation}
-  n \in \{250, 500, 1000, 1100:100:2000, 3000:1000:16000\}.
-\end{equation}
-Moreover, we used two different tolerances {\tt tol} for the SHAKE algorithm, namely
-\begin{equation}
-  \text{tol} \in \{10^{-4}, 10^{-12}\}.
-\end{equation}
+We conducted production simulations of 1 ps for both force fields, using $n \in \{250, 500, 1000, 1100:100:2000, 3000:1000:16000 \}$ steps to cover this interval.
+Moreover, we used two different values of the tolerance $\tau$ for the SHAKE algorithm, namely $\tau \in \{10^{-4}, 10^{-12}\}$. 
 For each experiment, we computed the total kinetic and potential energy of the system at the end of the simulation.
@@
 
+Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e, $\tau = 10^{-4}$.
+The amplitude of the oscillations is reduced when $\tau = 10^{-12}$.
+We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expection the computed approximation unless $\tau$ is very small.
+
 \begin{figure}
   \centering
-  \includegraphics[width=12cm]{oplsaaltol04.pdf}
+  \includegraphics[width=\linewidth]{oplsaaltol04.pdf}
   \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
@@
 
+Secondly, the total energy exhibits grows linearly with the number of time steps. This is not surprising as we expect the rounding error to grown with the number of operations.
+
+%The function {\tt gromacs\_fit} does a linear least squares fit of the total energy {\tt etol} to two models $M_1$ and $M_2$ where model $M_p$ assumes that ${\tt etol} = T + C_1n + C_2/n^p$ for suitable constants $T, C_1, C_2$.
+Since the SHAKE algorithm is 2nd order accurate in the time step, one would expect model $M_2$ to provide the best fit, but for this data model $M_1$ is a marginally better.
+While mildly interesting, this issue is vastly less important than the final point.
+
 \begin{figure}
   \centering
-  \includegraphics[width=12cm]{oplsaaltol12.pdf}
+  \includegraphics[width=\linewidth]{oplsaaltol12.pdf}
    \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
@@
 
-\begin{enumerate}
-\item The kinetic energy exhibits violent oscilations for ${\tt tol} = 10^{-4}$.
-Rapid oscillations are also shown by the potential energy, but they are less extreme.
-The applitude of all oscillations is significantly reduced when the constraint equations are solved accurately, i.e., ${\tt tol} = 10^{-12}$.
-This reduction is not unexpected.
-\item Apart from the oscillations, the total energy exhibits what appears to be linear growth with the number of time steps.
-This is not surprising as we expect the rounding error to grown with the number of operations.
-The function {\tt gromacs\_fit} does a linear least squares fit of the total energy {\tt etol} to two models $M_1$ and $M_2$ where model $M_p$ assumes that ${\tt etol} = T + C_1n + C_2/n^p$ for suitable constants $T, C_1, C_2$.
-Since the SHAKE algorithm is 2nd order accurate in the time step, one would expect model $M_2$ to provide the best fit, but for this data model $M_1$ is a marginally better.
-While mildly interesting, this issue is vastly less important than the final point.
-\item If the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} in the asymptotic range.
+Thirdly, if the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} in the asymptotic range.
 Why is this?
-If we where in the asymptotic range, then the computed energies $A_h$ would follows follow a power law, i.e.
-$A_h \approx T - \alpha h^p$ would be a good approximation for some power $p>0$.
+If we where in the asymptotic range, then the computed energies $\hat{A}_h$ would follow a power-law and $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some power $p>0$.
 In particular, the behaviour would be \emph{monotone} and the tiny oscillations that we have recorded would not be present.
-\end{enumerate}
-
-We are left to make the following conclusions:
-\begin{enumerate}
-  \item If we do not solve the constraint equations accurately, then we cannot expect the computed energies to change slowly with the number of time steps.
-  \item If we solve the constraint equations accurately, then there is no reason to believe that we have an asymptotic error expansion or that our time is in the asymptotic range.
-  \end{enumerate}
-  This is more serious than one might think.
-It is easy to verify that changing the force-field changes the final result, but how much this change is due to computational errors?
-If we cannot detect, when the computational error is irrelevant and estimate the discretization error, then we cannot compute the modelling error.
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=12cm]{charmm36tol04.pdf}
-%   \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
-The force-field is CHARMM36 and the constraint equations are not solved with great accuracy, i.e.
-${\tt tol} = 10^{-4}$.
-} \label{fig:charmm36tol04}
-% \end{figure}
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=12cm]{charmm36tol12.pdf}
-%    \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
-The force-field is CHARMM36 and the constraint equations are solved with great accuracy, i.e.
-${\tt tol} = 10^{-12}$.} \label{fig:charmm36tol12}
-% \end{figure}
 
@@
 
-In this section we contemplate the difference between the succesful and unsuccesful application of the theory.
-We identify two key conditions that are not always satisfied by the GROMACS library and we demonstrate that they are necessary for the succesful application of Richardson's techniques.
+In this section we identify two conditions that are not always satisfied by the GROMACS library and we demonstrate that they are necessary for the succesful application of Richardson's techniques.
 
@@
 In the case of Euler's explicit method, the relevant equation is linear, but in general this is nonlinear equation.
-The function {\tt maxrange\_rk1\_mwe1} and it is companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h \text{tol}$ using a wide range of tolerance {\tt tol}.
+The function {\tt maxrange\_rk1\_mwe1} and it is companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h\cdot\text{tol}$ using a wide range of tolerance {\tt tol}.
 It is instructive to observe the consequences of solving this equation inaccurately.
-Figures \ref{fig:maxrange_tol11} ($\text{tol} = 2^{-11}$) and Figures \ref{fig:maxrange_tol25} ($\text{tol} = 2^{-25}$) show the evolution of Richardson's fraction for two different values of the tolerance {\tt tol}.
-When the tolerance is large, there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
-When the tolerance is small, there is evidence to suggest that asymptotic error expansion exists and an asymptotic range can easily be identified for shells of type G1, G5 and G8.
+Figures \ref{fig:maxrange_rk2_tol11} ($\text{tol} = 2^{-11}$) and Figures \ref{fig:maxrange_rk2_tol25} ($\text{tol} = 2^{-25}$) show the evolution of Richardson's fraction for two different values of the tolerance {\tt tol}.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{maxrange_rk2_tol11.png} \caption{The evolution of $F_h$ for the maximum range of 6 different shells using using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-11}$.}
+  \label{fig:maxrange_rk2_tol11}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{maxrange_rk2_tol25.png} \caption{The evolution of $F_h$ for the maximum range of 6 different shells computed using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-25}$.}
+  \label{fig:maxrange_rk2_tol25}
+\end{figure}
+
+When the tolerance is large (Figure~\ref{maxrange_rk2_tol11.png}) there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
+When the tolerance is small (Figure~\ref{maxrange_rk2_tol25.png}), there is evidence to suggest that asymptotic error expansion exists and an asymptotic range can easily be identified for shells of type G1, G5 and G8.
 If we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is solve all central equations as accurately as the hardware will allow.
@@
 
-The GROMACS library is highly optimized in order to reduced the time-to-solution.
-In the case of the force-field calculation it is common to ignore the interaction between particles that are judged to be sufficiently far away.
-In practice, this is done by setting individual force-fields to zero outside of a sufficiently large ball.
-There is more than way to achieve this goal and the online documentation suggest that GROMACS operates with force-fields that are of class $C^1$.
-Why is this a concern in the context of practical error estimation?
-Proving that an asymptotic error expansion exist is often an exercise in manipulating Taylor expansions and this is only possible when the functions are sufficiently differentiable.
-We have designed a sequence of numerical experiment that illustrate this exact point.
-We consider a system of $m$ identical ions moving in a liquid.
-They repel each other according to Coulumb's law and they are pulled towards the origo according to Hooke's law.
-The friction between each ion and the fluid is proportional to its current velocity.
-This is a simple setup that ensure that particles will ultimate come to rest in a position where all forces balances each other.
-Can we estimate the value of the kinetic energy at at specific time $T$?
-This is the question that we explore in a sequence of scripts.
-Let $f$ denote the force-field emenating from an particle with charge $q$, i.e
+It is common to ignore the interaction between atoms that are far away.
+This can be done by setting force-fields to zero outside of a sufficiently large ball.
+There is more than way to accomplish this effect and the documentation for GROMACS 2021 discusses the use of force-fields that are of class $C^1$ rather than $C^{\infty}$.
+
+In order to explore the importance of smoothness we have conducted a sequence of simple experiments.
+We simulated the motion of a small set of identical ions moving in a liquid.
+The ions repell each other electrostatically, but they are pulled towards the origo by independent and identical springs that obey Hooke's law.
+The friction between each ion and the liquid is proportional to its velocity.
+This ensures the ions ultimately come to rest in a stable configuration.
+Let $f$ denote the forcefield generated by an ion located at $0$ with charge $q$. Then
 \begin{equation}
-  f(\bm{r}) = - c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
+  f(\br{r}) = c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
 \end{equation}
-In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$. In {\tt iontrap\_mwe2} we replace $f$ with
+where $c > 0$ is a suitable constant.
+The script {\tt iontrap\_mwe1} does not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedoron with edge length $\rho > 0$.
+In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$.
+In {\tt iontrap\_mwe2} and {\tt iontrap\mwe4} we replace $f$ with
 \begin{equation}
-  f_1(\bm{r}) = f(\bm{r}) g_1(r), \quad g_1(r) = \begin{cases} 1  & 0 \leq r < \rho \\ 0 & \rho \leq r \end{cases}
+  f_k(\bm{r}) = f(\bm{r}) g_k(r)
 \end{equation}
-This introduces jump discontinuites into the simulation.
-In {\tt iontrap\_mwe4} the transition to zero is done smoothly in the region from $0.5 \rho < r < 0.95 r$.
-It is clear that changing the force-fields is equivalent to physical changes, but can we estimate the discretization error?
+where $g_k$ is a switching function that assumes values in $[0,1]$.
+The function $g_2$ has a jump discontinuity and satisfies $g_1(r) = 1$ for $r < \rho$ and $g_1(r) = 0$ for $r \ge \rho$.
+The function $g_4$ is of class $C^\infty$ and satifies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.
+It is clear that changing the force-fields must impact the simulation, but can we actually estimate the discretization error for the kinetic energy?
 Figure \ref{fig:iontrap_mwe1}, \ref{fig:iontrap_mwe2}, \ref{fig:iontrap_mwe4} show the evolution of Richardson's fraction for the total kinetic energy at the end of each simulation.
@@
   \centering
-  \includegraphics[width=10cm]{iontrap_mwe1.pdf}
+  \includegraphics[width=\linewidth]{iontrap_mwe1.pdf}
   \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$}
@@
   \centering
-  \includegraphics[width=10cm]{iontrap_mwe2.pdf}
+  \includegraphics[width=\linewidth]{iontrap_mwe2.pdf}
   \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$.
@@
   \centering
-  \includegraphics[width=10cm]{iontrap_mwe4.pdf}
+  \includegraphics[width=\linewidth]{iontrap_mwe4.pdf}
   \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$.

### 2024-05-11 report/experiments/maxrange/src/maxrange_rk2.m

@@ -120,7 +120,7 @@ for t=1:numtol
         sf.Title=title(tit{j});
 
         % Set the fontsize for the axes
-        ax=fig.CurrentAxes; ax.FontSize=20; ax.GridLineWidth=2;
+        ax=fig.CurrentAxes; ax.FontSize=20; % ax.GridLineWidth=2;
 
         % Set the linewidth for the plot
         plt.LineWidth=2;

### 2024-05-11 report/src/draft.tex
Mainly changes to figures to save space and allow for black and white printing

@@
 \begin{subfigure}[h]{0.49\linewidth}
-\includegraphics[width=\linewidth]{rint_mwe1a.png}
+\includegraphics[width=\linewidth]{rint_mwe1a.pdf}
 \caption{The evolution of $F_h$} \label{fig:rint_mwe1a}
@@
 \begin{subfigure}[h]{0.49\linewidth}
-\includegraphics[width=\linewidth]{rint_mwe1b.png}
+\includegraphics[width=\linewidth]{rint_mwe1b.pdf}
 \caption{The size of $E_h$ and the accuracy of $R_h$} \label{fig:rint_mwe1b}
@@
 \begin{subfigure}[h]{0.49\linewidth}
-\includegraphics[width=\linewidth]{rint_mwe2a.png}
+\includegraphics[width=\linewidth]{rint_mwe2a.pdf}
 \caption{The evolution of $F_h$} \label{fig:rint_mwe2a}
@@
 \begin{subfigure}[h]{0.49\linewidth}
-\includegraphics[width=\linewidth]{rint_mwe2b.png}
+\includegraphics[width=\linewidth]{rint_mwe2b.pdf}
 \caption{The size of $E_h$ and the accuracy of $R_h$.} \label{fig:rint_mwe2b}
@@
   \centering
-  \includegraphics[width=\linewidth]{maxrange_rk1_tol53.png}
+  \includegraphics[width=\linewidth]{maxrange_rk1_tol53.pdf}
   \caption{The evolution of Richardson's fraction corresponding to the maximum range of 6 different shells fired from the D-20 howitzer.} \label{fig:maxrange_rk1_fraction_tol53}
@@
 The function {\tt maxrange\_rk1\_mwe1} and it is companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h\cdot\text{tol}$ using a wide range of tolerance {\tt tol}.
-It is instructive to observe the consequences of solving this equation inaccurately.
-Figures \ref{fig:maxrange_rk2_tol11} ($\text{tol} = 2^{-11}$) and Figures \ref{fig:maxrange_rk2_tol25} ($\text{tol} = 2^{-25}$) show the evolution of Richardson's fraction for two different values of the tolerance {\tt tol}.
 
@@
 
+It is instructive to observe the consequences of solving this equation inaccurately.
+Figures \ref{fig:maxrange_rk2_tol11} ($\text{tol} = 2^{-11}$) and Figures \ref{fig:maxrange_rk2_tol25} ($\text{tol} = 2^{-25}$) show the evolution of Richardson's fraction for two different values of the tolerance {\tt tol}.
+
 \begin{figure}
@@
 
-When the tolerance is large (Figure~\ref{maxrange_rk2_tol11.png}) there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
-When the tolerance is small (Figure~\ref{maxrange_rk2_tol25.png}), there is evidence to suggest that asymptotic error expansion exists and an asymptotic range can easily be identified for shells of type G1, G5 and G8.
+When the tolerance is large (Figure~\ref{fig:maxrange_rk2_tol11}) there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
+When the tolerance is small (Figure~\ref{fig:maxrange_rk2_tol25}), there is evidence to suggest that asymptotic error expansion exists and an asymptotic range can easily be identified for shells of type G1, G5 and G8.
 If we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is solve all central equations as accurately as the hardware will allow.
@@
 \begin{equation}
-  f(\br{r}) = c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
+  f(\bm{r}) = c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
 \end{equation}
@@
 In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$.
-In {\tt iontrap\_mwe2} and {\tt iontrap\mwe4} we replace $f$ with
+In {\tt iontrap\_mwe2} and {\tt iontrap\_mwe4} we replace $f$ with
 \begin{equation}

### 2024-05-11 report/experiments/integration/src/rint_mwe1.m

@@ -15,8 +15,8 @@ rule=@(y,a,b,N)trapezoid(y,a,b,N);
 % Theoretical order of the method 
 p=2; 
 
-% Number of refinements
-kmax=21; 
+% Number of approximations
+kmax=20; 
 
 % True value of the integral 
 val=exp(1)-1;
@@ -41,13 +41,15 @@ mypath='../experiments/integration/fig/';
 figa=figure(); plta=plot_fraction(data,p);
 
 % Set the fontsize for the axes
-ax=figa.CurrentAxes; ax.FontSize=20; ax.GridLineWidth=2;
+axa=figa.CurrentAxes; axa.FontSize=20; % axa.GridLineWidth=2;
 
 % Set the linewidth for the plot
 plta.LineWidth=2;
 
+% figa.Position=[900 800 700 400]
+
 % Save the figure
-fname='rint_mwe1a.png'; saveas(figa,strcat(mypath,fname));
+fname='rint_mwe1a.eps'; saveas(figa,strcat(mypath,fname));
 
 % ////////////////////////////////////////
 % Generate Figure rint_mwe1b
@@ -61,18 +63,20 @@ k=data(:,1); err=data(:,5); rel=data(:,6);
 
 % Plot the data
 hold on;
-pltb1=plot(k,log10(abs(err)));
-pltb2=plot(k,log10(abs(rel)));
+pltb1=plot(k,log10(abs(err)),'b-');
+pltb2=plot(k,log10(abs(rel)),'b:');
 axis([0 kmax -14 2]);
 grid; lgdb=legend('log_{10}(|E_h|)','log_{10}(|E_h-R_h|/|E_h|)', ...
-    'Location','North');
+    'Location','SouthWest');
 
 % Set the fontsize for the axes
-axb=figb.CurrentAxes; axb.FontSize=20; axb.GridLineWidth=2;
+axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
 
 % Set the linewidth for the plot
-pltb1.LineWidth=2; pltb2.LineWidth=2;
+pltb1.LineWidth=4; pltb2.LineWidth=4;
+
+% figb.Position=[900 800 700 400]
 
 % Save the figure
-fname='rint_mwe1b.png';
+fname='rint_mwe1b.eps';
 saveas(figb,strcat(mypath,fname));

### 2024-05-11 report/experiments/integration/src/rint_mwe2.m

@@ -18,7 +18,7 @@ rule=@(y,a,b,N)trapezoid(y,a,b,N);
 % Theoretical order of the method 
 p=1.5; 
 
-% Number of refinements
+% Number of approximations
 kmax=26; 
 
 % True value of the integral 
@@ -44,13 +44,13 @@ mypath='../experiments/integration/fig/';
 figa=figure(); plta=plot_fraction(data,p);
 
 % Set the fontsize for the axes
-ax=figa.CurrentAxes; ax.FontSize=20; ax.GridLineWidth=2;
+axa=figa.CurrentAxes; axa.FontSize=20; % axa.GridLineWidth=2;
 
 % Set the linewidth for the plot
 plta.LineWidth=2;
 
 % Save the figure
-fname='rint_mwe2a.png';
+fname='rint_mwe2a.eps';
 saveas(figa,strcat(mypath,fname));
 
 % ////////////////////////////////////////
@@ -65,18 +65,18 @@ k=data(:,1); err=data(:,5); rel=data(:,6);
 
 % Plot the data
 hold on;
-pltb1=plot(k,log10(abs(err)));
-pltb2=plot(k,log10(abs(rel)));
+pltb1=plot(k,log10(abs(err)),'b-');
+pltb2=plot(k,log10(abs(rel)),'b:');
 axis([0 kmax -14 2]);
 grid; lgdb=legend('log_{10}(|E_h|)','log_{10}(|E_h-R_h|/|E_h|)', ...
-    'Location','North');
+    'Location','SouthWest');
 
 % Set the fontsize for the axes
-axb=figb.CurrentAxes; axb.FontSize=20; axb.GridLineWidth=2;
+axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
 
 % Set the linewidth for the plot
-pltb1.LineWidth=2; pltb2.LineWidth=2;
+pltb1.LineWidth=4; pltb2.LineWidth=4;
 
 % Save the figure
-fname='rint_mwe2b.png';
+fname='rint_mwe2b.eps';
 saveas(figb,strcat(mypath,fname));

### 2024-05-11 report/experiments/maxrange/src/maxrange_rk1.m

@@ -102,36 +102,78 @@ end
 % Allocate space for the data
 data=zeros(kmax,4,numshell,numtol);
 
-% Process the data and generate the plots
-for t=1:numtol
-    % Get a handle to a new figure
-    fig=figure();
-    % Plan where to put the figure
-    q=floor((t-1)/4); r=(t-1)-4*q;
-    % Position on screen
-    fig.Position=[10+r*300 650-q*600 840 630];
-    % Loop over the shell types
-    for j=1:numshell
-        % Apply Richardson extrapolation
-        data(:,:,j,t)=richardson(raw_range(:,j,t),p);
-        % Plot the fraction
-        sf=subplot(2,3,j); plt=plot_fraction(data(:,:,j,t),p);
+if (0==1)
+
+    % Process all data and generate plots
+    for t=1:numtol
+        % Get a handle to a new figure
+        fig=figure();
+        % Plan where to put the figure
+        q=floor((t-1)/4); r=(t-1)-4*q;
+        % Position on screen
+        % fig.Position=[10+r*300 650-q*600 840 640];
+        fig.Position=[10+r*300 650-q*600 840 320];
+        % Loop over the shell types
+        for j=1:6
+            % Apply Richardson extrapolation
+            data(:,:,j,t)=richardson(raw_range(:,j,t),p);
+            % Plot the fraction
+            sf=subplot(2,3,j); plt=plot_fraction(data(:,:,j,t),p);
+            % Set the title
+            sf.Title=title(tit{j});
+            
+            % Set the fontsize for the axes
+            ax=fig.CurrentAxes; ax.FontSize=10; % ax.GridLineWidth=2;
+            
+            % Set the linewidth for the plot
+            plt.LineWidth=2;
+        end
         % Set the title
-        sf.Title=title(tit{j});
-
-        % Set the fontsize for the axes
-        ax=fig.CurrentAxes; ax.FontSize=20; % ax.GridLineWidth=2;
-
-        % Set the linewidth for the plot
-        plt.LineWidth=2;
+        sgtitle(['log2(tol) =' num2str(log2(tol(t)),'%d')]);
+        
+        % Save the figures
+        fname=strcat(figpath,'maxrange_rk1_tol',...
+            num2str(-log2(tol(t)),'%02d'),'.eps');
+        saveas(fig,fname);
     end
-    % Set the title
-    sgtitle(['log2(tol) =' num2str(log2(tol(t)),'%d')]);
-
-    % Save the figures 
[…]

### 2024-05-11 report/experiments/maxrange/src/maxrange_rk2.m

@@ -102,37 +102,81 @@ end
 % Allocate space for the data
 data=zeros(kmax,4,numshell,numtol);
 
-% Process the data and generate the plots
-for t=1:numtol
-    % Get a handle to a new figure
-    fig=figure();
-    % Plan where to put the figure
-    q=floor((t-1)/4); r=(t-1)-4*q;
-    % Position on screen
-    fig.Position=[10+r*300 650-q*600 840 630];
-    % Loop over the shell types
-    for j=1:numshell
-        % Apply Richardson extrapolation
-        data(:,:,j,t)=richardson(raw_range(:,j,t),p);
-        % Plot the fraction
-        sf=subplot(2,3,j); plt=plot_fraction(data(:,:,j,t),p);
+
+if (0==1)
+    
+    % Process all data and generate the plots
+    for t=1:numtol
+        % Get a handle to a new figure
+        fig=figure();
+        % Plan where to put the figure
+        q=floor((t-1)/4); r=(t-1)-4*q;
+        % Position on screen
+        fig.Position=[10+r*300 650-q*600 840 640];
+        % Loop over the shell types
+        for j=1:numshell
+            % Apply Richardson extrapolation
+            data(:,:,j,t)=richardson(raw_range(:,j,t),p);
+            % Plot the fraction
+            sf=subplot(2,3,j); plt=plot_fraction(data(:,:,j,t),p);
+            % Set the title
+            sf.Title=title(tit{j});
+            
+            % Set the fontsize for the axes
+            ax=fig.CurrentAxes; ax.FontSize=10; % ax.GridLineWidth=2;
+            
+            % Set the linewidth for the plot
+            plt.LineWidth=2;
+        end
         % Set the title
-        sf.Title=title(tit{j});
-
-        % Set the fontsize for the axes
-        ax=fig.CurrentAxes; ax.FontSize=20; % ax.GridLineWidth=2;
-
-        % Set the linewidth for the plot
-        plt.LineWidth=2;
+        sgtitle(['log2(tol) =' num2str(log2(tol(t)),'%d')]);
+        
+        % Save the figures
+        fname=strcat(figpath,'maxrange_rk2_tol',...
+            num2str(-log2(tol(t)),'%02d'),'.eps');
+        saveas(fig,fname);
     end
-    % Set the title
-    sgtitle(['log2(tol) =' num2str(log2(tol(t)),'%d')]);
-
-    % Save the figures 
-    fname=strcat(figpath,'maxrange_rk2_tol',...
[…]

### 2024-05-11 report/scripts/gatherfig.sh

@@ -3,6 +3,13 @@ find experiments -name \*.eps -exec cp {} fig \;
 
 # Crop white space from MATLAB eps files
 cd fig
+ps2pdf -DEPSCrop rint_mwe1a.eps rint_mwe1a.pdf
+ps2pdf -DEPSCrop rint_mwe1b.eps rint_mwe1b.pdf
+ps2pdf -DEPSCrop rint_mwe2a.eps rint_mwe2a.pdf
+ps2pdf -DEPSCrop rint_mwe2b.eps rint_mwe2b.pdf
+ps2pdf -DEPSCrop maxrange_rk1_tol53.eps maxrange_rk1_tol53.pdf
+ps2pdf -DEPSCrop maxrange_rk2_tol11.eps maxrange_rk2_tol11.pdf
+ps2pdf -DEPSCrop maxrange_rk2_tol25.eps maxrange_rk2_tol25.pdf
 ps2pdf -DEPSCrop charmm36tol04.eps charmm36tol04.pdf
 ps2pdf -DEPSCrop charmm36tol12.eps charmm36tol12.pdf
 ps2pdf -DEPSCrop oplsaaltol04.eps oplsaaltol04.pdf

### 2024-05-11 report/src/draft.tex
Minor updates to the text and juggling with the figures

@@
 \begin{abstract} We consider the problem of estimating the error when solving a system of differential algebraic equations.
-Richardson extrapolation is a standard technique that can be used to judge when rounding errors are irrelevant and estimate the discretization error.
-We have simulated molecular dynamics with constraints using the GROMACS library and found that the output is not always ameable to Richardson's analysis.
-We derive and illustrate Richardson's technique using a variety of numerical experiments.
+Richardson extrapolation is a classical technique that can be used to judge when rounding errors are irrelevant and estimate the discretization error.
+We have simulated molecular dynamics with constraints using the GROMACS library and found that the output is not always ameable to Richardson's method.
+We derive and illustrate Richardson's method using a variety of numerical experiments.
 We identify two necessary conditions that are not always satisfied by the GROMACS library.
@@
 
-\section{Motivation} Consider the problem resolving the motion of a system of atoms moving in a force-field $f$ subject to a set of constraints $g$.
+\section{Motivation} Consider the problem resolving the motion of a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints $g$.
 In this case, Newton's 2nd law takes the form the following system of differential algebraic equations
@@
 The nonsingular diagonal matrix $\bm{M}$ lists the masses of the atoms.
-The function $\bm{G}$ is the Jacobian of the constraint function $\bm{g}$.
+The function $\bm{G}$ is the Jacobian of the constraint function $\bm{g}$ and $\bm{\lambda}$ is a vector of Lagrange multipliers.
 In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm.
@@
 \end{equation}
-We conclude that $T_0 - T(f)$ is small, if the computational error $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the discretization error $T(\bm{f}) - A_h(\bm{f})$ are both small.
-If we cannot assert control over these errors, then we cannot say with certainty that our model is a good approximation of the physical reality.
+We conclude that $T_0 - T(\bm{f})$ is small, if the computational error $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the discretization error $T(\bm{f}) - A_h(\bm{f})$ are both small.
+If we cannot assert control over these errors, then we cannot say with certainty that our model delivers a good approximation of the physical reality.
 
@@
 In the past, we have sought to improve the solution of nonlinear constraint equations in the context of molecular dynamics by integrating Newton's method into the GROMACS library \cite{villelas2023accurate}
-Our goal has been to reduce the computational error at low cost in order to improve the long-term stability of constrained simulations of molecular dynamics.
+Our goal has been the efficient reduction of the computational error in order to improve the long-term stability of constrained simulations of molecular dynamics at low cost.
 We have now found that the output of GROMACS is not always amenable to Richardson's methods.
@@
 \begin{proof}
-  By assumption, there is a function $g(h)$ 
+  By assumption, there is a function $h \rightarrow g(h)$ 
   \begin{equation}
@@
   \end{equation}
-  and constants $C>0$ and $h_0 > 0$ such that
+  as well as constants $C>0$ and $h_0 > 0$ such that
   \begin{equation}
@@
 \end{proof}
-We shall now demonstrate how to investigate the existence of an asymptotic error expansion experimentally.
+We shall now show how to demonstrate the existence of an asymptotic error expansion experimentally.
 We define Richardson's fraction $F_h$ using the expression
@@
 The behavior of the function $h \rightarrow F_h$ is described by the following theorem.
-\begin{theorem} Assume that $T-A_h$ satisfies equation \eqref{equ:aex} and let $(m,n)$ be given by
+\begin{theorem} If $E_h$ satisfy equation \eqref{equ:aex} and if $(m,n)$ is given by
   \begin{equation}
-    m = q - p, \quad n = r - p.
+    m = q - p, \quad n = r - p,
   \end{equation}
-  Then Richardson's fraction satisfies
+  then Richardson's fraction $F_h$ satisfies
   \begin{equation}
@@
 However, it is often possible to assert that $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the error $T - \hat{A}_h$.
-We shall now demonstrate this procedure.
+We shall now demonstrate the procedure.
 To this end, we consider the familiar problem of computing definite integrals
@@
 We mention in passing that low order methods are more practical than high order methods in the sense that low order methods tend to have asymptotic ranges that are larger than high order methods.
-% This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small.
+This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small.
 This issue is more acute for high order methods than for low order methods, because $A_h$ tends to $T$ more rapidly for high order methods than for low order methods.
@@
   \includegraphics[width=\linewidth]{oplsaaltol04.pdf}
-  \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
-The force-field is OPLSAAL and the constraint equations are not solved with great accuracy, i.e.
-${\tt tol} = 10^{-4}$.
-} \label{fig:oplsaaltol04}
+  \caption{The evolution of the kinetic and potential energy of a system as the number of time-steps used to cover $1$ ps of real time.} \label{fig:oplsaaltol04}
 \end{figure}
@@
 
-%The function {\tt gromacs\_fit} does a linear least squares fit of the total energy {\tt etol} to two models $M_1$ and $M_2$ where model $M_p$ assumes that ${\tt etol} = T + C_1n + C_2/n^p$ for suitable constants $T, C_1, C_2$.
-Since the SHAKE algorithm is 2nd order accurate in the time step, one would expect model $M_2$ to provide the best fit, but for this data model $M_1$ is a marginally better.
-While mildly interesting, this issue is vastly less important than the final point.
+Thirdly, if the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} well inside the asymptotic range.
+Why is this?
+If we where in the asymptotic range, then the error $T - \hat{A}_h$ would very nearly obey a power law and $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some power $p>0$.
+In particular, the $\hat{A}_h$  would behave in be \emph{monotone} manner and the tiny oscillations that we have recorded would not be present.
 
@@
   \includegraphics[width=\linewidth]{oplsaaltol12.pdf}
-   \caption{The evolution of the mechanical energy of a molecule as a function of the number of time-steps used to cover $1$ps of real time.
-The force-field is OPLSAAL and the constraint equations are solved with great accuracy, i.e.
-${\tt tol} = 10^{-12}$.} \label{fig:oplsaaltol12}
+   \caption{The evolution of the kinetic and potential energy of a system as the number of time-steps used to cover $1$ ps of real time.} \label{fig:oplsaaltol12}
 \end{figure}
-
-Thirdly, if the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} in the asymptotic range.
-Why is this?
-If we where in the asymptotic range, then the computed energies $\hat{A}_h$ would follow a power-law and $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some power $p>0$.
-In particular, the behaviour would be \emph{monotone} and the tiny oscillations that we have recorded would not be present.
 
@@
 
-\section{The need for sufficient accuracy}
+\subsection{The need for sufficient accuracy}
 
-The output of GROMACS is clearly dependent how accurate we solve the constraint equations.
-When computing the range of a howitzer it necessary to adjust the final time-step to place the shell directly on the ground.
-In the case of Euler's explicit method, the relevant equation is linear, but in general this is nonlinear equation.
+When computing the range of a howitzer it necessary to adjust the final time-step to place the shell on the ground with great accuracy.
+In the case of Euler's explicit method, the relevant equation is linear, but in general it is a nonlinear equation.
 The function {\tt maxrange\_rk1\_mwe1} and it is companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h\cdot\text{tol}$ using a wide range of tolerance {\tt tol}.
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{maxrange_rk2_tol11.png} \caption{The evolution of $F_h$ for the maximum range of 6 different shells using using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-11}$.}
+  \includegraphics[width=\linewidth]{maxrange_rk2_tol11.pdf} \caption{The evolution of $F_h$ for the maximum range of 3 different shells using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-11}$.}
   \label{fig:maxrange_rk2_tol11}
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{maxrange_rk2_tol25.png} \caption{The evolution of $F_h$ for the maximum range of 6 different shells computed using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-25}$.}
+  \includegraphics[width=\linewidth]{maxrange_rk2_tol25.pdf} \caption{The evolution of $F_h$ for the maximum range of 6 different shells computed using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-25}$.}
   \label{fig:maxrange_rk2_tol25}
@@
 When the tolerance is large (Figure~\ref{fig:maxrange_rk2_tol11}) there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
-When the tolerance is small (Figure~\ref{fig:maxrange_rk2_tol25}), there is evidence to suggest that asymptotic error expansion exists and an asymptotic range can easily be identified for shells of type G1, G5 and G8.
-If we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is solve all central equations as accurately as the hardware will allow.
+When the tolerance is small (Figure~\ref{fig:maxrange_rk2_tol25}), there is evidence to suggest that asymptotic error expansion exists, but it is not trivial to determine an asymptotic range.
+Regardless, it is clear that if we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is solve all equations as accurately as the hardware will allow.
 
-\section{The need for sufficient smoothness}
+\subsection{The need for sufficient smoothness}
 
-It is common to ignore the interaction between atoms that are far away.
+In molecular dynamics, it is common to ignore the interaction between atoms that are far away.
 This can be done by setting force-fields to zero outside of a sufficiently large ball.
@@
 
-In order to explore the importance of smoothness we have conducted a sequence of simple experiments.
-We simulated the motion of a small set of identical ions moving in a liquid.
-The ions repell each other electrostatically, but they are pulled towards the origo by independent and identical springs that obey Hooke's law.
-The friction between each ion and the liquid is proportional to its velocity.
-This ensures the ions ultimately come to rest in a stable configuration.
-Let $f$ denote the forcefield generated by an ion located at $0$ with charge $q$. Then
-\begin{equation}
-  f(\bm{r}) = c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
-\end{equation}
-where $c > 0$ is a suitable constant.
-The script {\tt iontrap\_mwe1} does not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedoron with edge length $\rho > 0$.
-In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$.
-In {\tt iontrap\_mwe2} and {\tt iontrap\_mwe4} we replace $f$ with
-\begin{equation}
-  f_k(\bm{r}) = f(\bm{r}) g_k(r)
-\end{equation}
-where $g_k$ is a switching function that assumes values in $[0,1]$.
-The function $g_2$ has a jump discontinuity and satisfies $g_1(r) = 1$ for $r < \rho$ and $g_1(r) = 0$ for $r \ge \rho$.
-The function $g_4$ is of class $C^\infty$ and satifies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.
-It is clear that changing the force-fields must impact the simulation, but can we actually estimate the discretization error for the kinetic energy?
-Figure \ref{fig:iontrap_mwe1}, \ref{fig:iontrap_mwe2}, \ref{fig:iontrap_mwe4} show the evolution of Richardson's fraction for the total kinetic energy at the end of each simulation.
-
-\begin{figure}
+\begin{figure}[h!]
   \centering
@@
 
-When the force-fields are not perturbed, Figure~\ref{fig:iontrap_mwe1}, there experiments supports the conjecture that the exists an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate the dynamical system.
+In order to explore the importance of smoothness we have conducted a sequence of simple experiments.
+We simulated the motion of a small set of identical ions moving in a liquid.
+The ions repel each other electrostatically, but they are pulled towards the origo by independent and identical springs that obey Hooke's law.
+The friction between each ion and the liquid is proportional to its velocity.
+This ensures the ions eventually come to rest in a stable configuration.
+Let $\bm{f}$ denote the forcefield generated by an ion located at $0$ with charge $q$. Then
+\begin{equation}
+  \bm{f}(\bm{r}) = c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
+\end{equation}
+where $c > 0$ is a suitable constant.
+The script {\tt iontrap\_mwe1} does not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedoron with edge length $\rho > 0$.
+The scripts {\tt iontrap\_mwe2} and {\tt iontrap\_mwe4} replace $f$ with
+\begin{equation}
+  f_k(\bm{r}) = f(\bm{r}) g_k(r), \quad k \in \{2, 4\{
+\end{equation}
+where $g_k$ is a switching function that assumes values in $[0,1]$.
+The function $g_2$ has a jump discontinuity and satisfies $g_1(r) = 1$ for $r < \rho$ and $g_1(r) = 0$ for $r \ge \rho$.
+The function $g_4$ is of class $C^\infty$ and satifies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.
+It is clear that changing the force-fields must impact the simulation, but can we actually estimate the discretization error for the kinetic energy?
+Figure \ref{fig:iontrap_mwe1}, \ref{fig:iontrap_mwe2}, \ref{fig:iontrap_mwe4} show the evolution of Richardson's fraction for the kinetic energy at the end of each simulation.
 
-\begin{figure}
+When the force-fields are not perturbed, Figure~\ref{fig:iontrap_mwe1}, the experiments supports the conjecture that the exists an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate the dynamical system.
+
+\begin{figure}[h!]
   \centering
@@
 When the force-fields are truncated and jump-discontinuities are introduced, Figure~\ref{fig:iontrap_mwe2}, there is no evidence to suggest that there exists an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
+We cannot assert that the computational error is irrelevant and without an estimate of the discretization error we cannot quantify the impact of changing the force-field.
 

### 2024-05-11 report/experiments/gromacs/src/gromacs_figures.m

@@ -18,10 +18,10 @@ for i=1:4
     aux=c{i}.data; field=c{i}.field; tol=c{i}.tol;
 
     % Rename the columns to reduce the chance of misprints
-    ekin=aux(:,3); epot=aux(:,4); etol=ekin+epot;
+    ekin=aux(:,3); epot=aux(:,4); % etol=ekin+epot;
 
     % Call specialized plotting routine
-    [fig, ax]=gromacs_plot(field, tol, n, ekin, epot, etol);
+    [fig, ax]=gromacs_plot(field, tol, n, ekin, epot);
 
     fname=strcat(field,'tol',num2str(round(-log10(tol)),'%02d'),'.eps');
     saveas(fig,strcat(figpath,fname));

### 2024-05-11 report/experiments/gromacs/src/gromacs_plot.m

@@ -1,36 +1,36 @@
-function [fig, ax]=gromacs_plot(field, tol, n, ekin, epot, etol)
+function [fig, ax]=gromacs_plot(field, tol, n, ekin, epot)
 
-fig=figure(); fig.Position=[100 800 1800 400];
+fig=figure(); fig.Position=[100 800 1800 600];
 
-subplot(1,3,1); 
+subplot(1,2,1); 
 plt1=plot(n,ekin); grid; 
 xlabel('number of steps'); 
 ylabel('Kinetic energy (kJ/mol)'); 
 
 % Set the fontsize for the axes
-ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
+ax=fig.CurrentAxes; ax.FontSize=14; ax.LineWidth=2;
 
-subplot(1,3,2); 
+subplot(1,2,2); 
 plt2=plot(n,epot); grid;
 
 xlabel('number of steps');
 ylabel('Potential energy (kJ/mol)');
 
 % Set the fontsize for the axes
-ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
+ax=fig.CurrentAxes; ax.FontSize=14; ax.LineWidth=2;
 
-subplot(1,3,3); 
-plt3=plot(n,etol); grid;
-xlabel('number of steps');
-ylabel('Total energy (kJ/mol)');
-
-% Set the fontsize for the axes
-ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
+% subplot(1,3,3); 
+% plt3=plot(n,etol); grid;
+% xlabel('number of steps');
+% ylabel('Total energy (kJ/mol)');
+% 
+% % Set the fontsize for the axes
+% ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
 
 % Set the linewidth for the plot
 plt1.LineWidth=2;
 plt2.LineWidth=2;
-plt3.LineWidth=2;
+% plt3.LineWidth=2;
 
 % Set a common title
 tit=sgtitle(['Forcefield = ' field '     constraint tolerance = ' num2str(tol,'%.2e')]);

### 2024-05-12 report/src/draft.tex
Fixed misprints

@@
 The fundamental problem is that we cannot compute the exact values of $T(\bm{f})$ and $A_h(\bm{f})$.
-We must contend with the fact that we cannot solve the constraint equations exactly nor can we avoid rounding errors.
+We must contend with the fact that we can hardly solve the constraint equations exactly nor can we avoid rounding errors.
 Let $\hat{A}_h(\bm{f}, \tau, u)$ denote the value returned by our computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$.
 Suppose that $T_0 - A_h(\bm{f}, \tau, u)$ is small.
-Can we show that $T_0 - T(\bm{f})$ is small?
+Can we assert that $T_0 - T(\bm{f})$ is small?
 The triangle inequality implies the following inequality
@@
 We conclude that $T_0 - T(\bm{f})$ is small, if the computational error $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the discretization error $T(\bm{f}) - A_h(\bm{f})$ are both small.
-If we cannot assert control over these errors, then we cannot say with certainty that our model delivers a good approximation of the physical reality.
+If we cannot control these errors, then we cannot say with certainty that our model delivers a good approximation of the physical reality.
 
@@
 This is due to the fact the function $h \rightarrow F_h$ suffers from subtractive cancellation when $h$ is sufficiently small.
-This issue is more acute for high order methods than for low order methods, because $A_h$ tends to $T$ more rapidly for high order methods than for low order methods.
+This issue is more acute for high order methods than for low order methods, because $A_h$ tends to $T$ much more rapidly for high order methods than for low order methods.
 
@@
 The initial search bracket is $[0,\pi/2$] and this bracket is systematically reduced in length until it is shorter than $\frac{\pi}{2}u$.
-
 The script will either read the raw data from a file or generate it from scratch.
@@
   \includegraphics[width=\linewidth]{maxrange_rk1_tol53.pdf}
-  \caption{The evolution of Richardson's fraction corresponding to the maximum range of 6 different shells fired from the D-20 howitzer.} \label{fig:maxrange_rk1_fraction_tol53}
+  \caption{The evolution of Richardson's fraction corresponding to the maximum range of 3 different shells fired from the D-20 howitzer.} \label{fig:maxrange_rk1_fraction_tol53}
 \end{figure}
@@
 The amplitude of the oscillations is reduced when $\tau = 10^{-12}$.
-We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expection the computed approximation unless $\tau$ is very small.
+We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expectation for the computed approximation unless $\tau$ is very small.
 
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{maxrange_rk2_tol25.pdf} \caption{The evolution of $F_h$ for the maximum range of 6 different shells computed using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-25}$.}
+  \includegraphics[width=\linewidth]{maxrange_rk2_tol25.pdf} \caption{The evolution of $F_h$ for the maximum range of 3 different shells computed using Heyn's method (\text{'rk2'}) and $\text{tol} = 2^{-25}$.}
   \label{fig:maxrange_rk2_tol25}

### 2024-05-12 report/src/draft.tex
Drafted a conclusion

@@
 Richardson extrapolation is a classical technique that can be used to judge when rounding errors are irrelevant and estimate the discretization error.
-We have simulated molecular dynamics with constraints using the GROMACS library and found that the output is not always ameable to Richardson's method.
-We derive and illustrate Richardson's method using a variety of numerical experiments.
+We have simulated molecular dynamics with constraints using the GROMACS library and found that the output is not always ameable to Richardson extrapolation.
+We derive and illustrate Richardson extrapolation using a variety of numerical experiments.
 We identify two necessary conditions that are not always satisfied by the GROMACS library.
-\keywords{}
+\keywords{error estimation, Richardson extrapolation, multi-body dynamics, GROMACS}
 \end{abstract}
@@
 %
-
-\section{Motivation} Consider the problem resolving the motion of a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints $g$.
-In this case, Newton's 2nd law takes the form the following system of differential algebraic equations
+\section{Motivation} Consider the problem simulating the motion of a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints $\bm{g}$.
+In this case, Newton's 2nd law takes the form of the following system of differential algebraic equations
 \begin{align}
-  \bm{q}'(t) &= \bm{v}(t) \\
-  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t) \\
+  \bm{q}'(t) &= \bm{v}(t), \\
+  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t), \\
   \bm{g}(\bm{q}(t)) &= \bm{0}. 
@@
 The function $\bm{G}$ is the Jacobian of the constraint function $\bm{g}$ and $\bm{\lambda}$ is a vector of Lagrange multipliers.
-In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm.
+In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm \cite{shake1977}
 It uses a pair of staggered grids with uniform step size $h$ and takes the form
@@
   \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
-  \bm{g}(\bm{q}_{n+1}) &= \bm{0}.
+  \bm{g}(\bm{q}_{n+1}) &= \bm{0}. \label{equ:constraint}
 \end{align}
-The constraint equation is a usually a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
+The constraint equation \eqref{equ:constraint} is usually a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
 Now let $T \in \mathbb{R}$ denote any target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
@@
 The fundamental problem is that we cannot compute the exact values of $T(\bm{f})$ and $A_h(\bm{f})$.
-We must contend with the fact that we can hardly solve the constraint equations exactly nor can we avoid rounding errors.
+We must contend with the fact that we cannot expect to solve the constraint equations exactly nor can we avoid rounding errors in general.
 Let $\hat{A}_h(\bm{f}, \tau, u)$ denote the value returned by our computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$.
 Suppose that $T_0 - A_h(\bm{f}, \tau, u)$ is small.
-Can we assert that $T_0 - T(\bm{f})$ is small?
-The triangle inequality implies the following inequality
+Can we conclude that $T_0 - T(\bm{f})$ is small?
+The triangle inequality provided the following bound:
 \begin{equation}
@@
 \end{equation}
-We conclude that $T_0 - T(\bm{f})$ is small, if the computational error $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the discretization error $T(\bm{f}) - A_h(\bm{f})$ are both small.
+We conclude that $T_0 - T(\bm{f})$ is small, if the \emph{computational error} $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the \emph{discretization error} $T(\bm{f}) - A_h(\bm{f})$ are both small.
 If we cannot control these errors, then we cannot say with certainty that our model delivers a good approximation of the physical reality.
 
-As we shall demonstrate, Richardson extrapolation \cite{richardson1927} can often be used to determine when computational errors are insignificant and estimate the discretization error.
-In the past, we have sought to improve the solution of nonlinear constraint equations in the context of molecular dynamics by integrating Newton's method into the GROMACS library \cite{villelas2023accurate}
-Our goal has been the efficient reduction of the computational error in order to improve the long-term stability of constrained simulations of molecular dynamics at low cost.
-We have now found that the output of GROMACS is not always amenable to Richardson's methods.
-In this paper we shall derive and illustrate the use of Richardson's method using a variety of numerical experiments.
-We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson's method.
-Our data and software is freely available from our GitHub along with every script and function needed to regenerate every number, table and figure from scratch.
-We encourage the reader to clone our repository and execute all MATLAB functions that are explicitly named in this manuscript, such as {\tt plot\_shells}.
+Richardson extrapolation is a classical technique that is widely used in computational science and engieering applications \cite{roache1998}.
+It can be used to estimate size of discretization errors or improve the accuracy of an existing solution \cite{zahari2018} and it has applications in event location \cite{mannshardt1978one} as well.
+As we shall demonstrate, Richardson extrapolation can often be used to determine when computational errors are insignificant.
+
+In this paper we derive and illustrate the use of Richardson extrapolation using a variety of numerical experiments.
+We demonstrate that the output of the GROMACS \cite{gromacs2005} library for molecular dynamics is not always ameable to Richardson extrapolation. 
+We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation
+Our data and software is freely available from our GitHub \cite{lorien2024ppam} along with every script and function needed to replicate every number, table and figure from scratch.
+The names of our MATLAB functions are written with a typewriter font, e.g., \texttt{plot\_shells}.
 
@@
 The function {\tt maxrange\_rk1\_mwe1} and it is companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h\cdot\text{tol}$ using a wide range of tolerance {\tt tol}.
-
 \begin{figure}
@@
 \end{figure}
-
 It is instructive to observe the consequences of solving this equation inaccurately.
 Figures \ref{fig:maxrange_rk2_tol11} ($\text{tol} = 2^{-11}$) and Figures \ref{fig:maxrange_rk2_tol25} ($\text{tol} = 2^{-25}$) show the evolution of Richardson's fraction for two different values of the tolerance {\tt tol}.
-
 \begin{figure}
@@
 \end{figure}
-
 When the tolerance is large (Figure~\ref{fig:maxrange_rk2_tol11}) there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
@@
 This can be done by setting force-fields to zero outside of a sufficiently large ball.
-There is more than way to accomplish this effect and the documentation for GROMACS 2021 discusses the use of force-fields that are of class $C^1$ rather than $C^{\infty}$.
+There is more than way to accomplish this effect and the documentation for GROMACS 2021 discusses its use of force-fields that are not of class $C^{\infty}$.
 
@@
   \includegraphics[width=\linewidth]{iontrap_mwe1.pdf}
-  \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$}
+  \caption{The evolution of $F_h$ for the total kinetic energy of a system of ions using integrators of order $p \in \{1,2,3,4\}$.}
   \label{fig:iontrap_mwe1}
@@
 
-In order to explore the importance of smoothness we have conducted a sequence of simple experiments.
-We simulated the motion of a small set of identical ions moving in a liquid.
+In order to explore the importance of smoothness we have simulated the motion of a set of identical ions moving in a liquid.
 The ions repel each other electrostatically, but they are pulled towards the origo by independent and identical springs that obey Hooke's law.
 The friction between each ion and the liquid is proportional to its velocity.
-This ensures the ions eventually come to rest in a stable configuration.
+The friction drains the energy and ensures that the ions eventually come to rest in a stable configuration.
 Let $\bm{f}$ denote the forcefield generated by an ion located at $0$ with charge $q$. Then
@@
 The script {\tt iontrap\_mwe1} does not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedoron with edge length $\rho > 0$.
-The scripts {\tt iontrap\_mwe2} and {\tt iontrap\_mwe4} replace $f$ with
+The scripts {\tt iontrap\_mwe2} and {\tt iontrap\_mwe4} replace $\bm{f}$ with
 \begin{equation}
-  f_k(\bm{r}) = f(\bm{r}) g_k(r), \quad k \in \{2, 4\{
+  f_k(\bm{r}) = f(\bm{r}) g_k(r), \quad k \in \{2, 4\}
 \end{equation}
@@
 The function $g_4$ is of class $C^\infty$ and satifies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.
-It is clear that changing the force-fields must impact the simulation, but can we actually estimate the discretization error for the kinetic energy?
+It is clear that changing the force-fields must impact the simulation, but can we estimate the discretization error and quantify this effect?
 Figure \ref{fig:iontrap_mwe1}, \ref{fig:iontrap_mwe2}, \ref{fig:iontrap_mwe4} show the evolution of Richardson's fraction for the kinetic energy at the end of each simulation.
@@
 When the force-fields are not perturbed, Figure~\ref{fig:iontrap_mwe1}, the experiments supports the conjecture that the exists an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate the dynamical system.
-
 \begin{figure}[h!]
@@
   \includegraphics[width=\linewidth]{iontrap_mwe2.pdf}
-  \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$.
-The electrostatic force-fields are zero outside a small sphere and exhibit jump discontinuities.}
+  \caption{The evolution of $F_h$ for the total kinetic energy of a system of ions using integrators of order $p$. The fields are zero outside a small sphere and discontinuous.}
   \label{fig:iontrap_mwe2}
 \end{figure}
-
-When the force-fields are truncated and jump-discontinuities are introduced, Figure~\ref{fig:iontrap_mwe2}, there is no evidence to suggest that there exists an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
-We cannot assert that the computational error is irrelevant and without an estimate of the discretization error we cannot quantify the impact of changing the force-field.
-
+When the force-fields are truncated and discontinuities are introduced into the simulation, see Figure~\ref{fig:iontrap_mwe2}, there is no evidence to suggest that there exists an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
 \begin{figure}
@@
   \includegraphics[width=\linewidth]{iontrap_mwe4.pdf}
-  \caption{The evolution of Richardson's fraction for the total kinetic energy of a system of $m=4$ ions using Runge-Kutta methods of order $p \in \{1,2,3,4\}$.
-The electrostatic force-fields are zero outside a small sphere, but the transition to zero is done smoothly.}
+  \caption{The evolution of $F_h$ for the total kinetic energy of a system of ions using integrators of order $p$. The fields are zero outside a small sphere, but smooth.}
   \label{fig:iontrap_mwe4}
 \end{figure}
-
-When the force-field vanish outside a small ball, but retain their smoothness, Figure~\ref{fig:iontrap_mwe4}, we retain the ability to judge when Richardson's error estimates are reliable.
+When the force-fields vanish outside a small ball but are smooth, see Figure~\ref{fig:iontrap_mwe4}, we retain the ability to judge when Richardson's error estimates are reliable and we can quantify the effect of changing the force-field.
 
 \section{Conclusion}
+
+A central task in the field of computational science is to fit a system of DAEs to the results of a physical experiment.
+It is often possible to assert when the computational errors are irrelevant and estimate the discretization error using Richardson extrapolation.
+This classical technique hinges on the existence of an asymptotic error expansion.
+However, if the functions that describe our problem are not sufficient differentiable or if the computational error is not sufficiently small, then we cannot estimate the discretization error and we lose the ability to evaluate our model.
+In the absence of further analysis, the best strategy is therefore to use functions that are smooth and solve all central equations as accurately as possible.
 

### 2024-05-13 report/src/draft.tex
Added past support statement for Lorien. Some figures have been reduced minimally in size. Began the process of eliminating mispellings, misprints etc.

@@
 \begin{abstract} We consider the problem of estimating the error when solving a system of differential algebraic equations.
-Richardson extrapolation is a classical technique that can be used to judge when rounding errors are irrelevant and estimate the discretization error.
-We have simulated molecular dynamics with constraints using the GROMACS library and found that the output is not always ameable to Richardson extrapolation.
+Richardson extrapolation is a classical technique that can be used to judge when computational errors are irrelevant and estimate the discretization error.
+We have simulated molecular dynamics with constraints using the GROMACS library and found that the output is not always amenable to Richardson extrapolation.
 We derive and illustrate Richardson extrapolation using a variety of numerical experiments.
@@
 Can we conclude that $T_0 - T(\bm{f})$ is small?
-The triangle inequality provided the following bound:
+The triangle inequality delivers the following bound:
 \begin{equation}
@@
 
-Richardson extrapolation is a classical technique that is widely used in computational science and engieering applications \cite{roache1998}.
+Richardson extrapolation is a classical technique that is widely used in computational science and engineering applications \cite{roache1998}.
 It can be used to estimate size of discretization errors or improve the accuracy of an existing solution \cite{zahari2018} and it has applications in event location \cite{mannshardt1978one} as well.
@@
 In this paper we derive and illustrate the use of Richardson extrapolation using a variety of numerical experiments.
-We demonstrate that the output of the GROMACS \cite{gromacs2005} library for molecular dynamics is not always ameable to Richardson extrapolation. 
-We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation
+We demonstrate that the output of the GROMACS \cite{gromacs2005} library for molecular dynamics is not always amenable to Richardson extrapolation. 
+We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation.
 Our data and software is freely available from our GitHub \cite{lorien2024ppam} along with every script and function needed to replicate every number, table and figure from scratch.
@@
 
-Consider the problem of approximating a target value $T$ using a numerical method $A = A_h$ that depends on a single real parameter $h$.
+Consider the problem of approximating a target value $T$ using a method $A = A_h$ that depends on a single real parameter $h$.
 We shall assume there exists nonzero real constants $\alpha$ and $\beta$ and real exponents
@@
 We say that the error $E_h$ satisfies an asymptotic error expansion.
-Frequently, the exponents $(p,q,r)$ are all integers, but since we shall allow exponents that are not integers, we insist that $h$ is strictly positive.
+Frequently, the exponents $(p,q,r)$ are all integers, but since we shall encounter exponents that are not integers, we insist that $h$ is strictly positive.
 
@@
   \begin{equation} \label{equ:Dh:1}
-    A_h - A_{2h} = (2^p - 1) \alpha h^p + (2^q - 1) h^q + g(2h) - g(h).
+    A_h - A_{2h} = (2^p - 1) \alpha h^p + (2^q - 1) \beta h^q + g(2h) - g(h).
   \end{equation}
@@
   \begin{equation}
-    E_h = R_h + \left(1 - \frac{2^q-1}{2^p-1} \right) \beta h^q + O(h^r)
+    E_h = R_h + \left(1 - \frac{2^q-1}{2^p-1} \right) \beta h^q + O(h^r).
   \end{equation}
@@
 \end{equation}
-using the composite trazoidal rule $A_h$ given by
+using the composite trapezoidal rule $A_h$ given by
 \begin{equation}
@@
 Let $f : [0,1] \rightarrow \mathbb{R}$ be given by $f(x) = \sqrt{x}$ and let $T$ be given by equation \eqref{equ:integral}.
-Then $T = \frac{2}{3}$. Since $f$ is not differentible at $x=0$ we have no guarantee that there exists an asymptotic error expansion of the form given by equation \eqref{equ:trapezoidal-rule:aex}.
+Then $T = \frac{2}{3}$. Since $f$ is not differentiable at $x=0$ we have no guarantee that there exists an asymptotic error expansion of the form given by equation \eqref{equ:trapezoidal-rule:aex}.
 The script {\tt rint\_mwe2} computes $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,25\}$ and generates Figures \ref{fig:rint_mwe2a} and \ref{fig:rint_mwe2b}.
@@
 Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T-\hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe2b}.
-We observe that computed value of Richardson's error estimate is a good approxmation of the error $T - \hat{A}_h$.
+We observe that computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_h$.
 In fact, the corresponding relative error decreases when $k$ increases and we remain inside the asymptotic region.
@@
 The script {\tt maxrange\_rk1} models a shell as a point particle moving in a plane subject to Earth's standard gravity and the international standard atmospheric model.
-This model is suitable only for short range artillery, as it does not account for the rotation of the Earth or the oriention and spin of the projectile.
+This model is suitable only for short range artillery, as it does not account for the rotation of the Earth or the orientation and spin of the projectile.
 Each trajectory is integrated using Euler's explicit method ({\tt 'rk1'}) and all but the final step has the same size $h$.
@@
 This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step.
-We observe that for each drag coefficient, $k=12$ is still inside the asymtotic range and we have no reason to doubt the magnitude of the error estimate.
+We observe that for each drag coefficient, $k=12$ is still inside the asymptotic range and we have no reason to doubt the magnitude of the error estimate.
 We conclude that the best model for the D-20 howitzer is in fact the G7 shell.
@@
 
-\subsection{An unsuccesful application of the theory}
+\subsection{An unsuccessful application of the theory}
 
@@
 
-\section{The difference between succes and failure}
+\section{The difference between success and failure}
 
-In this section we identify two conditions that are not always satisfied by the GROMACS library and we demonstrate that they are necessary for the succesful application of Richardson's techniques.
+In this section we identify two conditions that are not always satisfied by the GROMACS library and we demonstrate that they are necessary for the successful application of Richardson's techniques.
 
@@
 The friction drains the energy and ensures that the ions eventually come to rest in a stable configuration.
-Let $\bm{f}$ denote the forcefield generated by an ion located at $0$ with charge $q$. Then
+Let $\bm{f}$ denote the force-field generated by an ion located at $0$ with charge $q$. Then
 \begin{equation}
@@
 where $c > 0$ is a suitable constant.
-The script {\tt iontrap\_mwe1} does not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedoron with edge length $\rho > 0$.
+The script {\tt iontrap\_mwe1} does not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho > 0$.
 The scripts {\tt iontrap\_mwe2} and {\tt iontrap\_mwe4} replace $\bm{f}$ with
@@
 where $g_k$ is a switching function that assumes values in $[0,1]$.
-The function $g_2$ has a jump discontinuity and satisfies $g_1(r) = 1$ for $r < \rho$ and $g_1(r) = 0$ for $r \ge \rho$.
-The function $g_4$ is of class $C^\infty$ and satifies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.
-It is clear that changing the force-fields must impact the simulation, but can we estimate the discretization error and quantify this effect?
+The function $g_2$ has a jump discontinuity and satisfies $g_1(r) = 1$ for $r < 0.5 \rho$ and $g_1(r) = 0$ for $r \ge 0.5 \rho$.
+The function $g_4$ is of class $C^\infty$ and satisfies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.
+It is clear that changing the force-fields impacts the motion, but can we estimate the discretization error and quantify this effect?
 Figure \ref{fig:iontrap_mwe1}, \ref{fig:iontrap_mwe2}, \ref{fig:iontrap_mwe4} show the evolution of Richardson's fraction for the kinetic energy at the end of each simulation.
 
-When the force-fields are not perturbed, Figure~\ref{fig:iontrap_mwe1}, the experiments supports the conjecture that the exists an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate the dynamical system.
+When the force-fields are not perturbed, Figure~\ref{fig:iontrap_mwe1}, the experiments supports the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
 \begin{figure}[h!]
@@
 The first author is supported by eSSENCE, a collaborative e-Science programme funded by the Swedish Research Council within the framework of the strategic research areas designated by the Swedish Government.
-
 % @Lorien: update these lines with current information.
-% This work has been partially supported by the Spanish Ministry of Science and Innovation (contract PID2019-107255GB-C21/AEI/10.13039/501100011033), by the Generalitat de Catalunya (contract 2017-SGR-1328), and by Lenovo-BSC Contract-Framework Contract (2020).
+This work has been partially supported by the Spanish Ministry of Science and Innovation (contract PID2019-107255GB-C21/AEI/10.13039/501100011033), by the Generalitat de Catalunya (contract 2017-SGR-1328), and by Lenovo-BSC Contract-Framework Contract (2020).
 

### 2024-05-13 report/experiments/gromacs/src/gromacs_figures.m

@@ -1,4 +1,7 @@
-% Generate figures from GROMACS experiments
+% Generate figures from GROMACS experiments by Lorien Lopez-Villellas
+
+% PROGRAMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  Spring 2024  Initial programming and testing
 
 % Set the path to figures
 figpath='../experiments/gromacs/fig/';

### 2024-05-13 report/experiments/gromacs/src/gromacs_plot.m

@@ -1,5 +1,7 @@
 function [fig, ax]=gromacs_plot(field, tol, n, ekin, epot)
 
+
+
 fig=figure(); fig.Position=[100 800 1800 600];
 
 subplot(1,2,1); 
@@ -19,18 +21,9 @@ ylabel('Potential energy (kJ/mol)');
 % Set the fontsize for the axes
 ax=fig.CurrentAxes; ax.FontSize=14; ax.LineWidth=2;
 
-% subplot(1,3,3); 
-% plt3=plot(n,etol); grid;
-% xlabel('number of steps');
-% ylabel('Total energy (kJ/mol)');
-% 
-% % Set the fontsize for the axes
-% ax=fig.CurrentAxes; ax.FontSize=20; ax.LineWidth=2;
-
 % Set the linewidth for the plot
 plt1.LineWidth=2;
 plt2.LineWidth=2;
-% plt3.LineWidth=2;
 
 % Set a common title
 tit=sgtitle(['Forcefield = ' field '     constraint tolerance = ' num2str(tol,'%.2e')]);

### 2024-05-13 report/experiments/integration/src/rint_mwe1.m

@@ -1,13 +1,19 @@
-% Apply Richardson's techniques to integrating a smooth function
+% Numerical integration of a function that is everywhere smooth
 
 % PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
-%   2024-04-09  Adapted from 5dv231ht23
+%   2024-04-09  Adapted from 5dv231ht23 taught by CCKM at UmU
+
+% Clear the work space
+clear;
 
 % Interval
 a=0; b=1; 
 
 % Function
 f=@(x)exp(x); 
+
+% Antiderivative
+F=@(x)exp(x)
 
 % Integration rule
 rule=@(y,a,b,N)trapezoid(y,a,b,N); 
@@ -19,7 +25,7 @@ p=2;
 kmax=20; 
 
 % True value of the integral 
-val=exp(1)-1;
+val=F(b)-F(a);
 
 % Apply Richardson's techniques
 data=rint(f,a,b,rule,p,kmax,val);
@@ -46,7 +52,8 @@ axa=figa.CurrentAxes; axa.FontSize=20; % axa.GridLineWidth=2;
 % Set the linewidth for the plot
 plta.LineWidth=2;
 
-% figa.Position=[900 800 700 400]
+% Set the position and size
+figa.Position=[900 800 560 400]
 
 % Save the figure
 fname='rint_mwe1a.eps'; saveas(figa,strcat(mypath,fname));
@@ -75,7 +82,8 @@ axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
 % Set the linewidth for the plot
 pltb1.LineWidth=4; pltb2.LineWidth=4;
 
-% figb.Position=[900 800 700 400]
+% Set the position and size
+figb.Position=[900 800 560 400];
 
 % Save the figure
 fname='rint_mwe1b.eps';

### 2024-05-13 report/experiments/integration/src/rint_mwe2.m

@@ -1,7 +1,10 @@
-% Apply Richardson's techniques to integrating a smooth function
+% Numerical integration of a function that is smooth in all but one point.
 
 % PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
-%   2024-04-09  Adapted from 5dv231ht23
+%   2024-04-09  Adapted from 5dv231ht23 taught by CCKM at UmU
+
+% Clear the work space
+clear;
 
 % Interval
 a=0; b=1; 
@@ -30,7 +33,7 @@ data=rint(f,a,b,rule,p,kmax,val);
 % Obtain table parameters
 tp=table_param('rdif',data);
 
-% Print the information nicely
+% Print the information nicely on the screen
 print_table(data,tp);
 
 % Define path to figures
@@ -49,9 +52,11 @@ axa=figa.CurrentAxes; axa.FontSize=20; % axa.GridLineWidth=2;
 % Set the linewidth for the plot
 plta.LineWidth=2;
 
+% Set the position and size
+figa.Position=[900 800 560 400];
+
 % Save the figure
-fname='rint_mwe2a.eps';
-saveas(figa,strcat(mypath,fname));
+fname='rint_mwe2a.eps'; saveas(figa,strcat(mypath,fname));
 
 % ////////////////////////////////////////
 % Generate Figure rint_mwe2b
@@ -76,6 +81,9 @@ axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
 
 % Set the linewidth for the plot
 pltb1.LineWidth=4; pltb2.LineWidth=4;
+
+% Set the position and size
+figb.Position=[900 800 560 400];
 
 % Save the figure
 fname='rint_mwe2b.eps';

### 2024-05-13 report/experiments/iontrap/src/iontrap_mwe1.m

@@ -3,6 +3,9 @@
 % The force-fields are smooth and have infinite range
 % 
 
+% PROGRAMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  Spring 2024  Initial programming and testing
+
 % Clear the work space
 clear;
 
@@ -176,7 +179,7 @@ for j=1:nm
 end
 
 % Adjust graphics
-fig.Position=[300 800 1200 400];
+fig.Position=[300 800 1200 360];
 
 % Save the graphics
 saveas(fig,fname);

### 2024-05-13 report/experiments/iontrap/src/iontrap_mwe2.m

@@ -3,6 +3,9 @@
 % The electrostatic forcefields have a jump discontinuity.
 % They are truncated at a finite distance.
 
+% PROGRAMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  Spring 2024  Initial programming and testing
+
 % Clear the work space
 clear;
 
@@ -176,7 +179,7 @@ for j=1:nm
 end
 
 % Adjust graphics
-fig.Position=[300 800 1200 400];
+fig.Position=[300 800 1200 360];
 
 % Save the graphics
 saveas(fig,fname);

### 2024-05-13 report/experiments/iontrap/src/iontrap_mwe3.m

@@ -1,10 +1,24 @@
 % Ion trap with dampning
 % Runge-Kutta methods
-% The forcefield transitions smoothly to zero 
+% The force-fields transitions smoothly to zero 
+%
 
-% Set the central file name
-datpath='../experiments/iontrap/dat/';
-fname=strcat(datpath,"iontrap_mwe3.mat");
+% PROGRAMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  Spring 2024  Initial programming and testing
+
+% Clear the work space
+clear;
+
+% Set the main path
+mpath='../experiments/iontrap/';
+
+% Set the data path
+dpath=strcat(mpath,'dat/');
+dname=strcat(dpath,'iontrap_mwe3.mat');
+
+% Set the figure path
+fpath=strcat(mpath,'fig/');
+fname=strcat(fpath,'iontrap_mwe3.eps');
 
 % Check if raw data exists
 if ~isfile(fname)

### 2024-05-13 report/experiments/iontrap/src/iontrap_mwe4.m

@@ -3,6 +3,9 @@
 % The forcefield transitions smoothly to zero 
 % 
 
+% PROGRAMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
+%  Spring 2024  Initial programming and testing
+
 % Clear the work space
 clear;
 
@@ -176,7 +179,7 @@ for j=1:nm
 end
 
 % Adjust graphics
-fig.Position=[300 800 1200 400];
+fig.Position=[300 800 1200 360];
 
 % Save the graphics
 saveas(fig,fname);

### 2024-05-13 report/experiments/maxrange/src/maxrange_rk1.m

@@ -1,9 +1,19 @@
-% Compute the range of shells of different types
+% Compute the maximum range of the D-20 using different shells and RK1
+%
+% To generate the figures that fit in the manuscript leave ppam2024=true.
+% If you want all possible figures, set ppam2024=false
+
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.um.se
+%   Spring 2024   Initial programming and testing
 
 % Clear the work space
 clear;
 
-% Load the gun
+% Logical switch
+ppam2024=true;
+
+% Define the howitzer, atmosphere, gravity
 d20;
 
 % Load all shells
@@ -102,7 +112,10 @@ end
 % Allocate space for the data
 data=zeros(kmax,4,numshell,numtol);
 
-if (0==1)
+% Height of a single row of figures in pixels
+height=290;
+
+if (ppam2024==false)
 
     % Process all data and generate plots
     for t=1:numtol
@@ -111,8 +124,7 @@ if (0==1)
         % Plan where to put the figure
         q=floor((t-1)/4); r=(t-1)-4*q;
         % Position on screen
-        % fig.Position=[10+r*300 650-q*600 840 640];
-        fig.Position=[10+r*300 650-q*600 840 320];
+        fig.Position=[10+r*300 650-q*600 840 2*height];
         % Loop over the shell types
         for j=1:6
             % Apply Richardson extrapolation
@@ -146,7 +158,7 @@ else
         % Plan where to put the figure
         q=floor((t-1)/4); r=(t-1)-4*q;
         % Position on screen
-        fig.Position=[10+r*300 650-q*600 840 320];
+        fig.Position=[10+r*300 650-q*600 840 height];
         % Loop over the shells
         for j=1:6
             % Apply Richardson extrapolation

### 2024-05-13 report/experiments/maxrange/src/maxrange_rk2.m

@@ -1,9 +1,19 @@
-% Compute the range of shells of different types
+% Compute the maximum range of the D-20 using different shells and RK2
+%
+% To generate the figures that fit in the manuscript leave ppam2024=true.
+% If you want all possible figures, set ppam2024=false
+
+
+% PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.um.se
+%   Spring 2024   Initial programming and testing
 
 % Clear the work space
 clear;
 
-% Load the gun
+% Logical switch
+ppam2024=true;
+
+% Define the howitzer, atmosphere, gravity
 d20;
 
 % Load all shells
@@ -102,8 +112,10 @@ end
 % Allocate space for the data
 data=zeros(kmax,4,numshell,numtol);
 
+% Height of a single row of figures in pixels
+height=290;
 
-if (0==1)
+if (ppam2024==false)
     
     % Process all data and generate the plots
     for t=1:numtol
@@ -112,7 +124,7 @@ if (0==1)
         % Plan where to put the figure
         q=floor((t-1)/4); r=(t-1)-4*q;
         % Position on screen
-        fig.Position=[10+r*300 650-q*600 840 640];
+        fig.Position=[10+r*300 650-q*600 840 2*height];
         % Loop over the shell types
         for j=1:numshell
             % Apply Richardson extrapolation
@@ -146,7 +158,7 @@ else
         % Plan where to put the figure
         q=floor((t-1)/4); r=(t-1)-4*q;
         % Position on screen
-        fig.Position=[10+r*300 650-q*600 840 320];
+        fig.Position=[10+r*300 650-q*600 840 height];
         % Loop over G5, G6 and G7
         for j=1:6
             % Apply Richardson extrapolation
@@ -174,9 +186,6 @@ else
     end
     
 end
-
-
-
 
 % Obtain table parameters
 tp=table_param('rdif',data(:,:,1,numtol));

### 2024-05-13 report/src/draft.tex
Fixed more misprints. Fiddled with location of figures

@@
 In practice, the computed value $\hat{A}_h$ is different from the exact value $A_h$.
-However, it is often possible to assert that $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the error $T - \hat{A}_h$.
+However, it is often possible to assert that the computational error $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the error $T - \hat{A}_h$.
 We shall now demonstrate the procedure.
@@
 Let $f : [0, 1] \rightarrow \mathbb{R}$ be given by $f(x) = e^x$ and $T$ be given by equation \eqref{equ:integral}.
-The script {\tt rint\_mwe1} computes the composite trapezoidal sum $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,20\}$ and generates \ref{fig:rint_mwe1a} and \ref{fig:rint_mwe1b}.
+The script {\tt rint\_mwe1} computes the composite trapezoidal sum $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,19\}$ and generates \ref{fig:rint_mwe1a} and \ref{fig:rint_mwe1b}.
 The raw data shows that $\hat{A}_{h_k}$ approaches $4 = 2^2$ as $k$ increases and $k \in \{2,3,\dots,14\}$.
@@
 
-In this section we present the results of more elaborate experiments that highlight the practical limitations of Richardson's extrapolation.
+In this section we present the results of more elaborate experiments that highlight both the utility and the practical limitations of Richardson's extrapolation.
 
@@
 To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}.
-For each of the 6 different drag coefficients we find that evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$.
+For each of the 3 different drag coefficients shown we see that evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$.
 This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step.
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{oplsaaltol04.pdf}
+  \caption{The evolution of the kinetic and potential energy of a system as the number of time-steps used to cover $1$ ps of real time.} \label{fig:oplsaaltol04}
+\end{figure}
+
 We utilized GROMACS v2021 to conduct experiments on the behavior of hen egg white lysozyme submerged in water within a cubic simulation box, following Justin Lemkul's Lysozyme in Water GROMACS Tutorial \cite{lemkul2019from}.
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{oplsaaltol04.pdf}
-  \caption{The evolution of the kinetic and potential energy of a system as the number of time-steps used to cover $1$ ps of real time.} \label{fig:oplsaaltol04}
-\end{figure}
-
 Secondly, the total energy exhibits grows linearly with the number of time steps. This is not surprising as we expect the rounding error to grown with the number of operations.
@@
 Why is this?
-If we where in the asymptotic range, then the error $T - \hat{A}_h$ would very nearly obey a power law and $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some power $p>0$.
+If we where in the asymptotic range, then $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some $\alpha \not = 0$ and a $p>0$.
 In particular, the $\hat{A}_h$  would behave in be \emph{monotone} manner and the tiny oscillations that we have recorded would not be present.

### 2024-05-13 report/src/draft.tex
Fixed numerous misspellings and missing words

@@
 We identify two necessary conditions that are not always satisfied by the GROMACS library.
-\keywords{error estimation, Richardson extrapolation, multi-body dynamics, GROMACS}
+\keywords{error estimation, Richardson extrapolation, numerical integration, external ballistics, multi-body dynamics, GROMACS}
 \end{abstract}
@@
 Richardson extrapolation is a classical technique that is widely used in computational science and engineering applications \cite{roache1998}.
-It can be used to estimate size of discretization errors or improve the accuracy of an existing solution \cite{zahari2018} and it has applications in event location \cite{mannshardt1978one} as well.
+It can be used to estimate the size of discretization errors or improve the accuracy of an existing solution \cite{zahari2018} and it has applications in event location \cite{mannshardt1978one} as well.
 As we shall demonstrate, Richardson extrapolation can often be used to determine when computational errors are insignificant.
@@
 We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation.
-Our data and software is freely available from our GitHub \cite{lorien2024ppam} along with every script and function needed to replicate every number, table and figure from scratch.
+Our data and software is freely available from our GitHub \cite{lorien2024ppam} repository along with every script and function needed to replicate every number, table and figure from scratch.
 The names of our MATLAB functions are written with a typewriter font, e.g., \texttt{plot\_shells}.
@@
 \begin{equation} \label{equ:aex}
-  E_h  = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
+  E_h  = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+.
 \end{equation}
@@
 
-Our first task to estimate the error $E_h$ for a specific value of $h$. Richardson's error estimate $R_h$ is defined by the equation
+Our first task is to estimate the error $E_h$ for a specific value of $h$. Richardson's error estimate $R_h$ is defined by the equation
 \begin{equation}
- R_h =  \frac{A_h - A_{2h}}{2^p - 1}
+ R_h =  \frac{A_h - A_{2h}}{2^p - 1}.
 \end{equation}
@@
 The behavior of the function $h \rightarrow F_h$ is described by the following theorem.
-\begin{theorem} If $E_h$ satisfy equation \eqref{equ:aex} and if $(m,n)$ is given by
+\begin{theorem} If $E_h$ satisfies equation \eqref{equ:aex} and if $(m,n)$ is given by
   \begin{equation}
@@
    \begin{equation}
-      A_{2h} - A_{4h} = 2^p (2^p-1) \alpha h^p \Big[ 1 + 2^m \nu h^m + \phi(2h) \Big]
+      A_{2h} - A_{4h} = 2^p (2^p-1) \alpha h^p \Big[ 1 + 2^m \nu h^m + \phi(2h) \Big].
   \end{equation}
@@
   \begin{equation}
-    F_h = \frac{ A_{2h} - A_{4h}}{ A_h - A_{2h}} = 2^p \left[ \frac{ 1 + 2^m \nu h^m + \phi(h)}{ 1  + \nu h^m + \phi(2h)} \right]
+    F_h = \frac{ A_{2h} - A_{4h}}{ A_h - A_{2h}} = 2^p \left[ \frac{ 1 + 2^m \nu h^m + \phi(h)}{ 1  + \nu h^m + \phi(2h)} \right].
   \end{equation}
@@
   \begin{equation}
-    f(h) = 2^m \nu h^m + \phi(h), \quad g(h) = \nu h^m + \phi(2h)
+    f(h) = 2^m \nu h^m + \phi(h), \quad g(h) = \nu h^m + \phi(2h).
   \end{equation}
@@
 \end{equation}
-and the difference $m = q-p$ can be determined from the fact
+and the difference $m = q-p$ can be determined from the fact that
 \begin{equation} \label{equ:asymtotic-behavior-of-Fh}
@@
 This is the so-called asymptotic range, where the computed value $\hat{A}_h$  behaves in a manner that is indistinguishable from the exact value $A_h$.
-We conclude that the experiments supports the existence of an asymptotic error expansion with $(p,q) = (p,p+m) = (2,4)$.
+We conclude that the experiment supports the existence of an asymptotic error expansion with $(p,q) = (p,p+m) = (2,4)$.
 Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T - \hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe1b}.
-We observe that computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_{h_h}$.
+We observe that the computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_{h_h}$.
 In fact, the corresponding relative error decreases when $k$ increases as long as we also remain inside the asymptotic region.
@@
 Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T-\hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe2b}.
-We observe that computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_h$.
+We observe that the computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_h$.
 In fact, the corresponding relative error decreases when $k$ increases and we remain inside the asymptotic region.
@@
 
-In this section we shall demonstrate how the theory can be applied to analyze a nontrivial problem.
+% In this section we shall demonstrate how the theory can be applied to analyze a nontrivial problem.
 
@@
 The final step is adjusted to place the shell on the ground.
-The drag function are interpolated from tables using cubic spline interpolation.
+The drag functions are interpolated from tables using cubic spline interpolation.
 The function {\tt plot\_shells} will plot the drag coefficients for the different shells as a function of the Mach number.
@@
 These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
-Figure \ref{fig:maxrange_rk1_table_tol53} list the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
-In each case the error estimate suggests that the computed range is exact to number of figures shown.
+Figure \ref{fig:maxrange_rk1_table_tol53} lists the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
+In each case the error estimate suggests that the computed range is exact to the number of figures shown.
 In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km.
-It is however, a fallacy to conclude anything on the basis of this table alone.
+However, it is a fallacy to conclude anything on the basis of this table alone.
 In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable.
 To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}.
-For each of the 3 different drag coefficients shown we see that evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$.
-This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step.
+For each of the 3 different drag coefficients shown we see that the evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$.
+This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step\footnote{The remaining figures are similar and have been omitted to save space.}.
 We observe that for each drag coefficient, $k=12$ is still inside the asymptotic range and we have no reason to doubt the magnitude of the error estimate.
@@
 These figures display the mechanical energies of the molecule at the end of the simulation as a function of the total number $n$ of time steps used to cover the interval.
-The figures have several features of interest.
+The figures present several features of interest.
 
@@
 
-Secondly, the total energy exhibits grows linearly with the number of time steps. This is not surprising as we expect the rounding error to grown with the number of operations.
+Secondly, the total energy grows linearly with the number of time steps. This is not surprising as we expect the rounding error to grow with the number of operations.
 
@@
 Why is this?
-If we where in the asymptotic range, then $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some $\alpha \not = 0$ and a $p>0$.
-In particular, the $\hat{A}_h$  would behave in be \emph{monotone} manner and the tiny oscillations that we have recorded would not be present.
+If we where in the asymptotic range, then $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some $\alpha \not = 0$ and $p>0$.
+In particular, the value of $\hat{A}_h$ should behave in a \emph{monotone} manner and the tiny oscillations that we have recorded should not be present.
 
@@
 
-In this section we identify two conditions that are not always satisfied by the GROMACS library and we demonstrate that they are necessary for the successful application of Richardson's techniques.
+In this section we identify two conditions that are not always satisfied by the GROMACS library and we demonstrate that they are necessary for the successful application of Richardson extrapolation.
 
@@
 
-When computing the range of a howitzer it necessary to adjust the final time-step to place the shell on the ground with great accuracy.
+It is clear that the output of constrained MD simulation depends on the tolerance passed to the constraint solver.
+Similarly, when computing the range of a howitzer it is necessary to adjust the final time-step to place the shell on the ground with great accuracy.
 In the case of Euler's explicit method, the relevant equation is linear, but in general it is a nonlinear equation.
-The function {\tt maxrange\_rk1\_mwe1} and it is companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h\cdot\text{tol}$ using a wide range of tolerance {\tt tol}.
+The function {\tt maxrange\_rk1\_mwe1} and its companion {\tt maxrange\_rk2\_mwe2} both use the bisection method to compute the final time step with an error that is bounded by $h\cdot\text{tol}$ using a wide range of tolerance {\tt tol}.
 \begin{figure}
@@
 \end{figure}
-When the tolerance is large (Figure~\ref{fig:maxrange_rk2_tol11}) there is no evidence that an asymptotic error expansion exists and is no reason to trust Richardson's error estimate.
-When the tolerance is small (Figure~\ref{fig:maxrange_rk2_tol25}), there is evidence to suggest that asymptotic error expansion exists, but it is not trivial to determine an asymptotic range.
-Regardless, it is clear that if we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is solve all equations as accurately as the hardware will allow.
+When the tolerance is large, see Figure~\ref{fig:maxrange_rk2_tol11}, there is no evidence that an asymptotic error expansion exists and there is no reason to trust Richardson's error estimate.
+When the tolerance is small, see Figure~\ref{fig:maxrange_rk2_tol25}, the fact that $\hat{F}_h$ tends to $2^p$ suggests that an asymptotic error expansion exists, but it is not trivial to determine an asymptotic range.
+Regardless, it is clear that if we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is to solve all equations as accurately as the hardware will allow.
 
@@
 This can be done by setting force-fields to zero outside of a sufficiently large ball.
-There is more than way to accomplish this effect and the documentation for GROMACS 2021 discusses its use of force-fields that are not of class $C^{\infty}$.
+There is more than one way to achieve this and the documentation for GROMACS 2021 discusses its use of force-fields that are not of class $C^{\infty}$.
 
@@
 \begin{equation}
-  f_k(\bm{r}) = f(\bm{r}) g_k(r), \quad k \in \{2, 4\}
+  \bm{f}_k(\bm{r}) = \bm{f}(\bm{r}) g_k(r), \quad k \in \{2, 4\}
 \end{equation}
@@
 
-When the force-fields are not perturbed, Figure~\ref{fig:iontrap_mwe1}, the experiments supports the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
+When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, the experiments supports the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
 \begin{figure}[h!]
@@
 \end{figure}
-When the force-fields are truncated and discontinuities are introduced into the simulation, see Figure~\ref{fig:iontrap_mwe2}, there is no evidence to suggest that there exists an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
+When the force-fields are truncated and discontinuities are introduced into the simulation, see Figure~\ref{fig:iontrap_mwe2}, there is no evidence to support the existence of an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
 \begin{figure}
@@
 This classical technique hinges on the existence of an asymptotic error expansion.
-However, if the functions that describe our problem are not sufficient differentiable or if the computational error is not sufficiently small, then we cannot estimate the discretization error and we lose the ability to evaluate our model.
+However, if the functions that describe our problem are not sufficiently differentiable or if the computational error is not sufficiently small, then we cannot estimate the discretization error and we lose the ability to evaluate our model.
 In the absence of further analysis, the best strategy is therefore to use functions that are smooth and solve all central equations as accurately as possible.

### 2024-05-13 report/experiments/gromacs/src/gromacs_figures.m

@@ -1,8 +1,5 @@
 % Generate figures from GROMACS experiments by Lorien Lopez-Villellas
 
-% PROGRAMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
-%  Spring 2024  Initial programming and testing
-
 % Set the path to figures
 figpath='../experiments/gromacs/fig/';
 
@@ -21,7 +18,7 @@ for i=1:4
     aux=c{i}.data; field=c{i}.field; tol=c{i}.tol;
 
     % Rename the columns to reduce the chance of misprints
-    ekin=aux(:,3); epot=aux(:,4); % etol=ekin+epot;
+    ekin=aux(:,3); epot=aux(:,4); 
 
     % Call specialized plotting routine
     [fig, ax]=gromacs_plot(field, tol, n, ekin, epot);

### 2024-05-13 report/src/draft.tex
Minor addition to the introduction

@@
 In this paper we derive and illustrate the use of Richardson extrapolation using a variety of numerical experiments.
-We demonstrate that the output of the GROMACS \cite{gromacs2005} library for molecular dynamics is not always amenable to Richardson extrapolation. 
+GROMACS is a state-of-the-art library for molecular dynamics that is widely used in academia \cite{gromacs2005}.
+We demonstrate that the output of the GROMACS is not always amenable to Richardson extrapolation. 
 We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation.

### 2024-05-13 report/src/draft.tex
Added acknowlegdement of Jesus, Pablo and Pablo

@@
 When the tolerance is large, see Figure~\ref{fig:maxrange_rk2_tol11}, there is no evidence that an asymptotic error expansion exists and there is no reason to trust Richardson's error estimate.
-When the tolerance is small, see Figure~\ref{fig:maxrange_rk2_tol25}, the fact that $\hat{F}_h$ tends to $2^p$ suggests that an asymptotic error expansion exists, but it is not trivial to determine an asymptotic range.
+When the tolerance is small, see Figure~\ref{fig:maxrange_rk2_tol25}, the fact that $\hat{F}_h$ approaches $2^p$ suggests that an asymptotic error expansion exists, but it is not trivial to determine an asymptotic range.
 Regardless, it is clear that if we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is to solve all equations as accurately as the hardware will allow.
@@
 
+The authors would like to thank Jesús Alastruey-Benedé, Pablo Ibáñez and Pablo García-Risueño for stimulating discussions on the subject matter.
 The first author is supported by eSSENCE, a collaborative e-Science programme funded by the Swedish Research Council within the framework of the strategic research areas designated by the Swedish Government.

### 2024-05-14 report/scripts/clean.sh
Generated all figures from scratch. Removed irrelevant comments from manuscript

@@ -1,0 +1,3 @@
+cd src
+rm *.aux *.bbl *.blg *.log *.out *~
+cd ..

### 2024-05-14 report/scripts/compile.sh

@@ -1,4 +1,6 @@
 cd src
 pdflatex draft.tex
+bibtex draft.aux
+pdflatex draft.tex
 pdflatex draft.tex
 cd ..

### 2024-05-14 report/scripts/gatherfig.sh

@@ -1,7 +1,6 @@
-find experiments -name \*.png -exec cp {} fig \;
 find experiments -name \*.eps -exec cp {} fig \;
 
-# Crop white space from MATLAB eps files
+# Crop white space from .eps files and generate .pdf files
 cd fig
 ps2pdf -DEPSCrop rint_mwe1a.eps rint_mwe1a.pdf
 ps2pdf -DEPSCrop rint_mwe1b.eps rint_mwe1b.pdf

### 2024-05-14 report/src/draft.tex
Generated all figures from scratch. Removed irrelevant comments from manuscript
Fixed misprints regarding the functions g2 and g4 used in the iontrap experiment
Affiliation updated

@@
 % to display URLs in blue roman font according to Springer's eBook style:
-%\usepackage{color}
-%\renewcommand\UrlFont{\color{blue}\rmfamily}
+\usepackage{color}
+\renewcommand\UrlFont{\color{blue}\rmfamily}
 %
-\usepackage{color}
-\newcommand{\todo}[1]{{\color{blue}TODO : {#1}}}
-
-% Set control to 1 to include all proofs.
-% Set control to 0 to shrink the paper to 12 pages
-\def\control{1}
 
@@
 %\linenumbers
-% \renewcommand\linenumberfont{\normalfont\scriptsize}
+%\renewcommand\linenumberfont{\normalfont\scriptsize}
 
@@
   Department of Computing Science, Ume{\aa} University, 90187 Ume{\aa}, Sweden \email{spock@cs.umu.se} \\
-  \and Barcelona Supercomputing Center, Barcelona, Spain, \\
-  \email{lorien.lopez@bsc.es} \\
+  \and Departamento de Inform\'atica e Ingenier\'ia de Sistemas / Arag\'on Institute for Engineering Research (I3A), Universidad de Zaragoza, Zaragoza, Spain \\
+  \email{lorien.lopez@unizar.es} \\
  %  \and Independent scholar \\
@@
 \begin{equation}
-  \bm{f}(\bm{r}) = c \frac{q}{\|\bm{r}\|^2} \frac{\bm{r}}{\|\bm{r}\|}
+  \bm{f}(\bm{r}) = c \frac{q}{r^2} \frac{\bm{r}}{r}, \quad r = \|\bm{r}\|_2,
 \end{equation}
@@
 where $g_k$ is a switching function that assumes values in $[0,1]$.
-The function $g_2$ has a jump discontinuity and satisfies $g_1(r) = 1$ for $r < 0.5 \rho$ and $g_1(r) = 0$ for $r \ge 0.5 \rho$.
+The function $g_2$ has a jump discontinuity and satisfies $g_2(r) = 1$ for $r < 0.5 \rho$ and $g_2(r) = 0$ for $r \ge 0.5 \rho$.
 The function $g_4$ is of class $C^\infty$ and satisfies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.

### 2024-05-14 report/src/draft.tex
Funding updated.

@@
 The first author is supported by eSSENCE, a collaborative e-Science programme funded by the Swedish Research Council within the framework of the strategic research areas designated by the Swedish Government.
-% @Lorien: update these lines with current information.
-This work has been partially supported by the Spanish Ministry of Science and Innovation (contract PID2019-107255GB-C21/AEI/10.13039/501100011033), by the Generalitat de Catalunya (contract 2017-SGR-1328), and by Lenovo-BSC Contract-Framework Contract (2020).
+This work has been partially supported by the Spanish Ministry of Science and Innovation MCIN/AEI/10.13039/501100011033 (grant PID2022-136454NB-C22), and by Government of Aragon (T58_23R research group).
 

### 2024-05-14 report/src/draft.tex
Fixed typos

@@
 
-Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e, $\tau = 10^{-4}$.
+Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e., $\tau = 10^{-4}$.
 The amplitude of the oscillations is reduced when $\tau = 10^{-12}$.
@@
 Why is this?
-If we where in the asymptotic range, then $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some $\alpha \not = 0$ and $p>0$.
+If we were in the asymptotic range, then $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some $\alpha \not = 0$ and $p>0$.
 In particular, the value of $\hat{A}_h$ should behave in a \emph{monotone} manner and the tiny oscillations that we have recorded should not be present.
@@
 
-When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, the experiments supports the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
+When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, the experiments support the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
 \begin{figure}[h!]

### 2024-05-14 report/src/draft.tex
Address of GitHub repository updated

--- 2024-05-14 report/src/draft.tex
@@
   Department of Computing Science, Ume{\aa} University, 90187 Ume{\aa}, Sweden \email{spock@cs.umu.se} \\
-  \and Departamento de Inform\'atica e Ingenier\'ia de Sistemas / Arag\'on Institute for Engineering Research (I3A), Universidad de Zaragoza, Zaragoza, Spain \\
-  \email{lorien.lopez@unizar.es} \\
+  \and Barcelona Supercomputing Center, Barcelona, Spain, \\
+  \email{lorien.lopez@bsc.es} \\
  %  \and Independent scholar \\
@@
 We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation.
-Our data and software is freely available from our GitHub \cite{lorien2024ppam} repository along with every script and function needed to replicate every number, table and figure from scratch.
+Our data and software is freely available from our GitHub \cite{spockcc2024ppam} repository along with every script and function needed to replicate every number, table and figure from scratch.
 The names of our MATLAB functions are written with a typewriter font, e.g., \texttt{plot\_shells}.

### 2024-05-14 report/src/draft.tex
Typo

@@
 GROMACS is a state-of-the-art library for molecular dynamics that is widely used in academia \cite{gromacs2005}.
-We demonstrate that the output of the GROMACS is not always amenable to Richardson extrapolation. 
+We demonstrate that the output of GROMACS is not always amenable to Richardson extrapolation. 
 We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation.
@@
 The first author is supported by eSSENCE, a collaborative e-Science programme funded by the Swedish Research Council within the framework of the strategic research areas designated by the Swedish Government.
-This work has been partially supported by the Spanish Ministry of Science and Innovation MCIN/AEI/10.13039/501100011033 (grant PID2022-136454NB-C22), and by Government of Aragon (T58_23R research group).
+This work has been partially supported by the Spanish Ministry of Science and Innovation MCIN/AEI/10.13039/501100011033 (grant PID2022-136454NB-C22), and by Government of Aragon (T58\_23R research group).
 

### 2024-05-14 report/experiments/gromacs/src/gromacs_plot.m
Various adjustments to make the text fit

@@ -2,7 +2,7 @@ function [fig, ax]=gromacs_plot(field, tol, n, ekin, epot)
 
 
 
-fig=figure(); fig.Position=[100 800 1800 600];
+fig=figure(); fig.Position=[100 800 1800 560];
 
 subplot(1,2,1); 
 plt1=plot(n,ekin); grid; 

### 2024-05-14 report/src/draft.tex
Various adjustments to make the text fit
The manuscript is now at the page limit

@@
 The script {\tt maxrange\_rk1} models a shell as a point particle moving in a plane subject to Earth's standard gravity and the international standard atmospheric model.
-This model is suitable only for short range artillery, as it does not account for the rotation of the Earth or the orientation and spin of the projectile.
+%This model is suitable only for short range artillery, as it does not account for the rotation of the Earth or the orientation and spin of the projectile.
 Each trajectory is integrated using Euler's explicit method ({\tt 'rk1'}) and all but the final step has the same size $h$.
@@
 The function {\tt plot\_shells} will plot the drag coefficients for the different shells as a function of the Mach number.
-
 For each drag coefficient our target value $T$ is the maximum range of the shell as the elevation of the howitzer varies continuously from $0$ to $\frac{\pi}{2}$.
@@
 In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}.
-These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
-Figure \ref{fig:maxrange_rk1_table_tol53} lists the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
-In each case the error estimate suggests that the computed range is exact to the number of figures shown.
-In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km.
-However, it is a fallacy to conclude anything on the basis of this table alone.
-In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable.
-To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}.
-For each of the 3 different drag coefficients shown we see that the evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$.
-This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step\footnote{The remaining figures are similar and have been omitted to save space.}.
-We observe that for each drag coefficient, $k=12$ is still inside the asymptotic range and we have no reason to doubt the magnitude of the error estimate.
-We conclude that the best model for the D-20 howitzer is in fact the G7 shell.
-
 \begin{figure}
@@
 \end{figure}
+These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
+Figure \ref{fig:maxrange_rk1_table_tol53} lists the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
+In each case the error estimate suggests that the computed range is exact to the number of figures shown.
+In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km.
+However, it is a fallacy to conclude anything on the basis of this table alone.
+In each case, we need to assert that we are inside the asymptotic range and that the error estimates are reliable.
+To this end, we examine the evolution of Richardson's fraction for the maximum range of each shell, see Figure \ref{fig:maxrange_rk1_fraction_tol53}.
+For each of the 3 different drag coefficients shown we see that the evolution of Richardson's fraction supports an asymptotic error expansion with $(p,q) = (1,2)$.
+This result is consistent with the use of Euler's explicit method which is 1st order accurate in the time step\footnote{The remaining figures are similar and have been omitted to save space.}.
+We observe that for each drag coefficient, $k=12$ is still inside the asymptotic range and we have no reason to doubt the magnitude of the error estimate.
+We conclude that the best model for the D-20 howitzer is in fact the G7 shell.
 
@@
 \subsection{An unsuccessful application of the theory}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{oplsaaltol04.pdf}
-  \caption{The evolution of the kinetic and potential energy of a system as the number of time-steps used to cover $1$ ps of real time.} \label{fig:oplsaaltol04}
-\end{figure}
 
@@
 The function {\tt gromacs\_figures} will generate Figure~\ref{fig:oplsaaltol04} and Figure~\ref{fig:oplsaaltol12} for the OPLS-AA/L force-file and similar figures for CHARMM36.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{oplsaaltol04.pdf}
+  \caption{The evolution of the kinetic and potential energy of a system as the number of time-steps used to cover $1$ ps of real time.} \label{fig:oplsaaltol04}
+\end{figure}
+
 These figures display the mechanical energies of the molecule at the end of the simulation as a function of the total number $n$ of time steps used to cover the interval.
 The figures present several features of interest.
-
 Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e., $\tau = 10^{-4}$.
@@
 We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expectation for the computed approximation unless $\tau$ is very small.
-
 Secondly, the total energy grows linearly with the number of time steps. This is not surprising as we expect the rounding error to grow with the number of operations.
-
 Thirdly, if the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} well inside the asymptotic range.
@@
 \begin{equation}
-  \bm{f}(\bm{r}) = c \frac{q}{r^2} \frac{\bm{r}}{r}, \quad r = \|\bm{r}\|_2,
+  \bm{f}(\bm{r}) = c q \bm{r}/r^3, \quad r = \|\bm{r}\|_2,
 \end{equation}
@@
 
-When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, the experiments support the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
+When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, or when the perturbation is smooth, see Figure~\ref{fig:iontrap_mwe4}, then the experiments support the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.x1
 \begin{figure}[h!]
@@
 \end{figure}
-When the force-fields vanish outside a small ball but are smooth, see Figure~\ref{fig:iontrap_mwe4}, we retain the ability to judge when Richardson's error estimates are reliable and we can quantify the effect of changing the force-field.
 

### 2024-05-14 report/experiments/integration/src/rint_mwe1.m

@@ -83,7 +83,7 @@ axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
 pltb1.LineWidth=4; pltb2.LineWidth=4;
 
 % Set the position and size
-figb.Position=[900 800 560 400];
+figb.Position=[900 800 560 320];
 
 % Save the figure
 fname='rint_mwe1b.eps';

### 2024-05-14 report/experiments/integration/src/rint_mwe2.m

@@ -83,7 +83,7 @@ axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
 pltb1.LineWidth=4; pltb2.LineWidth=4;
 
 % Set the position and size
-figb.Position=[900 800 560 400];
+figb.Position=[900 800 560 320];
 
 % Save the figure
 fname='rint_mwe2b.eps';

### 2024-05-14 report/experiments/maxrange/src/maxrange_rk1.m

@@ -113,7 +113,7 @@ end
 data=zeros(kmax,4,numshell,numtol);
 
 % Height of a single row of figures in pixels
-height=290;
+height=280;
 
 if (ppam2024==false)
 

### 2024-05-14 report/experiments/maxrange/src/maxrange_rk2.m

@@ -113,7 +113,7 @@ end
 data=zeros(kmax,4,numshell,numtol);
 
 % Height of a single row of figures in pixels
-height=290;
+height=280;
 
 if (ppam2024==false)
     

### 2024-06-04 report/src/draft.tex
Made several improvements to the figures and corrected misprints. These changes were suggested by Pablo Garcia-Risueno

@@
 The function $\bm{G}$ is the Jacobian of the constraint function $\bm{g}$ and $\bm{\lambda}$ is a vector of Lagrange multipliers.
-In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm \cite{shake1977}
+In the field of molecular dynamics, the standard algorithm for this problem is the SHAKE algorithm \cite{shake1977}.
 It uses a pair of staggered grids with uniform step size $h$ and takes the form
@@
 \end{align}
-The constraint equation \eqref{equ:constraint} is usually a nonlinear equation with respect to the Lagrange multiplier $\bm{\lambda}_n$. 
+The constraint equation \eqref{equ:constraint} is usually a nonlinear equation with respect to the Lagrange multipliers $\bm{\lambda}_n$. 
 Now let $T \in \mathbb{R}$ denote any target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
@@
 Let $\hat{A}_h(\bm{f}, \tau, u)$ denote the value returned by our computer when solving the constraint equations with a relative error bounded by $\tau$ and using floating point arithmetic with unit roundoff $u$.
-Suppose that $T_0 - A_h(\bm{f}, \tau, u)$ is small.
+Suppose that $T_0 - \hat{A}_h(\bm{f}, \tau, u)$ is small.
 Can we conclude that $T_0 - T(\bm{f})$ is small?
@@
 \begin{equation}
-  |T_0 - T(\bm{f}) | \leq  |T_0 - A_h(\bm{f},\tau,u)| + |A_h(\bm{f}) - A_h(\bm{f},\tau,u)| + |T(\bm{f}) - A_h(\bm{f})|.
+  |T_0 - T(\bm{f}) | \leq  |T_0 - \hat{A}_h(\bm{f},\tau,u)| + |A_h(\bm{f}) - \hat{A}_h(\bm{f},\tau,u)| + |T(\bm{f}) - A_h(\bm{f})|.
 \end{equation}
-We conclude that $T_0 - T(\bm{f})$ is small, if the \emph{computational error} $A_h(\bm{f}) - A_h(\bm{f}, \tau, u)$ and the \emph{discretization error} $T(\bm{f}) - A_h(\bm{f})$ are both small.
+We conclude that $T_0 - T(\bm{f})$ is small, if the \emph{computational error} $A_h(\bm{f}) - \hat{A}_h(\bm{f}, \tau, u)$ and the \emph{discretization error} $T(\bm{f}) - A_h(\bm{f})$ are both small.
 If we cannot control these errors, then we cannot say with certainty that our model delivers a good approximation of the physical reality.
@@
 We identify two conditions that are not always satisfied by GROMACS and we demonstrate that each condition is necessary for the successful application of Richardson extrapolation.
-Our data and software is freely available from our GitHub \cite{spockcc2024ppam} repository along with every script and function needed to replicate every number, table and figure from scratch.
+Our data and software are freely available from our GitHub \cite{spockcc2024ppam} repository along with every script and function needed to replicate every number, table and figure from scratch.
 The names of our MATLAB functions are written with a typewriter font, e.g., \texttt{plot\_shells}.
@@
 Consider the problem of approximating a target value $T$ using a method $A = A_h$ that depends on a single real parameter $h$.
-We shall assume there exists nonzero real constants $\alpha$ and $\beta$ and real exponents
+We shall assume that there exists nonzero real constants $\alpha$ and $\beta$ and real exponents
 \begin{equation}
@@
 Since the target value $T$ is known, we can treat Richardson's error estimate as an approximation of the error $T - \hat{A}_h$ and compute the corresponding relative error, see Figure \ref{fig:rint_mwe1b}.
-We observe that the computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_{h_h}$.
+We observe that the computed value of Richardson's error estimate is a good approximation of the error $T - \hat{A}_{h_k}$.
 In fact, the corresponding relative error decreases when $k$ increases as long as we also remain inside the asymptotic region.
@@
 The script {\tt rint\_mwe2} computes $A_h$ using $h_k = 2^{-k}$ for $k \in \{0,1,\dots,25\}$ and generates Figures \ref{fig:rint_mwe2a} and \ref{fig:rint_mwe2b}.
-From the raw data it is clear that $p = 2$ cannot be true, but it is plausible that $p \approx \frac{3}{2}$.
+The raw data shows that $p = 2$ cannot be true, but it is plausible that $p \approx \frac{3}{2}$.
 Figure \ref{fig:rint_mwe1a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
@@
 
-\paragraph{Example: Identify the shells fired by a howitzer}
+\paragraph{Example: Identify the shells fired by a howitzer.}
 
@@
 In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}.
-\begin{figure}
+\begin{table}
+ \caption{The computed maximum range for 6 shells fired from a D-20 howitzer.}
+ \label{fig:maxrange_rk1_table_tol53} 
  \centering
+ \vspace{.25cm}
  \begin{tabular}{c|c|c}
@@
    G8      &     15914    &            0.1 \\
+   \hline
  \end{tabular}
-  \caption{The computed maximum range for 6 shells fired from a D-20 howitzer.}
-  \label{fig:maxrange_rk1_table_tol53} 
-\end{figure}
+\end{table}
 These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
@@
 In each case the error estimate suggests that the computed range is exact to the number of figures shown.
-In particular, we see that a G7 type shell achieves a maximum range of 17.4 km and all other shells have ranges that are less than 16.9 km.
+In particular, we see that a G7 type shell achieves a maximum range of 17.5 km and all other shells have ranges that are less than 16.9 km.
 However, it is a fallacy to conclude anything on the basis of this table alone.
@@
 
-These figures display the mechanical energies of the molecule at the end of the simulation as a function of the total number $n$ of time steps used to cover the interval.
+These figures display the total potential and kinetic energy at the end of the simulation as a function of the total number $n$ of time steps used to cover the interval.
 The figures present several features of interest.
@@
 In order to explore the importance of smoothness we have simulated the motion of a set of identical ions moving in a liquid.
-The ions repel each other electrostatically, but they are pulled towards the origo by independent and identical springs that obey Hooke's law.
+The ions repel each other electrostatically, but they are pulled towards the origin by independent and identical springs that obey Hooke's law.
 The friction between each ion and the liquid is proportional to its velocity.

### 2024-06-04 report/experiments/gromacs/src/gromacs_plot.m

@@ -1,8 +1,15 @@
 function [fig, ax]=gromacs_plot(field, tol, n, ekin, epot)
 
+% Figure height in pixels
+height=600;
 
+% Linewidth
+lw=2;
 
-fig=figure(); fig.Position=[100 800 1800 600];
+% Fontsize 16
+fs=32;
+
+fig=figure(); fig.Position=[100 800 1800 height];
 
 subplot(1,2,1); 
 plt1=plot(n,ekin); grid; 
@@ -10,7 +17,7 @@ xlabel('number of steps');
 ylabel('Kinetic energy (kJ/mol)'); 
 
 % Set the fontsize for the axes
-ax=fig.CurrentAxes; ax.FontSize=14; ax.LineWidth=2;
+ax=fig.CurrentAxes; ax.FontSize=fs; ax.LineWidth=lw;
 
 subplot(1,2,2); 
 plt2=plot(n,epot); grid;
@@ -19,12 +26,12 @@ xlabel('number of steps');
 ylabel('Potential energy (kJ/mol)');
 
 % Set the fontsize for the axes
-ax=fig.CurrentAxes; ax.FontSize=14; ax.LineWidth=2;
+ax=fig.CurrentAxes; ax.FontSize=fs; ax.LineWidth=lw;
 
 % Set the linewidth for the plot
-plt1.LineWidth=2;
-plt2.LineWidth=2;
+plt1.LineWidth=lw;
+plt2.LineWidth=lw;
 
 % Set a common title
 tit=sgtitle(['Forcefield = ' field '     constraint tolerance = ' num2str(tol,'%.2e')]);
-tit.FontSize=20;
+tit.FontSize=32;

### 2024-06-04 report/experiments/integration/src/rint_mwe1.m

@@ -3,8 +3,18 @@
 % PROGRAMMING by Carl Christian Kjelgaard Mikkelsen (spock@cs.umu.se)
 %   2024-04-09  Adapted from 5dv231ht23 taught by CCKM at UmU
 
+
 % Clear the work space
 clear;
+
+% Dimensions of figures in pixel
+height=360;
+
+% Line width
+lw=4;
+
+% Fontsize
+fs=16;
 
 % Interval
 a=0; b=1; 
@@ -47,13 +57,13 @@ mypath='../experiments/integration/fig/';
 figa=figure(); plta=plot_fraction(data,p);
 
 % Set the fontsize for the axes
-axa=figa.CurrentAxes; axa.FontSize=20; % axa.GridLineWidth=2;
+axa=figa.CurrentAxes; axa.FontSize=16; % axa.GridLineWidth=2;
 
 % Set the linewidth for the plot
-plta.LineWidth=2;
+plta.LineWidth=lw;
 
 % Set the position and size
-figa.Position=[900 800 560 400]
+figa.Position=[900 800 560 height];
 
 % Save the figure
 fname='rint_mwe1a.eps'; saveas(figa,strcat(mypath,fname));
@@ -75,15 +85,16 @@ pltb2=plot(k,log10(abs(rel)),'b:');
 axis([0 kmax -14 2]);
 grid; lgdb=legend('log_{10}(|E_h|)','log_{10}(|E_h-R_h|/|E_h|)', ...
     'Location','SouthWest');
+xlabel('k');
 
 % Set the fontsize for the axes
-axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
+axb=figb.CurrentAxes; axb.FontSize=16; % axb.GridLineWidth=2;
 
 % Set the linewidth for the plot
-pltb1.LineWidth=4; pltb2.LineWidth=4;
+pltb1.LineWidth=lw; pltb2.LineWidth=lw;
 
 % Set the position and size
-figb.Position=[900 800 560 320];
+figb.Position=[900 800 560 height];
 
 % Save the figure
 fname='rint_mwe1b.eps';

### 2024-06-04 report/experiments/integration/src/rint_mwe2.m

@@ -6,6 +6,15 @@
 % Clear the work space
 clear;
 
+% Dimensions of figures in pixel
+height=360;
+
+% Line width
+lw=4;
+
+% Fontsize
+fs=16;
+
 % Interval
 a=0; b=1; 
 
@@ -47,13 +56,13 @@ mypath='../experiments/integration/fig/';
 figa=figure(); plta=plot_fraction(data,p);
 
 % Set the fontsize for the axes
-axa=figa.CurrentAxes; axa.FontSize=20; % axa.GridLineWidth=2;
+axa=figa.CurrentAxes; axa.FontSize=fs; % axa.GridLineWidth=2;
 
 % Set the linewidth for the plot
-plta.LineWidth=2;
+plta.LineWidth=lw;
 
 % Set the position and size
-figa.Position=[900 800 560 400];
+figa.Position=[900 800 560 height];
 
 % Save the figure
 fname='rint_mwe2a.eps'; saveas(figa,strcat(mypath,fname));
@@ -77,13 +86,13 @@ grid; lgdb=legend('log_{10}(|E_h|)','log_{10}(|E_h-R_h|/|E_h|)', ...
     'Location','SouthWest');
 
 % Set the fontsize for the axes
-axb=figb.CurrentAxes; axb.FontSize=20; % axb.GridLineWidth=2;
+axb=figb.CurrentAxes; axb.FontSize=fs; % axb.GridLineWidth=2;
 
 % Set the linewidth for the plot
-pltb1.LineWidth=4; pltb2.LineWidth=4;
+pltb1.LineWidth=lw; pltb2.LineWidth=lw;
 
 % Set the position and size
-figb.Position=[900 800 560 320];
+figb.Position=[900 800 560 height];
 
 % Save the figure
 fname='rint_mwe2b.eps';

### 2024-06-04 report/experiments/maxrange/src/maxrange_rk1.m

@@ -13,6 +13,15 @@ clear;
 % Logical switch
 ppam2024=true;
 
+% Height of a single row of figures in pixels
+height=300;
+
+% Line width
+lw=2;
+
+% Fontsize
+fs=16;
+
 % Define the howitzer, atmosphere, gravity
 d20;
 
@@ -112,12 +121,9 @@ end
 % Allocate space for the data
 data=zeros(kmax,4,numshell,numtol);
 
-% Height of a single row of figures in pixels
-height=300;
-
 if (ppam2024==false)
 
-    % Process all data and generate plots
+    % Process all data and generate the plots
     for t=1:numtol
         % Get a handle to a new figure
         fig=figure();
@@ -126,7 +132,7 @@ if (ppam2024==false)
         % Position on screen
         fig.Position=[10+r*300 650-q*600 840 2*height];
         % Loop over the shell types
-        for j=1:6
+        for j=1:numshell
             % Apply Richardson extrapolation
             data(:,:,j,t)=richardson(raw_range(:,j,t),p);
             % Plot the fraction
@@ -135,13 +141,13 @@ if (ppam2024==false)
             sf.Title=title(tit{j});
             
             % Set the fontsize for the axes
-            ax=fig.CurrentAxes; ax.FontSize=10; % ax.GridLineWidth=2;
+            ax=fig.CurrentAxes; ax.FontSize=fs; % ax.GridLineWidth=2;
             
             % Set the linewidth for the plot
-            plt.LineWidth=2;
+            plt.LineWidth=lw;
         end
         % Set the title
-        sgtitle(['log2(tol) =' num2str(log2(tol(t)),'%d')]);
+        sgtitle(['log_2(tol) =' num2str(log2(tol(t)),'%d')]);
         
         % Save the figures
         fname=strcat(figpath,'maxrange_rk1_tol',...
@@ -151,7 +157,7 @@ if (ppam2024==false)
     
 else
     
-    % Process all data, but only generate a subset of the plots   
+    % Process all data, but only generate plots for G5, G6, G7
     for t=1:numtol
         % Get a handle to a new figure
         fig=figure();
@@ -160,7 +166,7 @@ else
         % Position on screen
         fig.Position=[10+r*300 650-q*600 840 height];
         % Loop over the shells
-        for j=1:6
+        for j=1:numshell
             % Apply Richardson extrapolation
             data(:,:,j,t)=richardson(raw_range(:,j,t),p);
[…]

### 2024-06-04 report/experiments/maxrange/src/maxrange_rk2.m

@@ -13,6 +13,15 @@ clear;
 % Logical switch
 ppam2024=true;
 
+% Height of a single row of figures in pixels
+height=300;
+
+% Line width
+lw=2;
+
+% Fontsize
+fs=16;
+
 % Define the howitzer, atmosphere, gravity
 d20;
 
@@ -112,9 +121,6 @@ end
 % Allocate space for the data
 data=zeros(kmax,4,numshell,numtol);
 
-% Height of a single row of figures in pixels
-height=300;
-
 if (ppam2024==false)
     
     % Process all data and generate the plots
@@ -135,13 +141,13 @@ if (ppam2024==false)
             sf.Title=title(tit{j});
             
             % Set the fontsize for the axes
-            ax=fig.CurrentAxes; ax.FontSize=10; % ax.GridLineWidth=2;
+            ax=fig.CurrentAxes; ax.FontSize=fs; % ax.GridLineWidth=2;
             
             % Set the linewidth for the plot
-            plt.LineWidth=2;
+            plt.LineWidth=lw;
         end
         % Set the title
-        sgtitle(['log2(tol) =' num2str(log2(tol(t)),'%d')]);
+        sgtitle(['log_2(tol) =' num2str(log2(tol(t)),'%d')]);
         
         % Save the figures
         fname=strcat(figpath,'maxrange_rk2_tol',...
@@ -159,10 +165,11 @@ else
         q=floor((t-1)/4); r=(t-1)-4*q;
         % Position on screen
         fig.Position=[10+r*300 650-q*600 840 height];
-        % Loop over G5, G6 and G7
-        for j=1:6
+        % Loop over the shells
+        for j=1:numshell
             % Apply Richardson extrapolation
             data(:,:,j,t)=richardson(raw_range(:,j,t),p);
+            % Only generate plots for G5, G6, G7
             if (3<=j && j<=5)
                 % Plot the fraction
                 sf=subplot(1,3,j-2); plt=plot_fraction(data(:,:,j,t),p);
@@ -170,14 +177,14 @@ else
                 sf.Title=title(tit{j});
                 
                 % Set the fontsize for the axes
-                ax=fig.CurrentAxes; ax.FontSize=10; % ax.GridLineWidth=2;
+                ax=fig.CurrentAxes; ax.FontSize=fs; % ax.GridLineWidth=2;
                 
                 % Set the linewidth for the plot
-                plt.LineWidth=2;
+                plt.LineWidth=lw;
[…]

### 2024-06-04 report/src/draft.tex
Fixed the force-field versus force field issue. Thanks Pablo Garcia-Risueno

@@
 %
-\section{Motivation} Consider the problem simulating the motion of a system of atoms moving in a force-field $\bm{f}$ subject to a set of constraints $\bm{g}$.
+\section{Motivation} Consider the problem simulating the motion of a system of atoms moving in a force field subject to a set of constraints.
 In this case, Newton's 2nd law takes the form of the following system of differential algebraic equations
@@
 The vector $\bm{q}$ represents the position of the atoms.
-The vector $\bm{v}$ represents the velocities of the atoms. The function $\bm{f}$ represents the force-field acting on the atoms.
+The vector $\bm{v}$ represents the velocities of the atoms. The function $\bm{f}$ represents the force acting on the atoms.
 The nonsingular diagonal matrix $\bm{M}$ lists the masses of the atoms.
@@
 Now let $T \in \mathbb{R}$ denote any target value that can be computed in terms of the trajectory $t \rightarrow (q(t), v(t))$ and let $A_h$ denote the corresponding value obtained from the output of the SHAKE algorithm.
-It is clear that $T$ and $A_h$ are both functions of the force-field $\bm{f}$ and the question of adjusting $\bm{f}$ to match the outcome of a physical experiment naturally suggests itself.
+It is clear that $T$ and $A_h$ are both functions of the force field $\bm{f}$ and the question of adjusting $\bm{f}$ to match the outcome of a physical experiment naturally suggests itself.
 Let therefore $T_0 \in \mathbb{R}$ be given and consider the problem of solving the equation
@@
 For each experiment, we computed the total kinetic and potential energy of the system at the end of the simulation.
-The function {\tt gromacs\_figures} will generate Figure~\ref{fig:oplsaaltol04} and Figure~\ref{fig:oplsaaltol12} for the OPLS-AA/L force-file and similar figures for CHARMM36.
+The function {\tt gromacs\_figures} will generate Figure~\ref{fig:oplsaaltol04} and Figure~\ref{fig:oplsaaltol12} for the OPLS-AA/L force field and similar figures for CHARMM36.
 
@@
 In molecular dynamics, it is common to ignore the interaction between atoms that are far away.
-This can be done by setting force-fields to zero outside of a sufficiently large ball.
-There is more than one way to achieve this and the documentation for GROMACS 2021 discusses its use of force-fields that are not of class $C^{\infty}$.
+This can be done by setting force fields to zero outside of a sufficiently large ball.
+There is more than one way to achieve this and the documentation for GROMACS 2021 discusses its use of force fields that are not of class $C^{\infty}$.
 
@@
 where $c > 0$ is a suitable constant.
-The script {\tt iontrap\_mwe1} does not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho > 0$.
+The script {\tt iontrap\_mwe1} does not modify the electrostatic force fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho > 0$.
 The scripts {\tt iontrap\_mwe2} and {\tt iontrap\_mwe4} replace $\bm{f}$ with
@@
 The function $g_4$ is of class $C^\infty$ and satisfies $g_4(r) = 1$ for $r < 0.5 \rho$ and $g_4(r) = 0$ for $r \ge 0.95 \rho$.
-It is clear that changing the force-fields impacts the motion, but can we estimate the discretization error and quantify this effect?
+It is clear that changing the force fields impacts the motion, but can we estimate the discretization error and quantify this effect?
 Figure \ref{fig:iontrap_mwe1}, \ref{fig:iontrap_mwe2}, \ref{fig:iontrap_mwe4} show the evolution of Richardson's fraction for the kinetic energy at the end of each simulation.
 
-When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, or when the perturbation is smooth, see Figure~\ref{fig:iontrap_mwe4}, then the experiments support the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.x1
+When the force fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, or when the perturbation is smooth, see Figure~\ref{fig:iontrap_mwe4}, then the experiments support the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.x1
 \begin{figure}[h!]
@@
 \end{figure}
-When the force-fields are truncated and discontinuities are introduced into the simulation, see Figure~\ref{fig:iontrap_mwe2}, there is no evidence to support the existence of an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
+When the force fields are truncated and discontinuities are introduced into the simulation, see Figure~\ref{fig:iontrap_mwe2}, there is no evidence to support the existence of an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
 \begin{figure}

### 2024-06-04 report/experiments/gromacs/src/gromacs_plot.m

@@ -33,5 +33,5 @@ plt1.LineWidth=lw;
 plt2.LineWidth=lw;
 
 % Set a common title
-tit=sgtitle(['Forcefield = ' field '     constraint tolerance = ' num2str(tol,'%.2e')]);
+tit=sgtitle(['Force field = ' field '     constraint tolerance = ' num2str(tol,'%.2e')]);
 tit.FontSize=32;

### 2024-08-29 slides/ostrava/cckm-ppam-2024.tex

\documentclass[usenames,dvipsnames]{beamer}
\usepackage{graphicx}
\usepackage{tablefootnote}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{array, booktabs, multicol,hyperref}
\usepackage{ulem}

% Algorithms package
\usepackage[vlined, ruled, lines numbered]{algorithm2e}

% Tikz
\usepackage{tikz}
\usetikzlibrary{shapes}

\usepackage{url}

% LaTeX commands added by CCKM to process tables
% Adjust separation between rows
\renewcommand{\arraystretch}{1.1}
% Commands which allow the width to be set
\newcolumntype{L}[1]{>{\raggedright \let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering   \let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft  \let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

% Augmented vector notation
\newcommand{\av}[2]{\langle #1, #2 \rangle}

\graphicspath{{figures/}{data/}}

\beamertemplatenavigationsymbolsempty

\setbeamertemplate{footline}[frame number]
\definecolor{bostonuniversityred}{rgb}{0.8, 0.0, 0.0}
\usetheme{Copenhagen}

\title[Accuracy and smoothness]{The need for accuracy and smoothness in numerical simulations}

\author[Carl Christian Kjelgaard Mikkelsen et al.]{\underline{Carl Christian Kjelgaard Mikkelsen} \and
Lori{\'e}n L{\'o}pez-Villellas}

\institute{
  Department of Computer Science, Ume{\aa} University, Sweden, \\ {\tt spock@cs.umu.se} \and Department of Computer Science, Zaragoza University, Spain, \\ {\tt lorien.lopez@unizar.es}
}

\date{PPAM 2024, September 8-11th 2024,  Ostrava, Czechia}

\titlegraphic{
  \vspace*{0mm}
  \centering
  \includegraphics[height=1.0cm]{umu-logotyp-EN.png} \;\;
  \includegraphics[height=1.5cm]{Solemne_color.png} \;\;
  \includegraphics[height=1.0cm]{87660_essence_rgb.png}
}

\begin{document}

\frame{\titlepage}

\frame{\frametitle{Introduction}

  This is not a talk about, say, 
  \vspace{0.5cm}
\begin{enumerate}\setlength\itemsep{1em}
\item computational speed
\item parallelism
\item energy efficency
\end{enumerate}
\vspace{0.5cm}
This is a talk about
\vspace{0.5cm}
\begin{center}
  \underline{the relevance of the numbers returned by our simulations}
\end{center}
}

\frame{\frametitle{External ballistics}

  \includegraphics[width=12cm]{range_rk1_mwe2.pdf}

  The total force $\bm{F}$ acting on the shell 
  \begin{equation}
    \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
  \end{equation}
  is the combination of gravity and aerodynamic drag.
}

\frame{\frametitle{External ballistics}

\begin{block}{Key questions}
  Does it matter
  \begin{enumerate}
  \item if the drag coefficient
    \begin{equation*}
      \nu \rightarrow C_D(\nu)
    \end{equation*}
    is smooth or not?
  \item if the event equation
    \begin{equation*}
      y(t) = 0
    \end{equation*}
    is solved accurately or not?
      \end{enumerate}
 \end{block}
}

\frame{\frametitle{Molecular dynamics}

  The system of differential algebraic equations
  \begin{align}
  \bm{q}'(t) &= \bm{v}(t), \\
  \bm{M}\bm{v}'(t) &= \bm{f}(\bm{q}(t)) - \bm{G}(\bm{q}(t))^T\bm{\lambda}(t), \\
  \bm{g}(\bm{q}(t)) &= \bm{0}. 
  \end{align}
  is solved using the SHAKE algorithm
  \begin{align}
  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda}_n \right), \\
  \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
  \bm{g}(\bm{q}_{n+1}) &= \bm{0}. \label{equ:constraint}
\end{align}
}

\frame{\frametitle{Molecular dynamics}

  \begin{block}{Key questions}
    Does it matter
    \begin{enumerate}
    \item if the force-field 
      \begin{equation*}
        \bm{q} \rightarrow \bm{f}(\bm{q})
      \end{equation*}
      is smooth or not?
    \item if the nonlinear constraint equation
      \begin{equation}
        \bm{g}(\bm{q}_{n+1}(\bm{\lambda})) = \bm{0}
      \end{equation}
      is solved accurately with respect to $\bm{\lambda}$ or not?
    \end{enumerate}
  \end{block}
}

\frame{\frametitle{The primary point of this talk}

  If one of the following is true:
  \vspace{0.5cm}
  \begin{enumerate}
  \item the central functions are not smooth enough
  \item the central equations are not solved accurately enough
  \end{enumerate}
  \vspace{0.5cm}
  then we cannot:
  \vspace{0.5cm}
  \begin{enumerate}
  \item estimate the modelling error
  \item validate the model against the physical reality
  \end{enumerate}
  \vspace{1cm}
}

\frame{\frametitle{The secondary point of this talk}

  It might be possible to assert experimentally if
  \vspace{0.5cm}
  \begin{enumerate} \setlength\itemsep{1em}
  \item your functions are sufficiently smooth
  \item your equations are solved with sufficient accuracy
  \end{enumerate}
}

\frame{\frametitle{The key terms of this talk}

  \begin{enumerate} \setlength\itemsep{1em}
  \item $P$: physical quantity that can be measured
  \item $T$: approximation of $P$ predicted by our model
  \item $A_h$: approximation of $T$ returned by our algorithm
  \item $\hat{A}_h$: the computed value of $A_h$.
  \end{enumerate}
}

\frame{\frametitle{The three different error terms}

  \begin{enumerate}
  \item $P - T$ is the modelling error and
    \begin{equation}
      P - T \not = 0
    \end{equation}
    because our model is simpler than the real world.
  \item $T - A_h$ is the discretization error and
    \begin{equation}
      T - A_h \not = 0
    \end{equation}
    because we cannot solve most differential equations exactly.
  \item $A_h - \hat{A}_h$ is the computational error and
    \begin{equation}
      A_h - \hat{A}_h \not = 0
    \end{equation}
    due to rounding errors and truncation errors.
  \end{enumerate}
}

\frame{\frametitle{Why should we care?}

  We validate our models by demonstrating that
  \begin{equation}
    P - T \approx 0
  \end{equation}
  is a good approximation, but
  \begin{equation}
    P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
  \end{equation}
  so we need to assert that
  \begin{align}
    |T- A_h| & \ll |P - \hat{A}_h| \\
    |A_h - \hat{A}_h| & \ll |P - \hat{A}_h|
  \end{align}
  before we are certain that
  \begin{equation}
    P - T \approx P - \hat{A}_h
  \end{equation}
  is a good approximation.
}

\frame{\frametitle{Practical error estimation}

  It is frequently possible to simultanously
  \begin{enumerate}
  \item assert that the computational error
    \begin{equation*}
      A_h - \hat{A}_h
    \end{equation*}
    is irrelevant, and
  \item estimate the discretization error
    \begin{equation*}
      T_h - A_h
    \end{equation*}
    accurately
  \end{enumerate}
  The key is the existence of an asymptotic error expansion
  \begin{equation}
    E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
  \end{equation}
  where $0 < p < q < r$.
}

\frame{\frametitle{Elementary results}

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
If 
\begin{equation}
  E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
\end{equation}
then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
Why is this significant when the computer returns $\hat{A}_h$ and $$\hat{A}_h \not = A_h$$
}

\frame{\frametitle{An example from elementary numerical analysis}

  Our target value is 
  \begin{equation}
    T = \int_0^1 f(x) dx, \quad f(x) = \exp(x)
  \end{equation}
  Our approximation is the composite trapezoidal rule
  \begin{equation}
    A_h = \frac{h}{2} \sum_{j=0}^{n-1} \left( f(x_i) + f(x_{i+1}) \right), \quad x_i = ih, \quad nh = 1
  \end{equation}
  We compute $A_h$ for
  \begin{equation}
    h = h_k = 2^{-k}
  \end{equation}
}

\frame{\frametitle{The evolution of $F_h$ and the quality of $R_h$}

  \begin{figure}[ht]
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe1a.pdf}
            \caption{The evolution of $F_h$}
            \label{fig:a}
        \end{minipage}
        \hspace{0.5cm}
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe1b.pdf}
            \caption{The accuracy of $R_h$}
            \label{fig:b}
        \end{minipage}
    \end{figure}
    Strictly speaking we are not observing $F_h$ and $R_h$
  \begin{center}
    we are observing $\hat{F}_h$ and $\hat{R}_h$
  \end{center}
  The discrepancy is controlled by the computational error
  $$ A_h - \hat{A}_h$$
}

\frame{\frametitle{A spectacular failure}

  We used GROMACS to simulate the motion of hen egg white lysozyme submerged in water.
  \begin{enumerate} \setlength\itemsep{1em}
  \item $T$: total energy of the simulation at the end
  \item $A_h$: the approximation of $T$ computed using SHAKE
  \item $\hat{A}_h$: the value of $A_h$ returned by the computer
  \end{enumerate}
\vspace{0.5cm}
  Temporal matters:
  \begin{enumerate}
  \item The length of the simulations was $10^{-12}$ s ($1$ ps)
  \item A common step size in MD is $10^{-15} s$ ($1$ fs)
  \end{enumerate}

  \begin{center}
    So, a mere 1000 steps! What could possibly go wrong?
  \end{center}
}

\frame{\frametitle{A spectacular failure}

  \begin{figure}
    \includegraphics[width=12cm]{oplsaaltol12.pdf}
  \end{figure}
   The rapid growth of the energy for large value of $n$ 
    \begin{center}
      can likely be cured using compensated summation.
    \end{center}
  }

\frame{\frametitle{A spectacular failure}

  \begin{figure}
    \includegraphics[width=11cm]{oplsaaltol12.pdf}
  \end{figure}

  \begin{itemize}
  \item The \underline{wiggles} near $n=1000$ steps are a great concern.
  \item If there is an AEX, then $1$ fs is \underline{not} inside the asymp. range.
  \item We cannot quantifty the effect of changing the force-field.
  \end{itemize}

}

\frame{\frametitle{What can kill an AEX?}

}

\frame{\frametitle{Why should this concern the PPAM community?}

  We use every trick in the book to
  \begin{enumerate}
  \item reduce time-to-solution
  \item increase the parallel efficicency
  \item reduce energy-to-solution
  \end{enumerate}
  \vspace{1cm}
  We should not forget to ask the question:
  \begin{center}
    Did we just compromise our ability to validate our models against the real world?
  \end{center}
}

\end{document}

### 2024-08-30 slides/ostrava/cckm-ppam-2024.tex
Added more pictures to the slides for PPAM 2024

@@
 
-\graphicspath{{figures/}{data/}}
+\graphicspath{{../../report/fig/}}
 
@@
   \begin{align}
-  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda}_n \right), \\
+  \bm{v}_{n+1/2} &= \bm{v}_{n-1/2} + \bm{h} \bm{M}^{-1} \left( \bm{f}(\bm{q}_n) - \bm{G}(\bm{q}_n)^T \bm{\lambda} \right), \\
   \bm{q}_{n+1} &= \bm{q}_n + h \bm{v}_{n + 1/2}, \\
@@
   \begin{enumerate} \setlength\itemsep{1em}
+  \item rounding errors are irrelevant
   \item your functions are sufficiently smooth
@@
 \end{equation}
-Why is this significant when the computer returns $\hat{A}_h$ and $$\hat{A}_h \not = A_h$$
+Why is this significant? The computer returns $$\hat{A}_h \not = A_h$$
 }
@@
 
-\frame{\frametitle{What can kill an AEX?}
+\frame{\frametitle{When do we have an AEX?}
 
+  Deriving an AEX is an exercise in Taylor expansions. If
+  \begin{center}
+    our functions are not many times differentiable
+  \end{center}
+  then the foundation crumbles.
+  \vspace{1cm}
+    \newline
+  Every AEX refers to the exact value of the $A_h$. If
+  $$\hat{A}_h \approx A_h$$
+  is not a good approximation, then
+  \begin{center}
+    $\hat{A}_h$ will not behave nicely
+  \end{center}
+}
+
+\frame{\frametitle{External ballistics: Accurate event location}
+
+  \begin{figure}
+    \includegraphics[width=10cm]{maxrange_rk1_tol53.pdf}
+    \caption{RK1: The time to impact is computed with tolerance $\tau = 2^{-53}$. We can identify the asymptotic range and we trust the error estimate.}
+  \end{figure}
+}
+
+\frame{\frametitle{External ballistics: Inaccurate event location}
+
+  \begin{figure}
+    \includegraphics[width=10cm]{maxrange_rk2_tol11.pdf}
+    \caption{RK2: The time to impact is computed with tolerance $\tau = 2^{-11}$ and we cannot locate the asymptotic range nor trust the error estimate.}
+  \end{figure}
+}
+
+\frame{\frametitle{External ballistics: Sufficiently event location}
+
+  \begin{figure}
+    \includegraphics[width=10cm]{maxrange_rk2_tol25.pdf}
+    \caption{RK2: The time to impact is computed with tolerance $\tau = 2^{-25}$ and the asymptotic range begins to emerge}
+  \end{figure}
+}
+
+\frame{\frametitle{Molecular dynamics: Truncation of force-fields}
+
+  \begin{itemize}
+  \item Simple simulation with 4 ions
+  \item Each ion is attached to the origin with a spring
+  \item The electrostatic forcefields are trunctated.
+  \end{itemize}
 }

### 2024-09-04 slides/ostrava/cckm-ppam-2024.tex
Revised slides for PPAM 2024

--- 2024-08-29 slides/ostrava/cckm-ppam-2024.tex
@@
 
-\frame{\frametitle{External ballistics}
+\frame{\frametitle{Warning: Disturbing pictures to follow}
 
-  \includegraphics[width=12cm]{range_rk1_mwe2.pdf}
+  We did two sets of simulations of a molecule in water
+  \vspace{1cm}
+  \begin{enumerate} \setlength\itemsep{1em}
+    \item fixed real time
+    \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
+    \item identical initial conditions
+   \end{enumerate}
+  \vspace{1cm}
+  and tracked the energy as a function of the number of steps.
 
-  The total force $\bm{F}$ acting on the shell 
-  \begin{equation}
-    \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
-  \end{equation}
-  is the combination of gravity and aerodynamic drag.
 }
 
-\frame{\frametitle{External ballistics}
+\frame{\frametitle{Warning: Disturbing pictures below}
 
-\begin{block}{Key questions}
-  Does it matter
-  \begin{enumerate}
-  \item if the drag coefficient
-    \begin{equation*}
-      \nu \rightarrow C_D(\nu)
-    \end{equation*}
-    is smooth or not?
-  \item if the event equation
-    \begin{equation*}
-      y(t) = 0
-    \end{equation*}
-    is solved accurately or not?
-      \end{enumerate}
- \end{block}
+ \begin{figure}
+    \includegraphics[width=8cm]{oplsaaltol12.pdf}
+  \end{figure}
+
+  \begin{figure}
+    \includegraphics[width=8cm]{charmm36tol12.pdf}
+  \end{figure}
+
 }
+
+%% \frame{\frametitle{External ballistics}
+
+%%   \includegraphics[width=12cm]{range_rk1_mwe2.pdf}
+
+%%   The total force $\bm{F}$ acting on the shell 
+%%   \begin{equation}
+%%     \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
+%%   \end{equation}
+%%   is the combination of gravity and aerodynamic drag.
+%% }
+
+%% \frame{\frametitle{External ballistics}
+
+%% \begin{block}{Key questions}
+%%   Does it matter
+%%   \begin{enumerate}
+%%   \item if the drag coefficient
+%%     \begin{equation*}
+%%       \nu \rightarrow C_D(\nu)
+%%     \end{equation*}
+%%     is smooth or not?
+%%   \item if the event equation
+%%     \begin{equation*}
+%%       y(t) = 0
+%%     \end{equation*}
+%%     is solved accurately or not?
+%%       \end{enumerate}
+%%  \end{block}
+%% }
 
@@
 
-\frame{\frametitle{The secondary point of this talk}
+%% \frame{\frametitle{The secondary point of this talk}
 
-  It might be possible to assert experimentally if
-  \vspace{0.5cm}
-  \begin{enumerate} \setlength\itemsep{1em}
-  \item your functions are sufficiently smooth
-  \item your equations are solved with sufficient accuracy
-  \end{enumerate}
-}
+%%   It might be possible to assert experimentally if
+%%   \vspace{0.5cm}
+%%   \begin{enumerate} \setlength\itemsep{1em}
+%%   \item your functions are sufficiently smooth
+%%   \item your equations are solved with sufficient accuracy
+%%   \end{enumerate}
+%% }
 
@@
   \begin{equation}
+    P - T \approx P - \hat{A}_h
+  \end{equation}
+  is not necessarily a good approximation. We have
+
+  \begin{equation}
     P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
@@
   \end{align}
-  before we are certain that
-  \begin{equation}
-    P - T \approx P - \hat{A}_h
-  \end{equation}
-  is a good approximation.
 }
@@
 \end{equation}
-Why is this significant when the computer returns $\hat{A}_h$ and $$\hat{A}_h \not = A_h$$
+which implies
+\begin{equation}
+  \log | 2^p - F_h| \approx \log|\text{constant}| + (q-p) \log h
+\end{equation}
+
 }
@@
 
-  We used GROMACS to simulate the motion of hen egg white lysozyme submerged in water.
+  GROMACS simulation of hen egg white lysozyme in water:
+  \vspace{0.25cm}
   \begin{enumerate} \setlength\itemsep{1em}
@@
 Temporal matters:
+\vspace{0.25cm}
   \begin{enumerate}
@@
   \begin{figure}
-    \includegraphics[width=12cm]{oplsaaltol12.pdf}
+    \includegraphics[width=10cm]{oplsaaltol12.pdf}
   \end{figure}
@@
   \item If there is an AEX, then $1$ fs is \underline{not} inside the asymp. range.
-  \item We cannot quantifty the effect of changing the force-field.
+  \item We cannot assert that rounding errors are irrelevant
+  \item We cannot estimate the discretication error
   \end{itemize}
@@
 
-\frame{\frametitle{What can kill an AEX?}
+\frame{\frametitle{What can go wrong?}
+
+  \begin{enumerate}\setlength\itemsep{1.5em}
+  \item Proving the existence of an AEX is an exercise in
+    \begin{center}
+      Taylor expansion
+    \end{center}
+    We cannot complete the proof unless
+    \begin{center}
+      the functions are sufficiently smooth.
+      \end{center}
+    \item We cannot expect $\hat{A}_h$ to behave as predicted by the AEX unless
+    $$ \hat{A}_h \approx A_h $$
+    is a good approximation
+    \end{enumerate}
+}
+
+\frame{\frametitle{External ballistics}
+
+  \includegraphics[width=12cm]{range_rk1_mwe2.pdf}
+
+  The total force $\bm{F}$ acting on the shell 
+  \begin{equation}
+    \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
+  \end{equation}
+  is the combination of gravity and aerodynamic drag.
+}
+
+\frame{\frametitle{Computing the optimal range of the a howitzer}
+
+ \begin{figure}
+    \includegraphics[width=10cm]{maxrange_rk1_tol18.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and sufficiently accurate event location.}
+  \end{figure}
+
+}
+
+\frame{\frametitle{Computing the optimal range of the a howitzer}
+
+ \begin{figure}
+    \includegraphics[width=10cm]{maxrange_rk1_tol11.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and inaccurate event location.}
+  \end{figure}
+
+}
+
+\frame{\frametitle{Modelling ions}
 

### 2024-09-04 report/scripts/gatherfig.sh

@@ -6,6 +6,8 @@ ps2pdf -DEPSCrop rint_mwe1a.eps rint_mwe1a.pdf
 ps2pdf -DEPSCrop rint_mwe1b.eps rint_mwe1b.pdf
 ps2pdf -DEPSCrop rint_mwe2a.eps rint_mwe2a.pdf
 ps2pdf -DEPSCrop rint_mwe2b.eps rint_mwe2b.pdf
+ps2pdf -DEPSCrop maxrange_rk1_tol11.eps maxrange_rk1_tol11.pdf
+ps2pdf -DEPSCrop maxrange_rk1_tol18.eps maxrange_rk1_tol18.pdf
 ps2pdf -DEPSCrop maxrange_rk1_tol53.eps maxrange_rk1_tol53.pdf
 ps2pdf -DEPSCrop maxrange_rk2_tol11.eps maxrange_rk2_tol11.pdf
 ps2pdf -DEPSCrop maxrange_rk2_tol25.eps maxrange_rk2_tol25.pdf

### 2024-09-04 slides/ostrava/cckm-ppam-2024.tex

@@
 %% \frame{\frametitle{The secondary point of this talk}
-
 %%   It might be possible to assert experimentally if
@@
 \end{equation}
-
 }
@@
 
-\frame{\frametitle{What can go wrong?}
+\frame{\frametitle{When do we have an AEX?}
 
-  \begin{enumerate}\setlength\itemsep{1.5em}
-  \item Proving the existence of an AEX is an exercise in
+  Deriving an AEX is an exercise in Taylor expansions. If
   \begin{center}
-      Taylor expansion
+    our functions are not many times differentiable
   \end{center}
-    We cannot complete the proof unless
+  then the foundation crumbles.
+  \vspace{1cm}
+    \newline
+  Every AEX refers to the exact value of the $A_h$. If
+  $$\hat{A}_h \approx A_h$$
+  is not a good approximation, then
   \begin{center}
-      the functions are sufficiently smooth.
+    $\hat{A}_h$ will not behave nicely
   \end{center}
-    \item We cannot expect $\hat{A}_h$ to behave as predicted by the AEX unless
-    $$ \hat{A}_h \approx A_h $$
-    is a good approximation
-    \end{enumerate}
 }
@@
 
+\frame{\frametitle{Molecular dynamics: Truncation of force-fields}
+
+  \begin{itemize}
+  \item Simple simulation with 4 ions
+  \item Each ion is attached to the origin with a spring
+  \item The electrostatic forcefields are trunctated.
+  \end{itemize}
+}
+
 \frame{\frametitle{Why should this concern the PPAM community?}

### 2024-09-05 slides/ostrava/cckm-ppam-2024.tex
Slides for PPAM-2024 approaching final shape. Restructured slide folder to accommodate figures shared between talks

@@
 
-\graphicspath{{../../report/fig/}}
+\graphicspath{{../fig/}{../../report/fig/}}
 
@@
 
-\frame{\frametitle{Warning: Disturbing pictures to follow}
+\frame{\frametitle{A simple experiment}
 
@@
     \item identical initial conditions
+    \item ran the constraint solver to stagnation
    \end{enumerate}
   \vspace{1cm}
-  and tracked the energy as a function of the number of steps.
+  and tracked the final energy as a function of the number of steps.
 
@@
 
-\frame{\frametitle{Warning: Disturbing pictures below}
+\frame{% \frametitle{Warning: Disturbing pictures below}
 
  \begin{figure}
-    \includegraphics[width=8cm]{oplsaaltol12.pdf}
+    \includegraphics[width=10.5cm]{oplsaaltol12.pdf}
   \end{figure}
@@
   \begin{figure}
-    \includegraphics[width=8cm]{charmm36tol12.pdf}
+    \includegraphics[width=10.5cm]{charmm36tol12.pdf}
   \end{figure}
@@
 
-\frame{\frametitle{Molecular dynamics}
+\frame{\frametitle{Molecular dynamics with constraints}
 
@@
   \begin{block}{Key questions}
-    Does it matter
-    \begin{enumerate}
+    Does it matter\vspace{0.5cm}
+    \begin{enumerate}\setlength\itemsep{1.5em}
     \item if the force-field 
@@
     \item if the nonlinear constraint equation
-      \begin{equation}
+      \begin{equation*}
         \bm{g}(\bm{q}_{n+1}(\bm{\lambda})) = \bm{0}
-      \end{equation}
+      \end{equation*}
       is solved accurately with respect to $\bm{\lambda}$ or not?
@@
   \vspace{0.5cm}
-  then we cannot:
+  then we may lose the ability to
   \vspace{0.5cm}
   \begin{enumerate}
-  \item estimate the modelling error
-  \item validate the model against the physical reality
+  \item assert that rounding errors are irrelevant
+  \item estimate the discretization error
+  \item compare the model against the physical reality
   \end{enumerate}
@@
   is not necessarily a good approximation. We have
-
   \begin{equation}
@@
   It is frequently possible to simultanously
-  \begin{enumerate}
+   \vspace{0.5cm}
+  \begin{enumerate} \setlength\itemsep{1.5em}
   \item assert that the computational error
@@
   \end{enumerate}
-  The key is the existence of an asymptotic error expansion
+}
+
+\frame{\frametitle{Practical error estimation}
+
+  The key is to have an asymptotic error expansion (AEX)
   \begin{equation}
@@
   \end{equation}
-  where $0 < p < q < r$.
+  where
+  \begin{equation}
+    0 < p < q < r
+  \end{equation}
+  are not necessarily integers.
 }
@@
 \end{equation}
-If 
-\begin{equation}
-  E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
-\end{equation}
-then
+Then
 \begin{equation}
@@
   \end{center}
-  The discrepancy is controlled by the computational error
-  $$ A_h - \hat{A}_h$$
+  The difference is controlled by the computational error
+  \begin{equation}
+    A_h - \hat{A}_h
+  \end{equation}
 }
@@
   \item The length of the simulations was $10^{-12}$ s ($1$ ps)
-  \item A common step size in MD is $10^{-15} s$ ($1$ fs)
+  \item A common step size in MD is $10^{-15} s$ ($1$ fs) or $n=1000$ steps
+  \item $n \in \{250,500,1000:100:2000,3000:1000:16000\}$
   \end{enumerate}
@@
   \begin{center}
-    So, a mere 1000 steps! What could possibly go wrong?
+   What could possibly go wrong?
   \end{center}
@@
 
-  Deriving an AEX is an exercise in Taylor expansions. If
+  \begin{enumerate} \setlength\itemsep{1.5em}
+  \item Every AEX refers to the exact value of the $A_h$. If
+    $$\hat{A}_h \approx A_h$$
+    is not a good approximation, then
+    \vspace{0.25cm}
+    \begin{center}
+      $\hat{A}_h$ will not behave nicely
+    \end{center}
+    \vspace{0.25cm}
+    and we cannot identify the asymptotic range.
+  \item Deriving an AEX is an exercise in Taylor expansions. If
+    \vspace{0.25cm}
     \begin{center}
@@
     \end{center}
+    \vspace{0.25cm}
     then the foundation crumbles.
-  \vspace{1cm}
-    \newline
-  Every AEX refers to the exact value of the $A_h$. If
-  $$\hat{A}_h \approx A_h$$
-  is not a good approximation, then
-  \begin{center}
-    $\hat{A}_h$ will not behave nicely
-  \end{center}
+  \end{enumerate}
 }
@@
 
-\frame{\frametitle{Computing the optimal range of the a howitzer}
+\frame{\frametitle{Computing the optimal range of the a howitzer: Success}
 
  \begin{figure}
-    \includegraphics[width=10cm]{maxrange_rk1_tol18.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and sufficiently accurate event location.}
+    \includegraphics[width=10cm]{maxrange_rk1_tol18.pdf} \caption{The evolution of $F_h$ for 3 different drag coeffients, 1st order Runge-Kutta and sufficiently accurate event location.}
   \end{figure}
@@
 
-\frame{\frametitle{Computing the optimal range of the a howitzer}
+\frame{\frametitle{Computing the optimal range of the a howitzer: Failure}
 
  \begin{figure}
-    \includegraphics[width=10cm]{maxrange_rk1_tol11.pdf} \caption{The evolution of Richardson's fraction for 3 different drag coeffients, Euler's explicit method and inaccurate event location.}
+    \includegraphics[width=10cm]{maxrange_rk2_tol18.pdf} \caption{The evolution of $F_h$ for 3 different drag coeffients, 2nd order Runge-Kutta and inaccurate event location.}
   \end{figure}
@@
 
-\frame{\frametitle{Modelling ions}
+\frame{\frametitle{Modelling ions: Setup}
 
+  The total force on an ion is
+  \begin{equation}
+    \bm{F}(\bm{r_i}) =  - \alpha \sum_{j \not = i} \frac{1}{\|\bm{r}_i - \bm{r}_j\|^3}(\bm{r}_i - \bm{r}_j) - \beta \bm{r_i} - \gamma \bm{v}_i
+  \end{equation}
+  We measure the total kinetic energy $A_h$ at a fixed time as function of the stepsize $h_k = 2^{-k} h_0$. 
 }
 
-\frame{\frametitle{Molecular dynamics: Truncation of force-fields}
+\frame{\frametitle{Modelling ions: Success}
 
-  \begin{itemize}
-  \item Simple simulation with 4 ions
-  \item Each ion is attached to the origin with a spring
-  \item The electrostatic forcefields are trunctated.
-  \end{itemize}
+   \begin{figure}
+     \includegraphics[width=10cm]{iontrap_mwe1.pdf}
+     \caption{The evolution of $F_h$ for Runge-Kutta methods of order $p \in \{1,2,3,4\}$ and smooth force-fields with infinite range.}
+  \end{figure}
+}
+
+\frame{\frametitle{Modelling ions: Failure}
+ \begin{figure}
+    \includegraphics[width=10cm]{iontrap_mwe2.pdf} \caption{The evolution of $F_h$ for Runge-Kutta methods of order $p \in \{1,2,3,4\}$ and truncated force-fields with jump discontinuities.}
+  \end{figure}
+}
+
+\frame{\frametitle{Modelling ions: Success}
+ \begin{figure}
+    \includegraphics[width=10cm]{iontrap_mwe4.pdf} \caption{The evolution of $F_h$ for Runge-Kutta methods of order $p \in \{1,2,3,4\}$ and smoothly truncated force-fields.}
+  \end{figure}
 }
@@
   We use every trick in the book to
-  \begin{enumerate}
+  \vspace{0.5cm}
+  \begin{enumerate} \setlength\itemsep{1.5em}
   \item reduce time-to-solution
@@
   \begin{center}
-    Did we just compromise our ability to validate our models against the real world?
+    \begin{large}
+      \textbf{Can we still validate our models against the real world?}
+    \end{large}
   \end{center}

### 2024-09-05 report/scripts/gatherfig.sh

@@ -10,6 +10,7 @@ ps2pdf -DEPSCrop maxrange_rk1_tol11.eps maxrange_rk1_tol11.pdf
 ps2pdf -DEPSCrop maxrange_rk1_tol18.eps maxrange_rk1_tol18.pdf
 ps2pdf -DEPSCrop maxrange_rk1_tol53.eps maxrange_rk1_tol53.pdf
 ps2pdf -DEPSCrop maxrange_rk2_tol11.eps maxrange_rk2_tol11.pdf
+ps2pdf -DEPSCrop maxrange_rk2_tol18.eps maxrange_rk2_tol18.pdf
 ps2pdf -DEPSCrop maxrange_rk2_tol25.eps maxrange_rk2_tol25.pdf
 ps2pdf -DEPSCrop charmm36tol04.eps charmm36tol04.pdf
 ps2pdf -DEPSCrop charmm36tol12.eps charmm36tol12.pdf

### 2024-09-08 slides/pristina/cckm-fshmn-2024.tex
Adding the slides for Pristina presentation

--- 2024-09-05 slides/ostrava/cckm-ppam-2024.tex
@@
 
-\date{PPAM 2024, September 8-11th 2024,  Ostrava, Czechia}
+\date{Visit to FSHMN \\ Pristina, Kosovo, September 12th 2024}
 
@@
 
-  This is not a talk about, say, 
-  \vspace{0.5cm}
-\begin{enumerate}\setlength\itemsep{1em}
-\item computational speed
-\item parallelism
-\item energy efficency
-\end{enumerate}
-\vspace{0.5cm}
-This is a talk about
-\vspace{0.5cm}
-\begin{center}
-  \underline{the relevance of the numbers returned by our simulations}
-\end{center}
+  \begin{itemize}
+  \item Danish nationality
+  \item Ass. Prof. of Comp. Sci. at Ume{\aa} University, Sweden
+  \item Ph.D in mathematics from Purdue University, Indiana, USA
+  \item MSc. in mathematics from Aarhus University, Denmark
+  \item Research interests
+    \begin{itemize}
+    \item Structured linear system
+    \item Nonsymmetric eigenvalue problems
+    \item Solution of nonlinear constraint equations
+    \item High performance scientific computing
+    \end{itemize}
+  \item Software
+    \begin{itemize}
+    \item Coauthor of StarNEIG
+    \item Coauthor of ILVES
+    \end{itemize}
+  \end{itemize}
 }
 
-\frame{\frametitle{A simple experiment}
+%% \frame{\frametitle{A simple experiment}
 
-  We did two sets of simulations of a molecule in water
-  \vspace{1cm}
-  \begin{enumerate} \setlength\itemsep{1em}
-    \item fixed real time
-    \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
-    \item identical initial conditions
-    \item ran the constraint solver to stagnation
-   \end{enumerate}
-  \vspace{1cm}
-  and tracked the final energy as a function of the number of steps.
+%%   We did two sets of simulations of a molecule in water
+%%   \vspace{1cm}
+%%   \begin{enumerate} \setlength\itemsep{1em}
+%%     \item fixed real time
+%%     \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
+%%     \item identical initial conditions
+%%     \item ran the constraint solver to stagnation
+%%    \end{enumerate}
+%%   \vspace{1cm}
+%%   and tracked the final energy as a function of the number of steps.
 
+%% }
+
+%% \frame{% \frametitle{Warning: Disturbing pictures below}
+
+%%  \begin{figure}
+%%     \includegraphics[width=10.5cm]{oplsaaltol12.pdf}
+%%   \end{figure}
+
+%%   \begin{figure}
+%%     \includegraphics[width=10.5cm]{charmm36tol12.pdf}
+%%   \end{figure}
+
+%% }
+
+\frame{\frametitle{External ballistics}
+
+  \includegraphics[width=12cm]{range_rk1_mwe2.pdf}
+
+  The total force $\bm{F}$ acting on the shell 
+  \begin{equation}
+    \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
+  \end{equation}
+  is the combination of gravity and aerodynamic drag.
 }
 
-\frame{% \frametitle{Warning: Disturbing pictures below}
+\frame{\frametitle{External ballistics}
 
- \begin{figure}
-    \includegraphics[width=10.5cm]{oplsaaltol12.pdf}
-  \end{figure}
-
-  \begin{figure}
-    \includegraphics[width=10.5cm]{charmm36tol12.pdf}
-  \end{figure}
-
+\begin{block}{Key questions}
+  Does it matter
+  \begin{enumerate}
+  \item if the drag coefficient
+    \begin{equation*}
+      \nu \rightarrow C_D(\nu)
+    \end{equation*}
+    is smooth or not?
+  \item if the event equation
+    \begin{equation*}
+      y(t) = 0
+    \end{equation*}
+    is solved accurately or not?
+      \end{enumerate}
+ \end{block}
 }
-
-%% \frame{\frametitle{External ballistics}
-
-%%   \includegraphics[width=12cm]{range_rk1_mwe2.pdf}
-
-%%   The total force $\bm{F}$ acting on the shell 
-%%   \begin{equation}
-%%     \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
-%%   \end{equation}
-%%   is the combination of gravity and aerodynamic drag.
-%% }
-
-%% \frame{\frametitle{External ballistics}
-
-%% \begin{block}{Key questions}
-%%   Does it matter
-%%   \begin{enumerate}
-%%   \item if the drag coefficient
-%%     \begin{equation*}
-%%       \nu \rightarrow C_D(\nu)
-%%     \end{equation*}
-%%     is smooth or not?
-%%   \item if the event equation
-%%     \begin{equation*}
-%%       y(t) = 0
-%%     \end{equation*}
-%%     is solved accurately or not?
-%%       \end{enumerate}
-%%  \end{block}
-%% }
 
@@
   \vspace{0.5cm}
-  then we may lose the ability to
+  then we will almost certainly lose the ability to
   \vspace{0.5cm}
   \begin{enumerate}
-  \item assert that rounding errors are irrelevant
-  \item estimate the discretization error
-  \item compare the model against the physical reality
+  \item assert that \underline{rounding} errors are irrelevant
+  \item estimate the \underline{discretization} error
+  \item estimate the \underline{modelling} error
   \end{enumerate}
@@
 }
-
-%% \frame{\frametitle{The secondary point of this talk}
-%%   It might be possible to assert experimentally if
-%%   \vspace{0.5cm}
-%%   \begin{enumerate} \setlength\itemsep{1em}
-%%   \item your functions are sufficiently smooth
-%%   \item your equations are solved with sufficient accuracy
-%%   \end{enumerate}
-%% }
 
@@
   \begin{equation}
-    P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
+    \underset{\text{modelling error}}{\underbrace{P - T}} = (P - \hat{A}_h) - \underset{\text{discretization error}}{\underbrace{(T - A_h)}} - \underset{\text{computational error}}{\underbrace{(A_h - \hat{A}_h)}}
   \end{equation}
@@
   \end{equation}
-  are not necessarily integers.
+  are not necessarily integers and $\alpha$, $\beta$ are independent of $h$.
+}
+
+\frame{\frametitle{Basic definitions}
+
+We define Richardson's error estimate by
+\begin{equation}
+  R_h := \frac{A_h - A_{2h}}{2^p - 1}
+\end{equation}
+and Richardson's fraction by
+\begin{equation}
+\quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
+\end{equation}
 }
@@
 \frame{\frametitle{Elementary results}
-
-Define 
+  If
  \begin{equation}
-  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
+    E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
   \end{equation}
-Then
+ then
  \begin{equation}
@@
 \end{equation}
-which implies
+}
+
+\frame{\frametitle{Implications}
+
+  \begin{enumerate}
+  \item We can determine $p$ from
+    \begin{equation}
+      F_h \rightarrow 2^p
+    \end{equation}
+  \item We can determine $q$ from
     \begin{equation}
@@
     \end{equation}
+  \item We can then compute $R_h$ and estimate
+    \begin{equation}
+      E_h \approx R_h
+    \end{equation}
+    when $h$ is \textit{sufficiently} small.
+  \end{enumerate}
 }
 
-\frame{\frametitle{An example from elementary numerical analysis}
+\frame{\frametitle{Numerical integration \texttt{rint\_mwe1}}
 
@@
   \end{equation}
+}
+
+\frame{\frametitle{Numerical integration \texttt{rint\_mwe2}}
+
+  Our target value is 
+  \begin{equation}
+    T = \int_0^1 f(x) dx, \quad f(x) = \sqrt{x}.
+  \end{equation}
+  Our approximation is the composite trapezoidal rule
+  \begin{equation}
+    A_h = \frac{h}{2} \sum_{j=0}^{n-1} \left( f(x_i) + f(x_{i+1}) \right), \quad x_i = ih, \quad nh = 1
+  \end{equation}
+  We compute $A_h$ for
+  \begin{equation}
+    h = h_k = 2^{-k}
+  \end{equation}
+}
+
+\frame{\frametitle{The evolution of $F_h$ and the quality of $R_h$}
+
+  \begin{figure}[ht]
+        \begin{minipage}[b]{0.45\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{rint_mwe2a.pdf}
+            \caption{The evolution of $F_h$}
+            \label{fig:a}
+        \end{minipage}
+        \hspace{0.5cm}
+        \begin{minipage}[b]{0.45\linewidth}
+            \centering
+            \includegraphics[width=\textwidth]{rint_mwe2b.pdf}
+            \caption{The accuracy of $R_h$}
+            \label{fig:b}
+        \end{minipage}
+    \end{figure}
+  We observe that
+  \begin{enumerate}
+  \item the asymptotic range is much wider (good!)
+  \item the error estimate is less accurate (mostly harmless)
+  \end{enumerate}
+
 }
@@
 
-  The total force on an ion is
+  The total force on an ion is given
   \begin{equation}
@@
   \end{equation}
-  We measure the total kinetic energy $A_h$ at a fixed time as function of the stepsize $h_k = 2^{-k} h_0$. 
+  where $r_k$ is the position of the $k$th ion and $v_k$ is it velocity.
+  \vspace{1cm}
+  \begin{itemize}
+  \item We wish to know the total kinetic energy $T$ at a fixed time.
+  \item We compute approximation $A_{h_k}$ for $h_k = 2^{-k} h_0$.
+  \end{itemize}
 }
@@
 
-\frame{\frametitle{Why should this concern the PPAM community?}
+\frame{\frametitle{Why should this concern the computational scientist?}
 
@@
 
+\frame{\frametitle{Links}
+
+  \begin{figure}
+    \includegraphics[width=6cm]{homepage.png}
+  \end{figure}
+  \begin{center}
+    Thank your for your attention
+  \end{center}
+}
+
 \end{document}

### 2024-09-12 report/matlab/integration/mwe/rint_mwe3.m
Fixed typo

@@ -22,7 +22,7 @@ rule=@(y,a,b,N)trapezoid(y,a,b,N);
 % The value below is wrong!!! The function is not smooth enough!
 % p=2;
 % The value below is correct!
-p = 1.5 
+p=1.5; 
 
 % Number of refinements
 kmax=26; 

### 2024-11-03 report/src/revision.tex
Added the revised manuscript to the repository. We thank the anonymous reviewers from their time, effort and insights.

--- 2024-06-04 report/src/draft.tex
@@
 %
-\section{Motivation} Consider the problem simulating the motion of a system of atoms moving in a force field subject to a set of constraints.
+\section{Motivation} Consider the problem of simulating the motion of a system of atoms moving in a force field subject to a set of constraints.
 In this case, Newton's 2nd law takes the form of the following system of differential algebraic equations
@@
 \end{proof}
-We shall now show how to demonstrate the existence of an asymptotic error expansion experimentally.
+We shall now demonstrate how the values of $p$ and $q$ can be determined by observing the exact values of $A_h$ for different values of $h$.
 We define Richardson's fraction $F_h$ using the expression
@@
   \begin{equation}
-    F_h = \frac{ A_{2h} - A_{4h}}{ A_h - A_{2h}} = 2^p \left[ \frac{ 1 + 2^m \nu h^m + \phi(h)}{ 1  + \nu h^m + \phi(2h)} \right].
+    F_h = \frac{ A_{2h} - A_{4h}}{ A_h - A_{2h}} = 2^p \left[ \frac{ 1 + 2^m \nu h^m + \phi(2h)}{ 1  + \nu h^m + \phi(h)} \right].
   \end{equation}
@@
   \begin{equation}
-    f(h) = 2^m \nu h^m + \phi(h), \quad g(h) = \nu h^m + \phi(2h).
+    f(h) = 2^m \nu h^m + \phi(2h), \quad g(h) = \nu h^m + \phi(h).
   \end{equation}
@@
   \begin{equation}
-    F_h = 2^p \left( 1 + \frac{(2^m-1)\nu h^m}{1+g(h)} + \frac{\phi(h) - \phi(2h)}{1+g(h)}\right) \rightarrow 2^p, \quad h \rightarrow 0_+
+    F_h = 2^p \left( 1 + \frac{(2^m-1)\nu h^m}{1+g(h)} + \frac{\phi(2h) - \phi(h)}{1+g(h)}\right) \rightarrow 2^p, \quad h \rightarrow 0_+
   \end{equation}
@@
 \begin{equation}
-p = \underset{h \rightarrow 0_+}{\lim} F_h
+2^p = \underset{h \rightarrow 0_+}{\lim} F_h
 \end{equation}
@@
 \end{equation}
-is a good approximation for $h$ sufficiently small. In particular, we note that the high-hand side of equation \ref{equ:asymtotic-behavior-of-Fh} is a linear function of $\log(h)$ with slope $m$.
+is a good approximation for $h$ sufficiently small. In particular, we note that the right-hand side of equation \ref{equ:asymtotic-behavior-of-Fh} is a linear function of $\log(h)$ with slope $m$.
 \section{Elementary examples}
@@
 In practice, the computed value $\hat{A}_h$ is different from the exact value $A_h$.
-However, it is often possible to assert that the computational error $A_h - \hat{A}_h$ is irrelevant and issue precise estimates of the error $T - \hat{A}_h$.
-We shall now demonstrate the procedure.
-To this end, we consider the familiar problem of computing definite integrals
+However, as we shall demonstrate shortly, is often possible to assert that the computational error is irrelevant and estimate the error $T - \hat{A}_h$ accurately.
+We begin by considering the familiar problem of computing definite integrals
 \begin{equation} \label{equ:integral}
@@
 The raw data shows that $p = 2$ cannot be true, but it is plausible that $p \approx \frac{3}{2}$.
-Figure \ref{fig:rint_mwe1a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
+Figure \ref{fig:rint_mwe2a} illustrates the evolution of the \emph{computed} values of Richardson's fraction.
 We observe that $k \rightarrow \log_2|\hat{F}_{h_k} - 2^{3/2}|$ is essentially a linear function of $k$ with slope $-\frac{1}{2}$ for $k \in \{2,3,\dots,18\}$.
@@
 The script will either read the raw data from a file or generate it from scratch.
-In any case, the script produces several figures and tables including Figure \ref{fig:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}.
+In any case, the script produces several figures and tables including Table \ref{tab:maxrange_rk1_table_tol53} and Figure \ref{fig:maxrange_rk1_fraction_tol53}.
 \begin{table}
  \caption{The computed maximum range for 6 shells fired from a D-20 howitzer.}
- \label{fig:maxrange_rk1_table_tol53} 
+ \label{tab:maxrange_rk1_table_tol53} 
  \centering
@@
 These two figures represent calculations where the final time step is computed with an error that is bounded by $u h_{k}$ where $u = 2^{-53}$ is the double precision unit roundoff.
-Figure \ref{fig:maxrange_rk1_table_tol53} lists the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
+Table \ref{tab:maxrange_rk1_table_tol53} lists the maximum range and the corresponding error estimate for each of the 6 shell types in our library using a time step of $h = 2^{-9} s$.
 In each case the error estimate suggests that the computed range is exact to the number of figures shown.
@@
 \end{figure}
-When the tolerance is large, see Figure~\ref{fig:maxrange_rk2_tol11}, there is no evidence that an asymptotic error expansion exists and there is no reason to trust Richardson's error estimate.
-When the tolerance is small, see Figure~\ref{fig:maxrange_rk2_tol25}, the fact that $\hat{F}_h$ approaches $2^p$ suggests that an asymptotic error expansion exists, but it is not trivial to determine an asymptotic range.
+When the tolerance is $\tau = 2^{-11}$, see \ref{fig:maxrange_rk2_tol11}, there is no evidence that an asymptotic error expansion exists and there is no reason to trust Richardson’s error estimate.
+When the tolerance is $\tau = 2^{-25}$, see Figure \ref{fig:maxrange_rk2_tol25}, the fact that $\hat{F}_h$ approaches $2^p$ suggests that an asymptotic error expansion exists, but it is not trivial to determine an asymptotic range.
 Regardless, it is clear that if we do not know which tolerance is sufficient and if an error estimate is required, then our safest course of action is to solve all equations as accurately as the hardware will allow.
@@
 A central task in the field of computational science is to fit a system of DAEs to the results of a physical experiment.
-It is often possible to assert when the computational errors are irrelevant and estimate the discretization error using Richardson extrapolation.
+It is often possible to assert when the computational errors are irrelevant, estimate the discretization error using Richardson extrapolation.
 This classical technique hinges on the existence of an asymptotic error expansion.
@@
 
-The authors would like to thank Jesús Alastruey-Benedé, Pablo Ibáñez and Pablo García-Risueño for stimulating discussions on the subject matter.
+We would like to thank the reviewers for their work which allowed us to improve the manuscript.
+We would like to thank Jesús Alastruey-Benedé, Pablo Ibáñez and Pablo García-Risueño for stimulating discussions on the subject matter.
 The first author is supported by eSSENCE, a collaborative e-Science programme funded by the Swedish Research Council within the framework of the strategic research areas designated by the Swedish Government.

### 2024-11-03 report/experiments/integration/src/rint_mwe1.m

@@ -32,7 +32,7 @@ rule=@(y,a,b,N)trapezoid(y,a,b,N);
 p=2; 
 
 % Number of approximations
-kmax=20; 
+kmax=26; 
 
 % True value of the integral 
 val=F(b)-F(a);

### 2024-11-03 report/experiments/maxrange/src/maxrange_rk1.m

@@ -10,8 +10,10 @@
 % Clear the work space
 clear;
 
-% Logical switch
-ppam2024=true;
+% Logical switch:
+%  ppam2024=true   generate the figures that fit in the manuscript
+%  ppam2024=false  generate all figures 
+ppam2024=false;
 
 % Height of a single row of figures in pixels
 height=300;
